\documentclass[11pt,letterpaper,twoside]{book} 
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{color}
\usepackage[T1]{fontenc}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{makeidx}
\usepackage{paralist}
\usepackage{sectsty}
\usepackage{textcomp}
\geometry{left=2cm,right=2cm,top=2.5cm,bottom=2.5cm}

% Lowercase section headers
\let\MakeUppercase\relax

% No colored border on hyperlinks
\hypersetup{colorlinks=false,pdfborder={0 0 0}}

% For the horizontal rule under sections
\sectionfont{\sectionrule{0pt}{0pt}{-10pt}{0.8pt}}

% For displaying code
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\lstset{
  aboveskip=5mm,
  basicstyle={\footnotesize\ttfamily},
  belowskip=5mm,
  breaklines=true,
  breakatwhitespace=true,
  frame=none,
  columns=flexible,
  commentstyle=\color{dkgreen},
  keepspaces=true,
  keywordstyle=\color{blue},
  language=C++,
  % literate={\ \ }{{\ \ \ \ }}1,  % Convert 2 spaces to 4 
  numbers=left,
  numbersep=10pt,
  numberstyle=\scriptsize\ttfamily\color{gray},
  rulecolor=\color{black},
  showstringspaces=false,
  stringstyle=\color{mauve},
  tabsize=2
}

\begin{document} 
 
\frontmatter
\pagenumbering{gobble}

\begin{titlepage}
	\centering
	\newgeometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}

	\vspace*{6 cm}

	\textbf{\LARGE Alex's Anthology of Algorithms}
	\\[1.0\baselineskip]
	\textbf{\Large Common Code for Contests in Concise C++}
	\\[1.0\baselineskip]
	{\large\textit{Draft v0.9, Feb 16, 2020}}

	\vspace*{11 cm}

	{\Large Alex Li}
	\\[0.5\baselineskip]
	{\small \textcircled{c} 2015---. All rights reserved.}
\end{titlepage}

\pagebreak
\vspace*{8 cm}
\centering

{\large This page is intentionally left blank.}

\vspace{\fill}
\makeatletter

\let\cleardoublepage\clearpage
\pagenumbering{roman}
\chapter{Preface}

\raggedright
\setlength{\parskip}{0.5em}

Visit \href{https://github.com/alxli/algorithm-anthology}{\texttt{github.com/alxli/algorithm-anthology}} for the most up-to-date digital version of this book.

\section*{Introduction}

Welcome to a comprehensive collection of common algorithms and data structures. The ultimate goal of this book is not to explain concepts from the ground up, but instead to explore the finer details behind their \textit{implementations}. There are many potential ways you can use this, for instance:
\begin{itemize}
	\item as a reference to help you better understand topics that you have only studied on a high level,
	\item as a printed codebook, which is a permitted resource for contests such as the ACM ICPC, or
	\item to cross-check existing code you have written for contest or coding interview questions.
\end{itemize}

Before diving into any section, it is strongly recommended that you have already studied the algorithms involved. Reading the code first is never an ideal approach to properly understand an algorithm. You should instead try proving (its correctness and time/space complexity) and implementing it from scratch.

Every topic to be explored is easily researchable online. Thus instead of including theoretical discussions, I document just enough to establish the problem being solved, notation being used, and any special trickery involved. I have also included small, non-rigorous examples to demonstrate usage of the code.

We mentioned that the implementation itself is the focus, but what makes an implementation good? The code is written with the following principles in mind:
\begin{itemize}
	\item \textit{Clarity:} A reader already familiar with an algorithm should have no problem understanding how its implementation works. Consistency in naming conventions should be emphasized, and any tricks or language-specific hacks should be documented.
	\item \textit{Concision:} To minimize the amount of scrolling and searching during the frenzy of time contests, it is helpful for code to be compact. Shorter code is also generally easier to understand, as long as it is not overly cryptic. Finally, each implementation should fit in a single source file as required by nearly all online judging systems.
	\item \textit{Efficiency:} The code here is designed to be performant on real contests, and should maintain a low constant overhead. This is often challenging in the face of clarity and tweakability, but we can hope for contest setters to be liberal with time and memory limits. If the code here times out, you can reasonably rule out insufficient constant optimization and assume that you are choosing an algorithm from a suboptimal complexity class.
	\item \textit{Genericness:} It should be easy to adapt the implementation to achieve slightly different goals. This could be through tweaking some core logic, parameters, data types, etc. If we were tweaking the code to achieve a different purpose during a live contest, we would certainly want the process to be as painless as possible. C++ templates are often used to increase tweakability at the slight cost of simplicity.
	\item \textit{Portability:} Different contest environments use different versions of C++ (though almost all of them use GCC). In order to maximize compatibility, non-standard features are avoided. The decision to adhere to C++98 standards is due to many contest systems being stuck on an older version of the language. Moreover, minimizing newer C++ specific features will make the implementations more language-agnostic. Refer to the "Portability" section below.
\end{itemize}

As these points and the title both suggest, there is a slight bias towards contests. Compiling a codebook for my personal reference during contests was indeed how this project got started. This work has become much more multipurpose now; whatever your use case is, I hope you discover something enlightening.

\begin{flushright}
Cheers.\\
--- Alex\\[0.3\baselineskip]
\end{flushright}

\section*{Portability Note}

All programs were tested with GCC and compiled for a 32-bit target using the switches below:

\begin{verbatim}
   g++ -std=gnu++98 -pedantic -Wall -Wno-long-long -O2
\end{verbatim}

This means the following are assumed about data types:

\begin{compactitem}
	\item \texttt{bool} and \texttt{char} are 8-bit.
	\item \texttt{int} and \texttt{float} are 32-bit.
	\item \texttt{double} and \texttt{long long} are 64-bit.
	\item \texttt{long double} is 96-bit.
\end{compactitem}
\vspace{5pt}

Programs are highly portable (ISO C++ 1998 compliant), \textbf{except} in the following regards:

\begin{itemize}
	\item Usage of \texttt{long long} and dependent features, which are compliant in C99/C++0x or later. 64-bit integers are a must for many contest problems.
	\item Usage of variable sized arrays. While easily replaced by vectors, they are generally simpler and avoid dynamic memory (which some argue is a bad idea for contests).
	\item Usage of GCC's built-in functions like \texttt{\_\_builtin\_popcount()} and \texttt{\_\_builtin\_clz()}. These can be extremely convenient, but are straightforward to implement if unavailable. See here for a reference: \url{https://gcc.gnu.org/onlinedocs/gcc/Other-Builtins.html}
	\item Usage of compound-literals, e.g. \texttt{vec.push\_back((mystruct)\{a, b, c\})}. This adds a little more concision by not requiring a constructor definition.
	\item Hacks that may depend on the platform (e.g. endianness), such as getting the signbit with type-punned pointers. Be weary of portability for all bitwise/lower level code.
\end{itemize}

\tableofcontents

\mainmatter
\input{chapter1}
\input{chapter2}
\input{chapter3}
\input{chapter4}
\input{chapter5}
\input{chapter6}

\backmatter

\printindex

\end{document}
