\chapter{\texorpdfstring{Elementary Algorithms}{1. Elementary Algorithms}}

\section{\texorpdfstring{Array Transformations}{1.1 Array Transformations}}
\setcounter{section}{1}
\setcounter{subsection}{0}
\subsection{Sorting Algorithms}
The following functions are equivalent to \inlinecode{std::sort()}, taking random-access iterators as a range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ to be sorted. Elements between \inlinecode{lo} and \inlinecode{hi} (including the element pointed to by \inlinecode{lo} but excluding the element pointed to by \inlinecode{hi}) will be sorted into ascending order after the function call. Optionally, a comparison function object specifying a strict weak ordering may be specified to replace the default \inlinecode{operator\textless{}}.

\begin{compactitem}
  \item \inlinecode{quicksort(lo, hi)} sorts the range using quicksort.
  \item \inlinecode{mergesort(lo, hi)} sorts the range using merge sort, which is stable.
  \item \inlinecode{heapsort(lo, hi)} sorts the range using heapsort.
  \item \inlinecode{insertion\_sort(lo, hi)} sorts the range using insertion sort, which is stable.
  \item \inlinecode{combsort(lo, hi)} sorts the range using comb sort.
  \item \inlinecode{radix\_sort(lo, hi)} sorts the range using least-significant-byte radix sort.
\end{compactitem}

\inlinecode{radix\_sort()} is the exception to the shared interface above: it takes no comparator and requires an integer value type. These functions are not meant to compete with standard library implementations in terms of speed. Instead, they are meant to demonstrate how common sorting algorithms can be concisely implemented in C++.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <functional>
#include <iterator>
#include <type_traits>
#include <vector>
\end{lstlisting}
Quicksort repeatedly selects a pivot and partitions the range so that elements comparing less than the pivot precede it, elements comparing equal stay in the middle, and elements comparing greater follow it. Divide and conquer is then applied to the two outer ranges until the original range is sorted. Despite having a worst case of $O(n^{2})$, quicksort is often faster in practice than merge sort and heapsort, which both have a worst case time complexity of $O(n \log n)$.

The pivot chosen in this implementation is always a middle element of the range to be sorted. To reduce the likelihood of encountering the worst case, the pivot can be chosen in better ways (e.g. randomly, or using the ``median of three'' technique).

\textbf{Time Complexity (Average)}: $O(n \log n)$.

\textbf{Time Complexity (Worst)}: $O(n^{2})$.

\textbf{Space Complexity}: $O(\log n)$ auxiliary stack space.

\textbf{Stable?}: No.

\begin{lstlisting}[language=C++]
template<typename It, typename Compare = std::less<>>
void quicksort(It lo, It hi, Compare comp = Compare()) {
  while (hi - lo >= 2) {
    auto pivot = *(lo + (hi - lo) / 2);
    It lt = lo, i = lo, gt = hi;
    while (i != gt) {
      if (comp(*i, pivot)) {
        std::iter_swap(lt++, i++);
      } else if (comp(pivot, *i)) {
        std::iter_swap(i, --gt);
      } else {
        ++i;
      }
    }
    if (lt - lo < hi - gt) {
      quicksort(lo, lt, comp);
      lo = gt;
    } else {
      quicksort(gt, hi, comp);
      hi = lt;
    }
  }
}
\end{lstlisting}
Merge sort first divides a list into n sublists of one element each, then recursively merges the sublists into sorted order until only a single sorted sublist remains. Merge sort is a stable sort, meaning that it preserves the relative order of elements which compare equal by \inlinecode{operator\textless{}} or the custom comparator given.

An analogous function in the C++ standard library is \inlinecode{std::stable\_sort()}, except that the implementation here requires sufficient memory to be available. When $O(n)$ auxiliary memory is not available, \inlinecode{std::stable\_sort()} falls back to a time complexity of $O(n \log^{2} n)$ whereas the implementation here will simply fail.

\textbf{Time Complexity (Average)}: $O(n \log n)$.

\textbf{Time Complexity (Worst)}: $O(n \log n)$.

\textbf{Space Complexity}: $O(\log n)$ auxiliary stack space and $O(n)$ auxiliary heap space.

\textbf{Stable?}: Yes.

\begin{lstlisting}[language=C++]
template<typename It, typename Compare = std::less<>>
void mergesort(It lo, It hi, Compare comp = Compare()) {
  if (hi - lo < 2) {
    return;
  }
  It mid = lo + (hi - lo - 1) / 2, a = lo, c = mid + 1;
  mergesort(lo, mid + 1, comp);
  mergesort(mid + 1, hi, comp);
  using T = typename std::iterator_traits<It>::value_type;
  std::vector<T> merged;
  merged.reserve(hi - lo);
  while (a <= mid && c < hi) {
    merged.push_back(comp(*c, *a) ? *c++ : *a++);
  }
  merged.insert(merged.end(), a, mid + 1);
  merged.insert(merged.end(), c, hi);
  std::copy(merged.begin(), merged.end(), lo);
}
\end{lstlisting}
Heapsort first rearranges an array to satisfy the max-heap property. Then, it repeatedly pops the max element of the heap (the left, unsorted subrange), moving it to the beginning of the right, sorted subrange until the entire range is sorted. Heapsort has a better worst case time complexity than quicksort and also a better space complexity than merge sort.

The C++ standard library equivalent is calling \inlinecode{std::make\_heap(lo, hi)}, followed by \inlinecode{std::sort\_heap(lo, hi)}.

The loop below has two modes controlled by whether \inlinecode{i \textgreater{} lo}. While \inlinecode{i \textgreater{} lo}, it is heapifying (Floyd's $O(n)$ bottom-up heap build: sift down each non-leaf from right to left). Once \inlinecode{i == lo}, the heap is built, and the loop switches to extraction: repeatedly swap the root to the back of the unsorted region (\inlinecode{*j}) and sift down the new root.

\textbf{Time Complexity (Average)}: $O(n \log n)$.

\textbf{Time Complexity (Worst)}: $O(n \log n)$.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\textbf{Stable?}: No.

\begin{lstlisting}[language=C++]
template<typename It, typename Compare = std::less<>>
void heapsort(It lo, It hi, Compare comp = Compare()) {
  using T = typename std::iterator_traits<It>::value_type;
  T tmp;
  It i = lo + (hi - lo) / 2, j = hi, parent, child;
  while (true) {
    if (i <= lo) {
      if (--j == lo) {
        return;
      }
      tmp = *j;
      *j = *lo;
    } else {
      tmp = *(--i);
    }
    parent = i;
    child = lo + 2 * (i - lo) + 1;
    while (child < j) {
      if (child + 1 < j && comp(*child, *(child + 1))) {
        child++;
      }
      if (!comp(tmp, *child)) {
        break;
      }
      *parent = *child;
      parent = child;
      child = lo + 2 * (parent - lo) + 1;
    }
    *parent = tmp;
  }
}
\end{lstlisting}
Insertion sort builds the sorted range one element at a time. It scans left to right, and for each element shifts the larger elements of the already-sorted prefix one position to the right to open a slot where the element belongs. Although its average and worst cases are quadratic, it is simple, stable, in-place, and online (it can sort a stream as elements arrive).

Its strength is being adaptive: on an already-sorted or nearly-sorted range, each element travels only a short distance, giving $O(n)$ best-case time and $O(n + d)$ in general for \inlinecode{d} total inversions. This is why it outperforms the asymptotically faster sorts on small or nearly-sorted ranges, and why hybrid sorts such as introsort and Timsort fall back to it once a subrange is small enough.

The comparison \inlinecode{comp(key, *(j - 1))} is strict, so an element never moves past an earlier element it compares equal to, which keeps the sort stable.

\textbf{Time Complexity (Best)}: $O(n)$ on an already-sorted range.

\textbf{Time Complexity (Average)}: $O(n^{2})$.

\textbf{Time Complexity (Worst)}: $O(n^{2})$.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\textbf{Stable?}: Yes.

\begin{lstlisting}[language=C++]
template<typename It, typename Compare = std::less<>>
void insertion_sort(It lo, It hi, Compare comp = Compare()) {
  using T = typename std::iterator_traits<It>::value_type;
  for (It i = lo; i != hi; ++i) {
    T key = *i;
    It j = i;
    while (j != lo && comp(key, *(j - 1))) {
      *j = *(j - 1);
      --j;
    }
    *j = key;
  }
}
\end{lstlisting}
Comb sort is an improved bubble sort. While bubble sort increments the gap between swapped elements for every inner loop iteration, comb sort fixes the gap size in the inner loop, decreasing it by a particular shrink factor in every iteration of the outer loop. The shrink factor of 1.3 is empirically determined to be the most effective.

Even though the worst case time complexity is $O(n^{2})$, a well chosen shrink factor ensures that the gap sizes are co-prime, in turn requiring astronomically large n to make the algorithm exceed $O(n \log n)$ steps. On random arrays, comb sort is only 2-3 times slower than merge sort. Its short code length relative to its good performance makes it a worthwhile algorithm to remember.

\textbf{Time Complexity (Worst)}: $O(n^{2})$.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\textbf{Stable?}: No.

\begin{lstlisting}[language=C++]
template<typename It, typename Compare = std::less<>>
void combsort(It lo, It hi, Compare comp = Compare()) {
  int gap = static_cast<int>(hi - lo);
  bool swapped = true;
  while (gap > 1 || swapped) {
    if (gap > 1) {
      gap = gap * 10 / 13;
    }
    swapped = false;
    for (It it = lo; it + gap < hi; ++it) {
      if (comp(*(it + gap), *it)) {
        std::iter_swap(it, it + gap);
        swapped = true;
      }
    }
  }
}
\end{lstlisting}
Radix sort is used to sort integer elements with a constant number of bits in linear time. This implementation works on ranges pointing to any signed or unsigned integer primitive. Signed values are handled by sorting on a key that flips the sign bit, which maps the two's-complement order onto the unsigned order so the most negative value sorts first.

Digits are processed from least significant to most significant, and each counting-sort pass is stable. After one pass, the values are sorted by the processed digit; stability ensures that a later pass preserves the ordering already established by all less-significant digits. Inductively, after the final pass the values are sorted by the entire key.

In this implementation, a power of two is chosen to be the base for the sort so that bitwise operations can be easily used to extract digits. This avoids the need to use modulo/exponentiation, which are much more expensive operations. In practice, it's been demonstrated that $2^8$ is the best choice for sorting 32-bit integers (approximately 5 times faster than \inlinecode{std::sort()}, and typically 2-4 times faster than radix sort using any other power of two chosen as the base).

\textbf{Time Complexity}: $O(n \cdot w)$ for $n$ integers of $w$ bits each.

\textbf{Space Complexity}: $O(n + 2^{b})$ auxiliary for a radix of $b$ bits, i.e. $O(n)$ for constant $b$.

\begin{lstlisting}[language=C++]
template<typename It>
void radix_sort(It lo, It hi) {
  if (hi - lo < 2) {
    return;
  }
  const int radix_bits = 8;
  const int radix_base = 1 << radix_bits;  // e.g. 2^8 = 256
  const int radix_mask = radix_base - 1;   // e.g. 2^8 - 1 = 0xFF
  using T = typename std::iterator_traits<It>::value_type;
  using U = typename std::make_unsigned<T>::type;
  int num_bits = 8 * sizeof(T);  // 8 bits per byte
  // Sort on an unsigned key. For signed types, flipping the sign bit sends the most negative value
  // to 0, mapping the signed order onto the unsigned order; logical shifts then extract each digit.
  auto key = [](T x) -> U {
    U u = static_cast<U>(x);
    return std::is_signed<T>::value ? (u ^ (U(1) << (8 * sizeof(T) - 1))) : u;
  };
  std::vector<T> buf(hi - lo);
  for (int pos = 0; pos < num_bits; pos += radix_bits) {
    std::vector<int> count(radix_base);
    for (It it = lo; it != hi; ++it) {
      count[(key(*it) >> pos) & radix_mask]++;
    }
    std::vector<T *> bucket(radix_base);
    T *curr = buf.data();
    for (int i = 0; i < radix_base; curr += count[i++]) {
      bucket[i] = curr;
    }
    for (It it = lo; it != hi; ++it) {
      *bucket[(key(*it) >> pos) & radix_mask]++ = *it;
    }
    std::copy(buf.begin(), buf.end(), lo);
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cstdlib>
#include <ctime>
#include <iomanip>
#include <iostream>
#include <random>
#include <vector>
using namespace std;

template<typename It>
void print_range(It lo, It hi) {
  while (lo != hi) {
    cout << *lo++ << " ";
  }
  cout << endl;
}

template<typename It>
bool sorted(It lo, It hi) {
  while (++lo != hi) {
    if (*lo < *(lo - 1)) {
      return false;
    }
  }
  return true;
}

bool compare_as_ints(double i, double j) {
  return static_cast<int>(i) < static_cast<int>(j);
}

int main() {
  {  // Can be used to sort arrays like std::sort().
    vector<int> a{32, 71, 12, 45, 26, 80, 53, 33};
    quicksort(a.begin(), a.end());
    assert(sorted(a.begin(), a.end()));
  }
  {  // STL containers work too.
    vector<int> a{32, 71, 12, 45, 26, 80, 53, 33};
    quicksort(a.begin(), a.end());
    assert(sorted(a.begin(), a.end()));
  }
  {  // Reverse iterators work as expected.
    vector<int> a{32, 71, 12, 45, 26, 80, 53, 33};
    heapsort(a.rbegin(), a.rend());
    assert(sorted(a.rbegin(), a.rend()));
  }
  {  // We can sort doubles just as well.
    vector<double> a{1.1, -5.0, 6.23, 4.123, 155.2};
    combsort(a.begin(), a.end());
    assert(sorted(a.begin(), a.end()));
  }
  {  // Insertion sort is adaptive: an already-sorted range is left untouched in O(n).
    vector<int> a{12, 26, 32, 33, 45, 53, 71, 80};
    insertion_sort(a.begin(), a.end());
    assert(sorted(a.begin(), a.end()));
  }
  {  // radix_sort() handles signed integers (including negatives), unlike a plain counting sort.
    vector<int> a{32, -71, 12, -45, 26, -80, 53, 33};
    radix_sort(a.begin(), a.end());
    assert(sorted(a.begin(), a.end()));
  }
  {  // Empty and singleton ranges are valid no-ops.
    vector<int> empty, single{42};
    quicksort(empty.begin(), empty.end());
    mergesort(single.begin(), single.end());
    assert(empty.empty() && single[0] == 42);
  }

  // Example from: http://www.cplusplus.com/reference/algorithm/stable_sort
  const vector<double> a{3.14, 1.41, 2.72, 4.67, 1.73, 1.32, 1.62, 2.58};
  {
    vector<double> v(a);
    cout << "mergesort() with default comparisons: ";
    mergesort(v.begin(), v.end());
    print_range(v.begin(), v.end());
  }
  {
    vector<double> v(a);
    cout << "mergesort() with 'compare_as_ints()': ";
    mergesort(v.begin(), v.end(), compare_as_ints);
    print_range(v.begin(), v.end());
  }
  cout << "------" << endl;

  std::mt19937 rng(std::random_device{}());
  vector<int> data;
  for (int i = 0; i < 5000000; i++) {
    data.push_back(static_cast<int>(rng()));
  }
  cout << "Sorting five million integers..." << endl;
  cout.precision(3);

  // Each sort is wrapped in a lambda so its name resolves at the call site (the bare names are
  // function templates, not values), then timed on a fresh copy of the same unsorted data.
  auto benchmark = [&](const string &name, auto sort) {
    vector<int> v = data;
    clock_t start = clock();
    sort(v.begin(), v.end());
    double t = static_cast<double>(clock() - start) / CLOCKS_PER_SEC;
    cout << setw(14) << left << name + "(): " << fixed << t << "s" << endl;
    assert(sorted(v.begin(), v.end()));
  };
  benchmark("std::sort", [](auto lo, auto hi) { std::sort(lo, hi); });
  benchmark("quicksort", [](auto lo, auto hi) { quicksort(lo, hi); });
  benchmark("mergesort", [](auto lo, auto hi) { mergesort(lo, hi); });
  benchmark("heapsort", [](auto lo, auto hi) { heapsort(lo, hi); });
  benchmark("combsort", [](auto lo, auto hi) { combsort(lo, hi); });
  benchmark("radix_sort", [](auto lo, auto hi) { radix_sort(lo, hi); });
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
mergesort() with default comparisons: 1.32 1.41 1.62 1.73 2.58 2.72 3.14 4.67
mergesort() with 'compare_as_ints()': 1.41 1.73 1.32 1.62 2.72 2.58 3.14 4.67
------
Sorting five million integers...
std::sort():  0.223s
quicksort():  0.253s
mergesort():  0.621s
heapsort():   0.439s
combsort():   0.365s
radix_sort(): 0.016s
\end{exampleoutput}
\subsection{Array Rotation}
These functions are equivalent to \inlinecode{std::rotate()}, taking three iterators \inlinecode{lo}, \inlinecode{mid}, and \inlinecode{hi} (\inlinecode{lo} $\leq$ \inlinecode{mid} $\leq$ \inlinecode{hi}) to perform a left rotation on the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$. After the function call, $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ will consist of the concatenation of elements originally in $[{\inlinecodemath{mid}}, {\inlinecodemath{hi}})$ + $[{\inlinecodemath{lo}}, {\inlinecodemath{mid}})$. That is, the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ will be rearranged in such a way that the element at \inlinecode{mid} becomes the first element of the new range, and the element before \inlinecode{mid} becomes the last element, all while preserving the relative ordering of elements within the two rotated subarrays.

All three versions below achieve the same result using in-place algorithms.

\begin{compactitem}
  \item \inlinecode{rotate1(lo, mid, hi)} uses a straightforward swapping algorithm requiring ForwardIterators.
  \item \inlinecode{rotate2(lo, mid, hi)} requires BidirectionalIterators, employing a well-known trick with three simple reversals.
  \item \inlinecode{rotate3(lo, mid, hi)} requires random-access iterators, applying a juggling algorithm which first divides the range into \inlinecode{gcd(hi - lo, mid - lo)} sets and then rotates the corresponding elements in each set.
\end{compactitem}

Writing the original range as \inlinecode{A B}, the reversal method first produces \inlinecode{reverse(A) reverse(B)}; reversing that whole range yields \inlinecode{B A} while restoring the order within both parts. The juggling method follows the permutation that sends each position to its rotated destination. Those moves form \inlinecode{gcd(n, shift)} disjoint cycles, so walking each cycle once moves every element to its final position.

\textbf{Time Complexity}: $O(n)$ per call to all versions, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.

\textbf{Space Complexity}: $O(1)$ auxiliary for all versions

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <numeric>

template<typename It>
void rotate1(It lo, It mid, It hi) {
  if (lo == mid || mid == hi) {
    return;
  }
  It next = mid;
  while (lo != next) {
    std::iter_swap(lo++, next++);
    if (next == hi) {
      next = mid;
    } else if (lo == mid) {
      mid = next;
    }
  }
}

template<typename It>
void rotate2(It lo, It mid, It hi) {
  if (lo == mid || mid == hi) {
    return;
  }
  std::reverse(lo, mid);
  std::reverse(mid, hi);
  std::reverse(lo, hi);
}

template<typename It>
void rotate3(It lo, It mid, It hi) {
  if (lo == mid || mid == hi) {
    return;
  }
  int n = hi - lo, jump = mid - lo;
  int g = std::gcd(jump, n), cycle = n / g;
  for (int i = 0; i < g; i++) {
    int curr = i, next;
    for (int j = 0; j < cycle - 1; j++) {
      next = curr + jump;
      if (next >= n) {
        next -= n;
      }
      std::iter_swap(lo + curr, lo + next);
      curr = next;
    }
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <iostream>
#include <vector>
using namespace std;

int main() {
  vector<int> a0, a1, a2, a3;
  for (int i = 0; i < 10000; i++) {
    a0.push_back(i);
  }
  a1 = a2 = a3 = a0;
  int mid = 5678;
  std::rotate(a0.begin(), a0.begin() + mid, a0.end());
  rotate1(a1.begin(), a1.begin() + mid, a1.end());
  rotate2(a2.begin(), a2.begin() + mid, a2.end());
  rotate3(a3.begin(), a3.begin() + mid, a3.end());
  assert(a0 == a1 && a0 == a2 && a0 == a3);

  vector<int> no_op{1, 2, 3};
  rotate1(no_op.begin(), no_op.begin(), no_op.end());
  assert((no_op == vector<int>{1, 2, 3}));
  rotate2(no_op.begin(), no_op.end(), no_op.end());
  assert((no_op == vector<int>{1, 2, 3}));
  rotate3(no_op.begin(), no_op.begin(), no_op.end());
  assert((no_op == vector<int>{1, 2, 3}));

  // Example from: http://en.cppreference.com/w/cpp/algorithm/rotate
  vector<int> a{2, 4, 2, 0, 5, 10, 7, 3, 7, 1};
  cout << "before sort:  ";
  for (const auto &x : a) {
    cout << x << " ";
  }
  cout << endl;

  // Insertion sort.
  for (auto i = a.begin(); i != a.end(); ++i) {
    rotate1(std::upper_bound(a.begin(), i, *i), i, i + 1);
  }
  cout << "after sort:   ";
  for (const auto &x : a) {
    cout << x << " ";
  }
  cout << endl;

  // Simple rotation to the left.
  rotate2(a.begin(), a.begin() + 1, a.end());
  cout << "rotate left:  ";
  for (const auto &x : a) {
    cout << x << " ";
  }
  cout << endl;

  // Simple rotation to the right.
  rotate3(a.rbegin(), a.rbegin() + 1, a.rend());
  cout << "rotate right: ";
  for (const auto &x : a) {
    cout << x << " ";
  }
  cout << endl;
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
before sort:  2 4 2 0 5 10 7 3 7 1
after sort:   0 1 2 2 3 4 5 7 7 10
rotate left:  1 2 2 3 4 5 7 7 10 0
rotate right: 0 1 2 2 3 4 5 7 7 10
\end{exampleoutput}
\subsection{Counting Inversions}
The number of inversions for a sequence \inlinecode{a} is the number of ordered pairs $(i, j)$ such that $i < j$ and $a[i] > a[j]$. This is roughly how ``close'' an array is to being sorted, but is \textit{not} the minimum number of swaps required to sort the array. If the array is sorted, then the inversion count is $0$. If the array is sorted in decreasing order, then the inversion count is maximal. The following two functions are each techniques to efficiently count inversions. In the merge sort approach, whenever the merge step emits an element from the right half, that element jumps ahead of every unmerged left-half element, and exactly that many inversions are added to the count.

\begin{compactitem}
  \item \inlinecode{inversions(lo, hi)} uses merge sort to return the number of inversions given two random-access iterators as a range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$. The input range will be sorted after the function call. Optionally, a comparison function object specifying a strict weak ordering may be specified to replace the default \inlinecode{operator\textless{}}.
  \item \inlinecode{inversions(a)} uses coordinate compression and a Fenwick tree to return the number of inversions in an integer vector without modifying it.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log n)$ per call to \inlinecode{inversions(lo, hi)}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(n \log n)$ per call to \inlinecode{inversions(a)}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary space and $O(\log n)$ stack space for \inlinecode{inversions(lo, hi)}.
  \item $O(n)$ auxiliary heap space for \inlinecode{inversions(a)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstdint>
#include <functional>
#include <iterator>
#include <vector>

template<typename It, typename Compare = std::less<>>
int64_t inversions(It lo, It hi, Compare comp = Compare()) {
  if (hi - lo < 2) {
    return 0;
  }
  It mid = lo + (hi - lo - 1) / 2, a = lo, c = mid + 1;
  int64_t res = 0;
  res += inversions(lo, mid + 1, comp);
  res += inversions(mid + 1, hi, comp);
  using T = typename std::iterator_traits<It>::value_type;
  std::vector<T> merged;
  while (a <= mid && c < hi) {
    if (comp(*c, *a)) {
      merged.push_back(*(c++));
      res += (mid - a) + 1;
    } else {
      merged.push_back(*(a++));
    }
  }
  if (a > mid) {
    for (It it = c; it != hi; ++it) {
      merged.push_back(*it);
    }
  } else {
    for (It it = a; it <= mid; ++it) {
      merged.push_back(*it);
    }
  }
  for (It it = lo; it != hi; ++it) {
    *it = merged[it - lo];
  }
  return res;
}

int64_t inversions(const std::vector<int> &a) {
  int n = static_cast<int>(a.size());
  std::vector<int> values(a.begin(), a.end());
  std::sort(values.begin(), values.end());
  values.erase(std::unique(values.begin(), values.end()), values.end());
  std::vector<int> bit(values.size() + 1, 0);
  int64_t res = 0;
  for (int i = n - 1; i >= 0; i--) {
    int id =
        static_cast<int>(std::lower_bound(values.begin(), values.end(), a[i]) - values.begin()) + 1;
    for (int j = id - 1; j > 0; j -= j & -j) {
      res += bit[j];
    }
    for (int j = id; j < static_cast<int>(bit.size()); j += j & -j) {
      bit[j]++;
    }
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  {
    std::vector<int> a{6, 9, 1, 14, 8, 12, 3, 2};
    assert(inversions(a.begin(), a.end()) == 16);
    assert(std::is_sorted(a.begin(), a.end()));
  }
  {
    std::vector<int> a{6, 9, 1, 14, 8, 12, 3, 2};
    assert(inversions(a) == 16);
    assert((a == std::vector<int>{6, 9, 1, 14, 8, 12, 3, 2}));
  }
  {
    std::vector<int> a{2, 2, 1};
    assert(inversions(a) == 2);  // Equal elements are not inversions.
  }
  return 0;
}
\end{lstlisting}
\subsection{Coordinate Compression}
Given a range of $n$ numerical elements, reassign each element to an integer in the domain $[0, k)$, where $k$ is the number of distinct elements in the original range, while preserving the initial relative ordering of elements. That is, if $a$ is the array of original values and $b$ is the array of compressed values, then every pair of indices $i, j$ in $[0, n)$ shall satisfy $a[i] < a[j]$ if and only if $b[i] < b[j]$.

All implementations below take ranges as ForwardIterators and require \inlinecode{operator\textless{}} on the value type. The two \inlinecode{compress()} functions rewrite a range in place and then discard the mapping.

\begin{compactitem}
  \item \inlinecode{compress1(lo, hi)} performs the compression by sorting the array, removing duplicates, and binary searching for the position of each original value.
  \item \inlinecode{compress2(lo, hi)} achieves the same result by inserting all values into an \inlinecode{std::map}, which automatically removes duplicate values and supports efficient lookups of the compressed values.
\end{compactitem}

\inlinecode{CoordinateCompressor} is a class version that retains the sorted table of distinct values so that arbitrary values can be mapped to and from compressed ranks long after construction (e.g. for offline queries arriving separately from the array being compressed).

\begin{compactitem}
  \item \inlinecode{CoordinateCompressor\textless{}T\textgreater{}(lo, hi)} constructs a compressor over the values of $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, ready for querying. Alternatively, use the default constructor and register values with \inlinecode{add(x)}, then call \inlinecode{build()} once after all values are registered and before the first query.
  \item \inlinecode{size()} returns the number of distinct registered values $k$.
  \item \inlinecode{value(r)} returns the original value with rank \inlinecode{r}, inverting \inlinecode{rank()}.
  \item \inlinecode{rank(x)} returns the compressed value (rank) of \inlinecode{x} in $[0, k)$. \inlinecode{x} must have been registered.
  \item \inlinecode{contains(x)} returns whether \inlinecode{x} is a registered value.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log n)$ per call to \inlinecode{compress1(lo, hi)}, \inlinecode{compress2(lo, hi)}, and \inlinecode{CoordinateCompressor(lo, hi)}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(1)$ amortized per call to \inlinecode{add(x)}, and $O(m \log m)$ per call to \inlinecode{build()}, where $m$ is the total number of values registered.
  \item $O(\log k)$ per call to \inlinecode{rank(x)} and \inlinecode{contains(x)}, where $k$ is the number of distinct values.
  \item $O(1)$ per call to \inlinecode{value(r)} and \inlinecode{size()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary heap space for \inlinecode{compress1()} and \inlinecode{compress2()}.
  \item $O(k)$ storage for \inlinecode{CoordinateCompressor}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <iterator>
#include <map>
#include <vector>

template<typename It>
void compress1(It lo, It hi) {
  using T = typename std::iterator_traits<It>::value_type;
  std::vector<T> v(lo, hi);
  std::sort(v.begin(), v.end());
  v.resize(std::unique(v.begin(), v.end()) - v.begin());
  for (It it = lo; it != hi; ++it) {
    *it = static_cast<int>(std::lower_bound(v.begin(), v.end(), *it) - v.begin());
  }
}

template<typename It>
void compress2(It lo, It hi) {
  using T = typename std::iterator_traits<It>::value_type;
  std::map<T, int> m;
  for (It it = lo; it != hi; ++it) {
    m[*it] = 0;
  }
  int id = 0;
  for (auto &[key, val] : m) {
    val = id++;
  }
  for (It it = lo; it != hi; ++it) {
    *it = m[*it];
  }
}

template<typename T>
class CoordinateCompressor {
  std::vector<T> values;

 public:
  CoordinateCompressor() {}

  template<typename It>
  CoordinateCompressor(It lo, It hi) : values(lo, hi) {
    build();
  }

  int size() const { return static_cast<int>(values.size()); }
  const T &value(int r) const { return values[r]; }
  void add(const T &x) { values.push_back(x); }

  void build() {
    std::sort(values.begin(), values.end());
    values.resize(std::unique(values.begin(), values.end()) - values.begin());
  }

  int rank(const T &x) const {
    int r = static_cast<int>(std::lower_bound(values.begin(), values.end(), x) - values.begin());
    assert(r < size() && !(x < values[r]));  // x must be registered.
    return r;
  }

  bool contains(const T &x) const { return std::binary_search(values.begin(), values.end(), x); }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <iostream>
using namespace std;

template<typename It>
void print_range(It lo, It hi) {
  while (lo != hi) {
    cout << *lo++ << " ";
  }
  cout << endl;
}

int main() {
  {
    vector<int> a{1, 30, 30, 7, 9, 8, 99, 99};
    compress1(a.begin(), a.end());
    print_range(a.begin(), a.end());
  }
  {
    vector<int> a{1, 30, 30, 7, 9, 8, 99, 99};
    compress2(a.begin(), a.end());
    print_range(a.begin(), a.end());
  }
  {  // Non-integral types work too, as long as ints can be assigned to them.
    vector<double> a{0.5, -1.0, 3, -1.0, 20, 0.5};
    compress1(a.begin(), a.end());
    print_range(a.begin(), a.end());
  }
  {
    vector<int> a{10, 5, 30, 5, 20};
    CoordinateCompressor<int> cc(a.begin(), a.end());
    assert(cc.size() == 4);
    assert(cc.rank(5) == 0);
    assert(cc.rank(30) == 3);
    assert(cc.value(2) == 20);
    assert(cc.contains(10) && !cc.contains(7));
  }
  {  // Register values incrementally, then build once before querying.
    CoordinateCompressor<double> cc;
    cc.add(0.5);
    cc.add(-1.0);
    cc.add(0.5);
    cc.build();
    assert(cc.size() == 2);
    assert(cc.rank(0.5) == 1);
    assert(cc.value(0) == -1.0);
  }
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
0 4 4 1 3 2 5 5
0 4 4 1 3 2 5 5
1 0 2 0 3 1
\end{exampleoutput}
\subsection{Partitioning and Selection}
Partition a range around a pivot, or select the element with a given sorted rank without fully sorting the range. Three-way partitioning, also known as the Dutch National Flag algorithm, rearranges the elements into consecutive groups less than, equal to, and greater than the pivot. A single pass maintains these three groups plus an unclassified region. When an element greater than the pivot is swapped with the end of the unclassified region, the incoming element must still be examined before the scan advances.

Quickselect repeatedly applies this partition and continues only into the group containing the desired rank. This is the same task as \inlinecode{std::nth\_element()}: after the call, \inlinecode{*nth} is the value that would appear there in sorted order, all earlier positions contain values no larger than it, and all later positions contain values no smaller than it. Choosing pivots uniformly at random gives expected linear time, while the three-way split avoids unnecessary work on duplicate-heavy inputs.

\begin{compactitem}
  \item \inlinecode{partition\_three\_way(lo, hi, pivot)} rearranges $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ in-place and returns iterators $({\inlinecodemath{mid1}}, {\inlinecodemath{mid2}})$, where $[{\inlinecodemath{lo}}, {\inlinecodemath{mid1}})$ is less than \inlinecode{pivot}, $[{\inlinecodemath{mid1}}, {\inlinecodemath{mid2}})$ is equal to \inlinecode{pivot}, and $[{\inlinecodemath{mid2}}, {\inlinecodemath{hi}})$ is greater than \inlinecode{pivot}.
  \item \inlinecode{sort\_012(lo, hi)} sorts a range consisting only of the values $0$, $1$, and $2$.
  \item \inlinecode{nth\_element2(lo, nth, hi)} rearranges the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ around the 0-based rank represented by iterator \inlinecode{nth}. This requires random-access iterators and \inlinecode{operator\textless{}} on the value type.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{partition\_three\_way()} or \inlinecode{sort\_012()}, where $n$ is the range length.
  \item $O(n)$ expected per call to \inlinecode{nth\_element2()}, and $O(n^{2})$ in the worst case.
\end{compactitem}

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <iterator>
#include <random>
#include <utility>

template<typename It, typename T>
std::pair<It, It> partition_three_way(It lo, It hi, const T &pivot) {
  It lt = lo, i = lo, gt = hi;
  while (i != gt) {
    if (*i < pivot) {
      std::iter_swap(lt++, i++);
    } else if (pivot < *i) {
      std::iter_swap(i, --gt);
    } else {
      ++i;
    }
  }
  return {lt, gt};
}

template<typename It>
void sort_012(It lo, It hi) {
  partition_three_way(lo, hi, 1);
}

template<typename It>
void nth_element2(It lo, It nth, It hi) {
  static std::mt19937 rng(std::random_device{}());
  while (hi - lo > 1) {
    std::uniform_int_distribution<int> dist(0, hi - lo - 1);
    auto pivot = *(lo + dist(rng));
    auto [lt, gt] = partition_three_way(lo, hi, pivot);
    if (nth < lt) {
      hi = lt;
    } else if (nth >= gt) {
      lo = gt;
    } else {
      return;
    }
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <vector>
using namespace std;

int main() {
  vector<int> values{2, 0, 1, 2, 1, 0, 1};
  sort_012(values.begin(), values.end());
  for (int i = 0; i < 7; i++) {
    assert(values[i] == (i < 2 ? 0 : i < 5 ? 1 : 2));
  }

  vector<int> b{4, 2, 5, 3, 3, 1};
  auto [mid1, mid2] = partition_three_way(b.begin(), b.end(), 3);
  for (auto it = b.begin(); it != mid1; ++it) {
    assert(*it < 3);
  }
  for (auto it = mid1; it != mid2; ++it) {
    assert(*it == 3);
  }
  for (auto it = mid2; it != b.end(); ++it) {
    assert(*it > 3);
  }

  vector<int> a{5, 6, 4, 3, 2, 6, 7, 9, 3};
  int n = static_cast<int>(a.size());
  nth_element2(a.begin(), a.begin() + n / 2, a.end());
  assert(a[n / 2] == 5);
  // Values left of the median are <=, and values right are >= (the exact order within each side is
  // randomized, since the pivot is chosen at random).
  for (int i = 0; i < n / 2; i++) {
    assert(a[i] <= a[n / 2]);
  }
  for (int i = n / 2 + 1; i < n; i++) {
    assert(a[i] >= a[n / 2]);
  }
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Arrays and Dynamic Programming}{1.2 Arrays and Dynamic Programming}}
\setcounter{section}{2}
\setcounter{subsection}{0}
\subsection{Prefix Sums}
Precomputes prefix sums so range-sum queries can be answered in constant time. This is one of the most common array preprocessing tools, and also appears as a building block for subarray sums, difference arrays, and two-dimensional grids. Each table entry stores the sum of all elements before it, so any range sum is the difference of two entries; in two dimensions, rectangle sums combine four entries by inclusion-exclusion.

\begin{compactitem}
  \item \inlinecode{prefix\_sums(a)} returns array \inlinecode{pref} with \inlinecode{pref[0] = 0} and \inlinecode{pref[i + 1] = a[0] + ... + a[i]}.
  \item \inlinecode{range\_sum(pref, lo, hi)} returns the sum of the inclusive range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}}]$.
  \item \inlinecode{prefix\_sums\_2d(a)} returns a two-dimensional prefix sum table for matrix \inlinecode{a}.
  \item \inlinecode{rectangle\_sum(pref, r1, c1, r2, c2)} returns the sum of the rectangle with top left at $({\inlinecodemath{r1}}, {\inlinecodemath{c1}})$ and bottom right at $({\inlinecodemath{r2}}, {\inlinecodemath{c2}})$.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{prefix\_sums(a)}, where $n$ is the array size.
  \item $O(m \cdot n)$ per call to \inlinecode{prefix\_sums\_2d(a)}, where $m$ and $n$ are the number of rows and columns of \inlinecode{a}, respectively.
  \item $O(1)$ per range or rectangle query.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for \inlinecode{prefix\_sums(a)}.
  \item $O(m \cdot n)$ for \inlinecode{prefix\_sums\_2d(a)}.
  \item $O(1)$ auxiliary for \inlinecode{range\_sum()} and \inlinecode{rectangle\_sum()}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <vector>

std::vector<int64_t> prefix_sums(const std::vector<int> &a) {
  std::vector<int64_t> pref(a.size() + 1, 0);
  for (int i = 0; i < static_cast<int>(a.size()); i++) {
    pref[i + 1] = pref[i] + a[i];
  }
  return pref;
}

int64_t range_sum(const std::vector<int64_t> &pref, int lo, int hi) {
  return pref[hi + 1] - pref[lo];
}

std::vector<std::vector<int64_t>> prefix_sums_2d(const std::vector<std::vector<int>> &a) {
  int rows = static_cast<int>(a.size());
  int cols = a.empty() ? 0 : static_cast<int>(a[0].size());
  std::vector<std::vector<int64_t>> pref(rows + 1, std::vector<int64_t>(cols + 1, 0));
  for (int i = 0; i < rows; i++) {
    for (int j = 0; j < cols; j++) {
      pref[i + 1][j + 1] = a[i][j] + pref[i][j + 1] + pref[i + 1][j] - pref[i][j];
    }
  }
  return pref;
}

int64_t rectangle_sum(
    const std::vector<std::vector<int64_t>> &pref, int r1, int c1, int r2, int c2
) {
  return pref[r2 + 1][c2 + 1] - pref[r1][c2 + 1] - pref[r2 + 1][c1] + pref[r1][c1];
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> a{3, -1, 4, 1, 5};
  auto pref = prefix_sums(a);
  assert(range_sum(pref, 0, 4) == 12);  // Whole array a[0..4]
  assert(range_sum(pref, 1, 3) == 4);   // -1 + 4 + 1
  assert(range_sum(pref, 2, 2) == 4);   // Single element.

  vector<vector<int>> grid{
    {1, 2, 3},
    {4, 5, 6}
  };
  auto pre2 = prefix_sums_2d(grid);
  assert(rectangle_sum(pre2, 0, 1, 1, 2) == 16);  // rows 0..1, cols 1..2
  assert(rectangle_sum(pre2, 0, 0, 1, 2) == 21);  // Whole grid.
  assert(rectangle_sum(pre2, 1, 1, 1, 1) == 5);   // Single cell.
  return 0;
}
\end{lstlisting}
\subsection{Difference Array}
Applies many range-add updates to an array in linear total time using a difference array. Instead of immediately adding \inlinecode{delta} to every element of an inclusive range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}}]$, we can add \inlinecode{delta} at \inlinecode{diff[lo]} and subtract it at \inlinecode{diff[hi + 1]}; a final prefix sum reconstructs the updated values. This idea extends to two dimensions: a rectangle update touches just the four corner cells of a difference grid by inclusion-exclusion, and a final two-dimensional prefix sum reconstructs the grid.

\begin{compactitem}
  \item \inlinecode{DifferenceArray(n)} constructs an initially zero array of size \inlinecode{n}.
  \item \inlinecode{add(lo, hi, delta)} adds \inlinecode{delta} to every position in the inclusive range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}}]$.
  \item \inlinecode{build()} returns the final array after all range updates.
  \item \inlinecode{DifferenceArray2D(rows, cols)} constructs an initially zero grid with \inlinecode{rows} rows and \inlinecode{cols} columns.
  \item \inlinecode{add(r1, c1, r2, c2, delta)} adds \inlinecode{delta} to every cell of the rectangle with top left at $({\inlinecodemath{r1}}, {\inlinecodemath{c1}})$ and bottom right at $({\inlinecodemath{r2}}, {\inlinecodemath{c2}})$, inclusive.
  \item \inlinecode{build()} returns the final grid after all rectangle updates.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ per call to \inlinecode{add()} of either version.
  \item $O(n)$ per call to \inlinecode{build()} of \inlinecode{DifferenceArray}, where $n$ is the array size.
  \item $O(m \cdot n)$ per call to \inlinecode{build()} of \inlinecode{DifferenceArray2D}, where $m$ and $n$ are the number of rows and columns, respectively.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for the one-dimensional difference array.
  \item $O(m \cdot n)$ for the two-dimensional difference grid.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <vector>

class DifferenceArray {
  std::vector<int64_t> diff;

 public:
  explicit DifferenceArray(int n) : diff(n + 1, 0) {}

  void add(int lo, int hi, int64_t delta) {
    diff[lo] += delta;
    diff[hi + 1] -= delta;
  }

  std::vector<int64_t> build() const {
    std::vector<int64_t> res(static_cast<int>(diff.size()) - 1);
    int64_t cur = 0;
    for (int i = 0; i < static_cast<int>(res.size()); i++) {
      cur += diff[i];
      res[i] = cur;
    }
    return res;
  }
};

class DifferenceArray2D {
  std::vector<std::vector<int64_t>> diff;

 public:
  DifferenceArray2D(int rows, int cols) : diff(rows + 1, std::vector<int64_t>(cols + 1, 0)) {}

  void add(int r1, int c1, int r2, int c2, int64_t delta) {
    diff[r1][c1] += delta;
    diff[r1][c2 + 1] -= delta;
    diff[r2 + 1][c1] -= delta;
    diff[r2 + 1][c2 + 1] += delta;
  }

  std::vector<std::vector<int64_t>> build() const {
    int rows = static_cast<int>(diff.size()) - 1;
    int cols = static_cast<int>(diff[0].size()) - 1;
    std::vector<std::vector<int64_t>> res(rows, std::vector<int64_t>(cols));
    for (int i = 0; i < rows; i++) {
      for (int j = 0; j < cols; j++) {
        int64_t up = i > 0 ? res[i - 1][j] : 0;
        int64_t left = j > 0 ? res[i][j - 1] : 0;
        int64_t diag = i > 0 && j > 0 ? res[i - 1][j - 1] : 0;
        res[i][j] = diff[i][j] + up + left - diag;
      }
    }
    return res;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  DifferenceArray d(5);
  d.add(0, 2, 2);   // a[0..2] += 2
  d.add(1, 4, 4);   // a[1..4] += 4
  d.add(2, 3, -1);  // a[2..3] -= 1
  d.add(4, 4, 3);   // Single index update.
  vector<int64_t> a = d.build();
  vector<int64_t> expected{2, 6, 5, 3, 7};
  for (int i = 0; i < static_cast<int>(expected.size()); i++) {
    assert(a[i] == expected[i]);
  }

  DifferenceArray2D d2(3, 4);
  d2.add(0, 0, 1, 2, 1);  // Rows 0..1, cols 0..2 += 1
  d2.add(1, 1, 2, 3, 2);  // Rows 1..2, cols 1..3 += 2
  d2.add(2, 0, 2, 0, 5);  // Single cell update.
  vector<vector<int64_t>> g = d2.build();
  vector<vector<int64_t>> expected2{
    {1, 1, 1, 0},
    {1, 3, 3, 2},
    {5, 2, 2, 2}
  };
  assert(g == expected2);
  return 0;
}
\end{lstlisting}
\subsection{Two Pointers and Sliding Window}
Uses two moving indices to avoid trying all pairs of positions. This technique has two common forms: opposite-end pointers over sorted arrays, and a sliding window over contiguous subarrays. For sorted 2-sum, if the current sum is too small, pairing the left value with any smaller right value also fails, so the left pointer can safely advance; if the sum is too large, the symmetric argument lets the right pointer retreat. Each step therefore discards an entire row or column of candidate pairs.

A sliding window applies the same idea when extending and shrinking a range changes its state monotonically. With nonnegative values, extending a window cannot decrease its sum and removing its leftmost value cannot increase it, so no shorter feasible window is skipped. For a bound such as at most \inlinecode{k} distinct values, the left pointer advances only until the window becomes valid again. Since neither endpoint moves backward, each element enters and leaves the window at most once.

For fixed-length windows, a monotonic deque of indices can report every minimum or maximum in linear total time. It stores only candidates that could still become the answer, ordered from the current extreme at the front to the least useful value at the back. A new element removes every older value that it dominates, while expired indices are removed from the front. Each index is again pushed and popped at most once.

The examples below cover sorted 2-sum/3-sum variants, finding the shortest subarray with sum at least a target when all values are nonnegative, and finding the longest subarray with at most \inlinecode{k} distinct values. They also cover the minimum or maximum of every fixed-length window and an online dynamic programming recurrence.

\begin{compactitem}
  \item \inlinecode{two\_sum\_sorted(a, target)} returns two indices whose values sum to \inlinecode{target}, or $(-1, -1)$ if no such pair exists. The input array must be sorted.
  \item \inlinecode{three\_sum(a, target)} returns three original indices whose values sum to \inlinecode{target}, or $(-1, -1, -1)$ if none exist.
  \item \inlinecode{min\_length\_at\_least(a, target)} returns $({\inlinecodemath{length}}, {\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, the minimum length and inclusive endpoints of a contiguous subarray with sum at least \inlinecode{target}, or length $-1$ if none exists. Values in \inlinecode{a} must be nonnegative. If \inlinecode{target} is nonpositive, it returns the empty subarray.
  \item \inlinecode{longest\_at\_most\_k\_distinct(a, k)} returns $({\inlinecodemath{length}}, {\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, the maximum length and inclusive endpoints of a contiguous subarray containing at most \inlinecode{k} distinct values.
  \item \inlinecode{sliding\_window\_extrema(a, k, comp)} returns the extreme value in each window of length \inlinecode{k}. With the default \inlinecode{less\textless{}\textgreater{}} comparator it returns minimums; passing \inlinecode{greater\textless{}\textgreater{}} returns maximums.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{two\_sum\_sorted()} and \inlinecode{min\_length\_at\_least()}, where $n$ is the array size.
  \item $O(n^{2})$ per call to \inlinecode{three\_sum()}.
  \item $O(n)$ expected per call to \inlinecode{longest\_at\_most\_k\_distinct()}.
  \item $O(n)$ per call to \inlinecode{sliding\_window\_extrema()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for \inlinecode{two\_sum\_sorted(a, target)}.
  \item $O(n)$ auxiliary heap space for \inlinecode{three\_sum(a, target)}.
  \item $O(1)$ auxiliary for \inlinecode{min\_length\_at\_least(a, target)}.
  \item $O(k)$ auxiliary heap space for \inlinecode{longest\_at\_most\_k\_distinct(a, k)}.
  \item $O(n)$ auxiliary heap space for the result of \inlinecode{sliding\_window\_extrema()}, plus $O(k)$ for the deque.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <array>
#include <cassert>
#include <cstdint>
#include <deque>
#include <functional>
#include <tuple>
#include <unordered_map>
#include <utility>
#include <vector>

std::pair<int, int> two_sum_sorted(const std::vector<int> &a, int64_t target) {
  for (int l = 0, r = static_cast<int>(a.size()) - 1; l < r;) {
    int64_t sum = static_cast<int64_t>(a[l]) + a[r];
    if (sum == target) {
      return {l, r};
    } else if (sum < target) {
      l++;
    } else {
      r--;
    }
  }
  return {-1, -1};
}

std::array<int, 3> three_sum(const std::vector<int> &a, int64_t target) {
  int n = static_cast<int>(a.size());
  std::vector<std::pair<int, int>> sorted;
  sorted.reserve(a.size());
  for (int i = 0; i < n; i++) {
    sorted.push_back({a[i], i});
  }
  std::sort(sorted.begin(), sorted.end());
  for (int i = 0; i < n; i++) {
    int l = i + 1, r = n - 1;
    while (l < r) {
      int64_t sum = static_cast<int64_t>(sorted[i].first) + sorted[l].first + sorted[r].first;
      if (sum == target) {
        return {sorted[i].second, sorted[l].second, sorted[r].second};
      } else if (sum < target) {
        l++;
      } else {
        r--;
      }
    }
  }
  return {-1, -1, -1};
}

std::tuple<int, int, int> min_length_at_least(const std::vector<int> &a, int64_t target) {
  if (target <= 0) {
    return {0, 0, -1};
  }
  int best = static_cast<int>(a.size()) + 1, best_lo = 0, best_hi = -1;
  int64_t sum = 0;
  for (int l = 0, r = 0; r < static_cast<int>(a.size()); r++) {
    sum += a[r];
    while (sum >= target) {
      if (r - l + 1 < best) {
        best = r - l + 1;
        best_lo = l;
        best_hi = r;
      }
      sum -= a[l++];
    }
  }
  if (best == static_cast<int>(a.size()) + 1) {
    return {-1, 0, -1};
  }
  return {best, best_lo, best_hi};
}

std::tuple<int, int, int> longest_at_most_k_distinct(const std::vector<int> &a, int k) {
  if (k <= 0) {
    return {0, 0, -1};
  }
  std::unordered_map<int, int> count;
  int best = 0, best_lo = 0, best_hi = -1;
  for (int l = 0, r = 0; r < static_cast<int>(a.size()); r++) {
    count[a[r]]++;
    while (static_cast<int>(count.size()) > k) {
      if (--count[a[l]] == 0) {
        count.erase(a[l]);
      }
      l++;
    }
    if (best < r - l + 1) {
      best = r - l + 1;
      best_lo = l;
      best_hi = r;
    }
  }
  return {best, best_lo, best_hi};
}

template<typename T, typename Compare = std::less<>>
std::vector<T> sliding_window_extrema(const std::vector<T> &a, int k, Compare comp = Compare()) {
  int n = static_cast<int>(a.size());
  assert(1 <= k && k <= n);
  std::deque<int> window;
  std::vector<T> res;
  res.reserve(n - k + 1);
  for (int i = 0; i < n; i++) {
    while (!window.empty() && !comp(a[window.back()], a[i])) {
      window.pop_back();
    }
    window.push_back(i);
    if (window.front() <= i - k) {
      window.pop_front();
    }
    if (i >= k - 1) {
      res.push_back(a[window.front()]);
    }
  }
  return res;
}
\end{lstlisting}
A monotone queue can be used to maintain the minimum or maximum value in an online sliding window. This is useful for dynamic programming recurrences where each transition may only come from one of the last $w$ states, such as \inlinecode{dp[i] = a[i] + min(dp[j])} for $j \in [i - w, i - 1]$.

The queue stores candidate (\inlinecode{index}, \inlinecode{value}) pairs in monotone order. Expired indices are removed from the front, and dominated values are removed from the back before inserting a new candidate.

\begin{compactitem}
  \item \inlinecode{MonotoneQueue\textless{}T, Compare\textgreater{}()} constructs an empty queue. Use \inlinecode{std::less\textless{}T\textgreater{}} for minimum queries and \inlinecode{std::greater\textless{}T\textgreater{}} for maximum queries.
  \item \inlinecode{push(index, value)} inserts candidate \inlinecode{value} at position \inlinecode{index}, removing dominated candidates from the back.
  \item \inlinecode{expire(first\_valid)} removes candidates with index less than \inlinecode{first\_valid}.
  \item \inlinecode{top()} returns the best active (\inlinecode{index}, \inlinecode{value}) pair. The queue must be nonempty.
  \item \inlinecode{empty()} returns whether the queue has no active candidates.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ for any sequence of $n$ calls to \inlinecode{push(index, value)} and \inlinecode{expire(first\_valid)}.
  \item $O(1)$ amortized per call to \inlinecode{push(index, value)} and \inlinecode{expire(first\_valid)}.
  \item $O(1)$ worst-case per call to \inlinecode{top()} and \inlinecode{empty()}.
\end{compactitem}

\textbf{Space Complexity}: $O(w)$ auxiliary heap space for a sliding window of width $w$.

\begin{lstlisting}[language=C++]
template<typename T, typename Compare>
class MonotoneQueue {
  std::deque<std::pair<int, T>> q;
  Compare better;

 public:
  bool empty() const { return q.empty(); }

  void push(int index, const T &value) {
    while (!q.empty() && !better(q.back().second, value)) {
      q.pop_back();
    }
    q.emplace_back(index, value);
  }

  void expire(int first_valid) {
    while (!q.empty() && q.front().first < first_valid) {
      q.pop_front();
    }
  }

  std::pair<int, T> top() const {
    assert(!q.empty());
    return q.front();
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
using namespace std;

int main() {
  vector<int> c{-3, -1, 2, 4, 8, 11};
  assert(two_sum_sorted(c, 10) == make_pair(1, 5));  // -1 + 11.
  assert(two_sum_sorted(c, 7) == make_pair(1, 4));   // -1 + 8.
  assert(two_sum_sorted(c, 100) == make_pair(-1, -1));

  vector<int> d{4, -1, 8, 2, -3, 11};
  array<int, 3> triple = three_sum(d, 9);
  assert(d[triple[0]] + d[triple[1]] + d[triple[2]] == 9);
  assert((three_sum(d, 30) == array<int, 3>{-1, -1, -1}));

  vector<int> a{2, 3, 1, 2, 4, 3};
  auto [length, lo, hi] = min_length_at_least(a, 7);
  assert(length == 2);
  assert(lo == 4 && hi == 5);  // [4, 3].
  auto [empty_length, empty_lo, empty_hi] = min_length_at_least(a, 0);
  assert(empty_length == 0);
  assert(empty_lo == 0 && empty_hi == -1);

  vector<int> b{1, 2, 1, 3, 4, 3, 5};
  auto [longest, longest_lo, longest_hi] = longest_at_most_k_distinct(b, 2);
  assert(longest == 3);
  assert(longest_lo == 0 && longest_hi == 2);  // [1, 2, 1].

  vector<int> e{1, 3, -1, -3, 5, 3, 6, 7};
  assert((sliding_window_extrema(e, 3) == vector<int>{-1, -3, -3, -3, 3, 3}));
  assert((sliding_window_extrema(e, 3, greater<>()) == vector<int>{3, 3, 5, 5, 6, 7}));

  // Toy DP: dp[i] = f[i] + min(dp[j]) for j in [i - max_jump, i - 1].
  // Scanning the window costs O(max_jump) per state; the monotone queue makes it O(1) amortized.
  vector<int> f{4, 2, 7, 1, 3, 6};
  int max_jump = 3;
  vector<int> dp(f.size());
  MonotoneQueue<int, less<int>> best;
  dp[0] = f[0];
  best.push(0, dp[0]);
  for (int i = 1; i < static_cast<int>(f.size()); i++) {
    best.expire(i - max_jump);
    dp[i] = f[i] + best.top().second;
    best.push(i, dp[i]);
  }
  assert(dp[5] == 11);
  return 0;
}
\end{lstlisting}
\subsection{Monotonic Stack}
A monotonic stack maintains its elements in sorted order as values are pushed, popping away any entries that would violate the ordering. Scanning an array once while keeping such a stack of indices answers, for every position, where the nearest smaller or larger neighbor lies. These ``nearest smaller/greater'' relationships underlie many array problems, the most famous being the largest rectangle in a histogram. Each index is pushed and popped at most once, so a full scan runs in linear time. A largest all-zero submatrix can then be found by sweeping rows, turning each row into a histogram of consecutive zeros ending at that row.

For a histogram bar \inlinecode{i}, the nearest strictly lower bars on its left and right are exactly the first positions that prevent a rectangle of height \inlinecode{heights[i]} from extending farther. Its widest such rectangle therefore spans the positions between those boundaries. For a binary matrix, the height at each column counts consecutive zeros ending in the current row; every all-zero rectangle has some bottom row, so solving one histogram per row considers every possible rectangle.

All functions below take a vector \inlinecode{a} of $n$ comparable values and return a vector of $n$ indices. A ``less'' query uses a strictly smaller neighbor and a ``greater'' query uses a strictly larger one; changing the comparison from strict to non-strict (e.g. \inlinecode{\textgreater{}=} to \inlinecode{\textgreater{}}) toggles how ties are handled.

\begin{compactitem}
  \item \inlinecode{previous\_less(a)} returns, for each $i$, the largest index $j < i$ with $a[j] < a[i]$, or $-1$ if there's no such index.
  \item \inlinecode{next\_less(a)} returns, for each $i$, the smallest index $j > i$ with $a[j] < a[i]$, or $n$ if there's no such index.
  \item \inlinecode{previous\_greater(a)} returns, for each $i$, the largest index $j < i$ with $a[j] > a[i]$, or $-1$ if there's no such index.
  \item \inlinecode{next\_greater(a)} returns, for each $i$, the smallest index $j > i$ with $a[j] > a[i]$, or $n$ if there's no such index.
  \item \inlinecode{largest\_rectangle(heights)} returns the maximum area of an axis-aligned rectangle that fits under the given histogram, given an array of \inlinecode{heights} where each bar has width $1$.
  \item \inlinecode{largest\_zero\_submatrix(a)} returns the area of the largest all-zero rectangular submatrix of the 0/1 matrix \inlinecode{a}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to all one-dimensional functions, where $n$ is the size of the input.
  \item $O(r \cdot c)$ per call to \inlinecode{largest\_zero\_submatrix()}, where $r$ and $c$ are the matrix dimensions.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary heap space for the stack and result.
  \item $O(c)$ auxiliary heap space for \inlinecode{largest\_zero\_submatrix()}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <stack>
#include <vector>

template<typename T>
std::vector<int> previous_less(const std::vector<T> &a) {
  int n = static_cast<int>(a.size());
  std::vector<int> res(n);
  std::stack<int> s;
  for (int i = 0; i < n; i++) {
    while (!s.empty() && a[s.top()] >= a[i]) {
      s.pop();
    }
    res[i] = s.empty() ? -1 : s.top();
    s.push(i);
  }
  return res;
}

template<typename T>
std::vector<int> next_less(const std::vector<T> &a) {
  int n = static_cast<int>(a.size());
  std::vector<int> res(n);
  std::stack<int> s;
  for (int i = n - 1; i >= 0; i--) {
    while (!s.empty() && a[s.top()] >= a[i]) {
      s.pop();
    }
    res[i] = s.empty() ? n : s.top();
    s.push(i);
  }
  return res;
}

template<typename T>
std::vector<int> previous_greater(const std::vector<T> &a) {
  int n = static_cast<int>(a.size());
  std::vector<int> res(n);
  std::stack<int> s;
  for (int i = 0; i < n; i++) {
    while (!s.empty() && a[s.top()] <= a[i]) {
      s.pop();
    }
    res[i] = s.empty() ? -1 : s.top();
    s.push(i);
  }
  return res;
}

template<typename T>
std::vector<int> next_greater(const std::vector<T> &a) {
  int n = static_cast<int>(a.size());
  std::vector<int> res(n);
  std::stack<int> s;
  for (int i = n - 1; i >= 0; i--) {
    while (!s.empty() && a[s.top()] <= a[i]) {
      s.pop();
    }
    res[i] = s.empty() ? n : s.top();
    s.push(i);
  }
  return res;
}

template<typename T>
T largest_rectangle(const std::vector<T> &heights) {
  int n = static_cast<int>(heights.size());
  std::vector<int> left = previous_less(heights), right = next_less(heights);
  T best = 0;
  for (int i = 0; i < n; i++) {
    T area = heights[i] * (right[i] - left[i] - 1);
    if (area > best) {
      best = area;
    }
  }
  return best;
}

int largest_zero_submatrix(const std::vector<std::vector<char>> &a) {
  if (a.empty()) {
    return 0;
  }
  int rows = static_cast<int>(a.size()), cols = static_cast<int>(a[0].size());
  for (const auto &row : a) {
    assert(static_cast<int>(row.size()) == cols);
  }
  int best = 0;
  std::vector<int> height(cols, 0);
  for (int i = 0; i < rows; i++) {
    for (int j = 0; j < cols; j++) {
      height[j] = a[i][j] ? 0 : height[j] + 1;
    }
    best = std::max(best, largest_rectangle(height));
  }
  return best;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
using namespace std;

int main() {
  vector<int> a{2, 1, 4, 3};
  assert((previous_less(a) == vector<int>{-1, -1, 1, 1}));
  assert((next_less(a) == vector<int>{1, 4, 3, 4}));
  assert((previous_greater(a) == vector<int>{-1, 0, -1, 2}));
  assert((next_greater(a) == vector<int>{2, 2, 4, 4}));

  vector<int> hist{2, 1, 5, 6, 2, 3};
  // The best histogram rectangle uses bars 2 and 3: min height 5 times width 2.
  assert(largest_rectangle(hist) == 10);

  vector<vector<char>> grid{
      {1, 0, 1, 1, 0, 0},
      {1, 0, 0, 1, 0, 0},
      {0, 0, 0, 0, 0, 1},
      {1, 0, 0, 1, 0, 0},
      {1, 0, 1, 0, 0, 1}
  };
  // The largest zero rectangle has 3 rows and 2 columns.
  assert(largest_zero_submatrix(grid) == 6);
  return 0;
}
\end{lstlisting}
\subsection{Aggregate Queue}
Maintain a first-in, first-out queue that additionally answers, in $O(1)$ amortized time, the fold of an associative operation over all elements currently in the queue. This is the ``sliding window aggregation'' (SWAG) technique: by pushing new elements and popping old ones as a window advances, it reports an aggregate of every window in linear total time.

Unlike the monotonic deque used for sliding-window extrema (see 1.2.3), the operation need only be associative, not order-based, so the same structure handles min, max, sum, gcd, matrix products, and other monoids. It is sometimes called a ``monotonic queue,'' but that term usually denotes the monotonic deque of 1.2.3 (which handles only min/max); the two are different structures. The queue is built from two stacks: a back stack receives pushes while a front stack serves pops, and each stack stores running folds so the two halves combine in $O(1)$. Whenever the front stack empties, the back stack is flipped onto it; this costs $O(1)$ amortized per element, since each element is moved between stacks at most once.

The fold is defined by an associative \inlinecode{combine(a, b)} function. The default code below returns the ``min'' of the elements; for ``sum'', \inlinecode{combine(a, b)} should return \inlinecode{a + b}. The fold is taken in queue order (front to back), so non-commutative operations such as matrix products are also supported.

\begin{compactitem}
  \item \inlinecode{AggregateQueue\textless{}T\textgreater{}()} constructs an empty queue.
  \item \inlinecode{empty()} returns whether the queue is empty.
  \item \inlinecode{size()} returns the number of elements in the queue.
  \item \inlinecode{push(x)} appends \inlinecode{x} to the back of the queue.
  \item \inlinecode{pop()} removes the element at the front of the queue, which must be non-empty.
  \item \inlinecode{front()} returns the element at the front of the queue, which must be non-empty.
  \item \inlinecode{fold()} returns \inlinecode{combine()} applied to all elements in queue order; the queue must be non-empty.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ per call to the constructor, \inlinecode{empty()}, \inlinecode{size()}, \inlinecode{front()}, and \inlinecode{fold()}.
  \item $O(1)$ amortized per call to \inlinecode{push()} and \inlinecode{pop()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for storage of the queue elements, where $n$ is the number of elements.
  \item $O(1)$ auxiliary per operation, amortized (a \inlinecode{pop()} may flip the back stack onto the front stack).
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <utility>
#include <vector>

template<typename T>
class AggregateQueue {
  static T combine(const T &a, const T &b) { return std::min(a, b); }

  // Each stack entry is (value, fold), where fold combines the value with all entries between it
  // and its end of the queue (the front for front_stack, the back for back_stack).
  std::vector<std::pair<T, T>> front_stack, back_stack;

 public:
  bool empty() const { return front_stack.empty() && back_stack.empty(); }
  int size() const { return static_cast<int>(front_stack.size() + back_stack.size()); }

  void push(const T &x) {
    T acc = back_stack.empty() ? x : combine(back_stack.back().second, x);
    back_stack.emplace_back(x, acc);
  }

  void pop() {
    assert(!empty());
    if (front_stack.empty()) {
      // Flip the back stack onto the front stack, reversing order so the oldest element ends up on
      // top, and recompute the front folds in queue order.
      while (!back_stack.empty()) {
        T x = back_stack.back().first;
        back_stack.pop_back();
        T acc = front_stack.empty() ? x : combine(x, front_stack.back().second);
        front_stack.emplace_back(x, acc);
      }
    }
    front_stack.pop_back();
  }

  T front() const {
    assert(!empty());
    return front_stack.empty() ? back_stack.front().first : front_stack.back().first;
  }

  T fold() const {
    assert(!empty());
    if (front_stack.empty()) {
      return back_stack.back().second;
    }
    if (back_stack.empty()) {
      return front_stack.back().second;
    }
    return combine(front_stack.back().second, back_stack.back().second);
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
using namespace std;

int main() {
  // Used directly as a FIFO queue with an O(1) min query.
  AggregateQueue<int> q;
  assert(q.empty());
  q.push(5);
  q.push(3);
  q.push(8);
  assert(!q.empty());
  assert(q.front() == 5 && q.size() == 3);
  assert(q.fold() == 3);  // min(5, 3, 8)
  q.pop();                // Removes 5.
  assert(q.fold() == 3);  // min(3, 8)
  q.pop();                // Removes 3.
  assert(q.fold() == 8);
  q.pop();  // Removes 8.
  assert(q.empty());

  // Canonical use: the minimum of every width-3 window, advancing the window by push/pop.
  vector<int> a{4, 2, 5, 1, 3, 6};
  AggregateQueue<int> w;
  vector<int> mins;
  for (int i = 0; i < static_cast<int>(a.size()); i++) {
    w.push(a[i]);
    if (w.size() > 3) {
      w.pop();
    }
    if (w.size() == 3) {
      mins.push_back(w.fold());
    }
  }
  assert((mins == vector<int>{2, 1, 1, 1}));  // min of [4,2,5], [2,5,1], [5,1,3], [1,3,6]
  return 0;
}
\end{lstlisting}
\subsection{Maximum Subarray Sum (Kadane)}
Given an array of numbers, determine the maximum possible sum of any contiguous subarray. Kadane's algorithm scans the array while maintaining the maximum nonnegative sum of a subarray ending at the current position. A negative running sum is discarded because it can only reduce the sum of every subarray that extends it. Thus, by including \inlinecode{a[i]}, the running sum becomes \inlinecode{max(0, sum + a[i])}, and the largest running sum seen is the answer. This can be adapted to compute the maximal submatrix sum as well.

\begin{compactitem}
  \item \inlinecode{max\_subarray\_sum(lo, hi)} returns \inlinecode{(sum, begin, end)} for the maximal subarray in $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, where \inlinecode{lo} and \inlinecode{hi} are random-access iterators to numeric types. The endpoints are inclusive. This implementation requires operators \inlinecode{+} and \inlinecode{\textless{}} on the iterators' value type. By convention, the empty subarray is allowed, so an input range containing only negative values returns sum $0$ and endpoints $[0, -1]$.
  \item \inlinecode{max\_submatrix\_sum(a)} returns \inlinecode{(sum, r1, c1, r2, c2)} for the largest rectangular submatrix of a matrix \inlinecode{a} with $m$ rows and $n$ columns. The corners are inclusive. This implementation requires operators \inlinecode{+} and \inlinecode{\textless{}} on the matrix value type. By convention, the empty submatrix is allowed, so a matrix containing only negative values returns sum $0$ with empty row and column intervals.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{max\_subarray\_sum()}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(m \cdot n^{2})$ per call to \inlinecode{max\_submatrix\_sum()}, where $m$ is the number of rows and $n$ is the number of columns in the matrix.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for \inlinecode{max\_subarray\_sum()}.
  \item $O(m)$ auxiliary heap space for \inlinecode{max\_submatrix\_sum()}, where $m$ is the number of rows in the matrix.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstddef>
#include <iterator>
#include <tuple>
#include <vector>

template<typename It>
auto max_subarray_sum(It lo, It hi) {
  using T = typename std::iterator_traits<It>::value_type;
  if (lo == hi) {
    return std::make_tuple(T(), 0, -1);
  }
  int curr_begin = 0, begin = 0, end = -1;
  T sum = 0, max_sum = 0;
  for (It it = lo; it != hi; ++it) {
    sum += *it;
    if (sum < 0) {
      sum = 0;
      curr_begin = (it - lo) + 1;
    } else if (max_sum < sum) {
      max_sum = sum;
      begin = curr_begin;
      end = it - lo;
    }
  }
  return std::make_tuple(max_sum, begin, end);
}

template<typename T>
std::tuple<T, int, int, int, int> max_submatrix_sum(const std::vector<std::vector<T>> &a) {
  if (a.empty() || a[0].empty()) {
    return {T(), 0, 0, -1, -1};
  }
  int rows = static_cast<int>(a.size()), cols = static_cast<int>(a[0].size());
  int r1 = 0, c1 = 0, r2 = -1, c2 = -1;
  T max_sum = 0;
  for (int clo = 0; clo < cols; clo++) {
    std::vector<T> sums(rows, 0);
    for (int chi = clo; chi < cols; chi++) {
      for (int i = 0; i < rows; i++) {
        sums[i] += a[i][chi];
      }
      auto [sum, rlo, rhi] = max_subarray_sum(sums.begin(), sums.end());
      if (max_sum < sum) {
        max_sum = sum;
        r1 = rlo;
        c1 = clo;
        r2 = rhi;
        c2 = chi;
      }
    }
  }
  return {max_sum, r1, c1, r2, c2};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <iostream>
using namespace std;

int main() {
  {
    vector<int> a{-2, -1, -3, 4, -1, 2, 1, -5, 4};
    auto [negative_sum, negative_lo, negative_hi] = max_subarray_sum(a.begin(), a.begin() + 3);
    assert(negative_sum == 0 && negative_lo == 0 && negative_hi == -1);
    auto [sum, lo, hi] = max_subarray_sum(a.begin(), a.end());
    assert(sum == 6);
    cout << "Maximal sum subarray:" << endl;
    for (int i = lo; i <= hi; i++) {
      cout << a[i] << " ";
    }
    cout << endl;
  }
  {
    vector<vector<int>> a{{0, -2, -7, 0, 5},
                          {9, 2, -6, 2, -4},
                          {-4, 1, -4, 1, 0},
                          {-1, 8, 0, -2, 3}};
    auto [sum, r1, c1, r2, c2] = max_submatrix_sum(a);
    assert(sum == 15);
    cout << "\nMaximal sum submatrix:" << endl;
    for (int i = r1; i <= r2; i++) {
      for (int j = c1; j <= c2; j++) {
        cout << a[i][j] << " ";
      }
      cout << endl;
    }
  }
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
Maximal sum subarray:
4 -1 2 1

Maximal sum submatrix:
9 2
-4 1
-1 8
\end{exampleoutput}
\subsection{Longest Increasing Subsequence}
Given a range of elements, determine a longest subsequence of the range where all elements are in strictly ascending order. The subsequence is not necessarily contiguous or unique, so only one such answer will be found. This implementation requires random-access iterators and \inlinecode{operator\textless{}} on the value type.

The answer is computed by scanning left to right while maintaining, for each length, the smallest possible tail value of an increasing subsequence of that length. Each element extends or improves one of these tails, located in $O(\log n)$ by binary search; predecessor links then reconstruct the subsequence.

\begin{compactitem}
  \item \inlinecode{longest\_increasing\_subsequence(lo, hi)} returns the indices relative to \inlinecode{lo} of one longest strictly increasing subsequence of $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, in order.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call to either function, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <vector>

template<typename It>
std::vector<int> longest_increasing_subsequence(It lo, It hi) {
  int len = 0, n = static_cast<int>(hi - lo);
  if (n == 0) {
    return {};
  }
  // tail[i] = index (into [lo, hi)) of the last element of the best known LIS of length i+1.
  // prev[i] = index of the predecessor of element i in its LIS (or -1 if first).
  std::vector<int> prev(n), tail(n);
  for (int i = 0; i < n; i++) {
    // Find the tail to extend or improve. Comparing by value through the stored indices keeps
    // `tail` index-based for reconstruction; swap `<` for `<=` to allow a non-decreasing result.
    int pos = static_cast<int>(
        std::lower_bound(
            tail.begin(), tail.begin() + len, i, [&](int t, int x) { return *(lo + t) < *(lo + x); }
        ) -
        tail.begin()
    );
    if (pos == len) {
      len++;
    }
    prev[i] = pos > 0 ? tail[pos - 1] : -1;
    tail[pos] = i;
  }
  std::vector<int> res(len);
  for (int i = tail[len - 1]; i != -1; i = prev[i]) {
    res[--len] = i;
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <iostream>
using namespace std;

template<typename It>
void print_range(It lo, const vector<int> &indices) {
  for (int i : indices) {
    cout << *(lo + i) << " ";
  }
  cout << endl;
}

int main() {
  vector<int> a{-2, -5, 1, 9, 10, 8, 11, 10, 13, 11};
  vector<int> res = longest_increasing_subsequence(a.begin(), a.end());
  assert((res == vector<int>{1, 2, 3, 4, 6, 8}));
  print_range(a.begin(), res);
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
-5 1 9 10 11 13
\end{exampleoutput}
\subsection{Minimum Excluded Value}
Computes the minimum excluded nonnegative (MEX) integer of a set of integers, that is, the smallest integer $x \geq 0$ that does not appear in the set. It appears frequently in array problems, game theory with Grundy numbers, and dynamic programming states. Since the MEX of $n$ values never exceeds $n$, the one-shot version only marks values in $[0, {\inlinecodemath{n}}]$. The dynamic version stores counts for present values and an ordered set of currently missing candidates.

\begin{compactitem}
  \item \inlinecode{mex(lo, hi)} returns the MEX of the values in $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$.
  \item \inlinecode{DynamicMex()} maintains the MEX of a multiset of nonnegative integers.
  \item \inlinecode{add(x)} inserts one copy of value \inlinecode{x} if \inlinecode{x} $\geq 0$.
  \item \inlinecode{remove(x)} removes one copy of value \inlinecode{x} if present.
  \item \inlinecode{get()} returns the current MEX.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{mex(lo, hi)}, where $n$ is the number of values.
  \item $O(\log n)$ per call to \inlinecode{add(x)} and \inlinecode{remove(x)} for \inlinecode{DynamicMex}.
  \item $O(1)$ per call to \inlinecode{get()} for \inlinecode{DynamicMex}.
\end{compactitem}

\textbf{Space Complexity}: $O(n)$ auxiliary space for both approaches.

\begin{lstlisting}[language=C++]
#include <iterator>
#include <set>
#include <unordered_map>
#include <vector>

template<typename It>
int mex(It lo, It hi) {
  int n = std::distance(lo, hi);
  std::vector<char> seen(n + 1, false);
  for (It it = lo; it != hi; ++it) {
    if (0 <= *it && *it <= n) {
      seen[*it] = true;
    }
  }
  for (int x = 0; x <= n; x++) {
    if (!seen[x]) {
      return x;
    }
  }
  return n + 1;
}

class DynamicMex {
  std::unordered_map<int, int> count;
  std::set<int> missing;
  int size = 0;

  void ensure_candidate(int x) {
    if (!count.count(x)) {
      missing.insert(x);
    }
  }

 public:
  DynamicMex() { missing.insert(0); }

  void add(int x) {
    if (x < 0) {
      return;
    }
    ensure_candidate(size + 1);
    if (count[x]++ == 0) {
      missing.erase(x);
    }
    size++;
    ensure_candidate(size + 1);
  }

  void remove(int x) {
    auto it = count.find(x);
    if (it == count.end()) {
      return;
    }
    if (--it->second == 0) {
      count.erase(it);
      missing.insert(x);
    }
    size--;
  }

  int get() const { return *missing.begin(); }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> a{0, 1, 4, 2, 1};
  assert(mex(a.begin(), a.end()) == 3);
  vector<int> b{-1, 0, 1, 2};
  assert(mex(b.begin(), b.end()) == 3);  // Negative values are ignored.

  DynamicMex m;
  m.add(0);
  m.add(1);
  m.add(1);
  m.add(3);
  assert(m.get() == 2);
  m.add(2);
  assert(m.get() == 4);
  m.remove(1);
  assert(m.get() == 4);  // One copy of 1 remains.
  m.remove(1);
  assert(m.get() == 1);
  m.remove(100);
  assert(m.get() == 1);
  return 0;
}
\end{lstlisting}
\subsection{Iterated Function Cycle Detection}
Given a function $f$ mapping a finite set to itself and a starting value $x_0$, return the position and length of the cycle reached by repeatedly applying $f$. Formally, the sequence $x_0, x_1 = f(x_0), x_2 = f(x_1), \ldots, x_n = f(x_{n - 1}), \ldots$ must eventually repeat some value. If $x_i = x_j$ for $i < j$, then the sequence from $x_i$ to $x_{j - 1}$ repeats forever. This is useful for detecting cycles in functional graphs, pseudo-random generators, Pollard rho style algorithms, degenerate linked lists, and arrays where each index points to the next index. For example, if \inlinecode{nums} has size $n + 1$ and all values are in $[1, n]$, then $f(i)$ = \inlinecode{nums[i]} forms a functional graph; the duplicate value is the entry point of the cycle reached from $0$.

Floyd's cycle-finding algorithm, a.k.a. the ``tortoise and the hare algorithm'', is a space-efficient default choice: it moves two pointers through the sequence at different speeds, finds a meeting point inside the cycle, then locates the cycle start. Brent's algorithm is a compact variant that usually calls $f$ fewer times by only resetting the tortoise at powers of two; prefer it when evaluating $f$ is expensive and constant factors matter. In the detection phase, Brent advances one pointer with one call to $f$ per step, while Floyd advances the tortoise once and the hare twice, using three calls to $f$ per step.

For Floyd's algorithm, suppose the tail has length $m$ and the cycle has length $n$. At a meeting, the hare has traveled twice as far as the tortoise, so the tortoise's distance is a multiple of $n$. Consequently, the distance from the meeting point around the cycle to its entry is congruent to $m$ modulo $n$. Resetting one pointer to $x_0$ and advancing both one step at a time makes them meet at the cycle entry after $m$ steps. Brent first determines $n$, then advances one pointer $n$ steps; moving both together from that fixed separation likewise makes them meet at the entry.

\begin{compactitem}
  \item \inlinecode{find\_cycle\_floyd(f, x0)} returns the reached cycle as a pair $({\inlinecodemath{start}}, {\inlinecodemath{length}})$, where \inlinecode{start} is the smallest index $i$ such that $x_i$ is in the cycle, and \inlinecode{length} is the number of distinct values in the cycle.
  \item \inlinecode{find\_cycle\_brent(f, x0)} does the same with fewer calls to $f$ in many cases.
\end{compactitem}

\textbf{Time Complexity}: $O(m + n)$ per call, where $m$ is the first index in the cycle and $n$ is the cycle length.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
#include <utility>

template<typename Fn, typename T>
std::pair<int, int> find_cycle_floyd(Fn f, T x0) {
  // Find a meeting point inside the cycle.
  T tortoise = f(x0), hare = f(f(x0));
  while (tortoise != hare) {
    tortoise = f(tortoise);
    hare = f(f(hare));
  }
  // Find the cycle entry.
  int start = 0;
  tortoise = x0;
  while (tortoise != hare) {
    tortoise = f(tortoise);
    hare = f(hare);
    start++;
  }
  // Measure the cycle length.
  int length = 1;
  hare = f(tortoise);
  while (tortoise != hare) {
    hare = f(hare);
    length++;
  }
  return {start, length};
}

template<typename Fn, typename T>
std::pair<int, int> find_cycle_brent(Fn f, T x0) {
  // Find the cycle length by doubling the search block.
  int power = 1, length = 1;
  T tortoise = x0, hare = f(x0);
  while (tortoise != hare) {
    if (power == length) {
      tortoise = hare;
      power *= 2;
      length = 0;
    }
    hare = f(hare);
    length++;
  }
  // Separate the pointers by one cycle length.
  hare = x0;
  for (int i = 0; i < length; i++) {
    hare = f(hare);
  }
  // Find the cycle entry.
  int start = 0;
  tortoise = x0;
  while (tortoise != hare) {
    tortoise = f(tortoise);
    hare = f(hare);
    start++;
  }
  return {start, length};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <set>
#include <vector>
using namespace std;

int f(int x) {
  return (123 * x * x + 4567890) % 1337;
}

void verify(int x0, int start, int length) {
  set<int> s;
  int x = x0;
  for (int i = 0; i < start; i++) {
    assert(!s.count(x));
    s.insert(x);
    x = f(x);
  }
  int startx = x;
  s.clear();
  for (int i = 0; i < length; i++) {
    assert(!s.count(x));
    s.insert(x);
    x = f(x);
  }
  assert(startx == x);
}

// Given n + 1 integers within [1, n], where one value in [1, n] is present 2 or more times (but
// other values in [1, n] are present at most once), find the duplicate without modifying the array.
// Interpret each value as a pointer to the next index; the duplicate is the cycle entry.
int find_duplicate_integer(const vector<int> &nums) {
  auto [start, _] = find_cycle_floyd([&](int i) { return nums[i]; }, 0);
  int res = 0;
  for (int i = 0; i < start; i++) {
    res = nums[res];
  }
  return res;
}

int main() {
  int x0 = 0;
  auto [floyd_start, floyd_len] = find_cycle_floyd(f, x0);
  assert(floyd_start == 4 && floyd_len == 2);
  verify(x0, floyd_start, floyd_len);

  auto [brent_start, brent_len] = find_cycle_brent(f, x0);
  assert(brent_start == floyd_start && brent_len == floyd_len);

  assert(find_duplicate_integer({1, 3, 4, 2, 2}) == 2);
  assert(find_duplicate_integer({3, 1, 3, 4, 2}) == 3);
  return 0;
}
\end{lstlisting}
\subsection{0-1 Knapsack}
Solves the 0-1 knapsack problem: given items with nonnegative integer weights and values, choose each item at most once to maximize total value without exceeding a capacity limit.

The one-dimensional dynamic programming array \inlinecode{dp[w]} stores the best value that can be achieved with total weight at most \inlinecode{w} after processing some prefix of the items. Iterating weights in decreasing order is what enforces the 0-1 constraint; it prevents the same item from being reused during the current item update.

\begin{compactitem}
  \item \inlinecode{knapsack\_01(weight, value, capacity)} returns a pair $({\inlinecodemath{value}}, {\inlinecodemath{items}})$ containing the maximum value and the selected item indices in increasing order, where item \inlinecode{i} has weight \inlinecode{weight[i]} and value \inlinecode{value[i]}. All weights must be nonnegative integers. Items with weight $0$ are allowed and can each be taken once. The take/skip decision for every item and capacity is stored to reconstruct the optimal subset.
\end{compactitem}

\textbf{Time Complexity}: $O(n \cdot W)$ per call, where $n$ is the number of items and $W$ is \inlinecode{capacity}.

\textbf{Space Complexity}: $O(n \cdot W)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <cstdint>
#include <utility>
#include <vector>

std::pair<int64_t, std::vector<int>> knapsack_01(
    const std::vector<int> &weight, const std::vector<int64_t> &value, int capacity
) {
  int n = static_cast<int>(weight.size());
  assert(static_cast<int>(value.size()) == n && capacity >= 0);
  std::vector<int64_t> dp(capacity + 1, 0);
  std::vector<std::vector<char>> take(n, std::vector<char>(capacity + 1, false));
  for (int i = 0; i < n; i++) {
    assert(weight[i] >= 0);
    for (int w = capacity; w >= weight[i]; w--) {
      int64_t candidate = dp[w - weight[i]] + value[i];
      if (dp[w] < candidate) {
        dp[w] = candidate;
        take[i][w] = true;
      }
    }
  }
  std::vector<int> items;
  for (int i = n - 1, w = capacity; i >= 0; i--) {
    if (take[i][w]) {
      items.push_back(i);
      w -= weight[i];
    }
  }
  std::reverse(items.begin(), items.end());
  return {dp[capacity], items};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> weight{3, 4, 5, 9};
  vector<int64_t> value{4, 5, 7, 10};

  auto [best, items] = knapsack_01(weight, value, 8);
  assert(best == 11);
  assert((items == vector<int>{0, 2}));
  return 0;
}
\end{lstlisting}
\subsection{Unbounded Knapsack and Coin Change}
Solves two common unbounded knapsack variants, where each item may be used any number of times. In the maximum-value version, each item has a weight and value, and the goal is to maximize total value subject to a capacity limit. In the coin change version, each coin denomination may be used any number of times, and the goal is to count unordered ways to form a target sum.

For unbounded maximum-value knapsack, capacities are processed in increasing order. Thus, when computing \inlinecode{dp[w]}, every smaller-capacity state \inlinecode{dp[w - weight[i]]} is already final and may itself contain item \inlinecode{i}, allowing unlimited reuse. For unordered coin change, coins are the outer loop so each multiset of coins is counted once, independent of ordering.

\begin{compactitem}
  \item \inlinecode{unbounded\_knapsack(weight, value, capacity)} returns a pair $({\inlinecodemath{value}}, {\inlinecodemath{count}})$ containing the maximum value achievable with total weight at most \inlinecode{capacity} and \inlinecode{count[i]}, the number of copies selected of each item \inlinecode{i}.
  \item \inlinecode{count\_coin\_change(coins, target)} returns the number of unordered ways to make sum \inlinecode{target} using the given coin denominations.
\end{compactitem}

All capacities must be nonnegative integers. Weights and coin denominations should be positive to avoid infinitely many uses of a zero-weight item.

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \cdot W)$ for \inlinecode{unbounded\_knapsack()}, where $n$ is the number of items and $W$ is \inlinecode{capacity}.
  \item $O(c \cdot t)$ for \inlinecode{count\_coin\_change(coins, target)}, where $c$ is the number of coin denominations and $t$ is \inlinecode{target}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(W)$ auxiliary heap space for \inlinecode{unbounded\_knapsack()}.
  \item $O(t)$ auxiliary heap space for \inlinecode{count\_coin\_change(coins, target)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <cstdint>
#include <utility>
#include <vector>

std::pair<int64_t, std::vector<int>> unbounded_knapsack(
    const std::vector<int> &weight, const std::vector<int64_t> &value, int capacity
) {
  int n = static_cast<int>(weight.size());
  assert(static_cast<int>(value.size()) == n && capacity >= 0);
  for (int w : weight) {
    assert(w > 0);
  }
  std::vector<int64_t> dp(capacity + 1, 0);
  std::vector<int> choice(capacity + 1, -1);
  for (int w = 1; w <= capacity; w++) {
    for (int i = 0; i < n; i++) {
      if (weight[i] <= w && dp[w] < dp[w - weight[i]] + value[i]) {
        dp[w] = dp[w - weight[i]] + value[i];
        choice[w] = i;
      }
    }
  }
  std::vector<int> count(n, 0);
  for (int w = capacity; choice[w] != -1; w -= weight[choice[w]]) {
    count[choice[w]]++;
  }
  return {dp[capacity], count};
}

int64_t count_coin_change(const std::vector<int> &coins, int target) {
  std::vector<int64_t> dp(target + 1, 0);
  dp[0] = 1;
  for (int coin : coins) {
    for (int sum = coin; sum <= target; sum++) {
      dp[sum] += dp[sum - coin];
    }
  }
  return dp[target];
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> weight{3, 4, 5};
  vector<int64_t> value{4, 5, 7};
  auto [best, count] = unbounded_knapsack(weight, value, 10);
  assert(best == 14);
  assert((count == vector<int>{0, 0, 2}));

  vector<int> coins{1, 2, 5};
  assert(count_coin_change(coins, 5) == 4);  // 5, 2+2+1, 2+1+1+1, 1*5.
  return 0;
}
\end{lstlisting}
\subsection{Subset Sum}
Given a collection of integers and a target $v$, the subset sum problem asks for the maximum sum of any subset that is at most $v$. Two algorithms below solve this problem in complementary regimes, and which one wins depends on whether the bottleneck is the number of items or the size of the weights.

The meet-in-the-middle method splits the collection in half, enumerates every subset sum for each half, sorts one side, and binary searches for the best compatible partner. It makes no assumption about the sign or size of the values, so it handles negative inputs, and it is the method of choice when the number of items $n$ is small (up to roughly 40).

Pisinger's algorithm targets the opposite regime: nonnegative weights whose maximum value is small, with possibly many items. Let $W$ be the largest weight. It greedily takes a maximal prefix with sum $s$ that fits the target; because the next weight is at most $W$ and does not fit, $s$ is greater than \inlinecode{target} $- W$. The prefix itself is feasible, so the optimum lies between $s$ and \inlinecode{target} and only that width-$W$ window matters. A balanced sweep explores exchanges that add later items to sums below the target and remove earlier items from sums above it. For each sum, it retains the farthest prefix boundary reached, which dominates smaller boundaries because it permits at least the same future exchanges. This gives a running time linear in the number of items times $W$, independent of the target beyond that window. This routine is sometimes called ``fast knapsack,'' a slight misnomer: it maximizes a subset sum rather than packing items with distinct values, so it is really the bounded-weight case of subset sum rather than the 0-1 value knapsack.

\begin{compactitem}
  \item \inlinecode{max\_subset\_sum\_at\_most(lo, hi, v)} returns a pair $({\inlinecodemath{sum}}, {\inlinecodemath{items}})$ containing that maximum sum and the selected item indices relative to \inlinecode{lo}. Values may be negative. The range is supplied as random-access iterators, and 64-bit integers are used in calculations to avoid overflow.
  \item \inlinecode{max\_subset\_sum\_bounded(lo, hi, target)} returns the maximum sum of any subset of the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$ that does not exceed \inlinecode{target}. All weights and \inlinecode{target} must be nonnegative integers. This is faster than meet-in-the-middle when the largest weight is small relative to $2^{n/2}$.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(2^{n/2})$ per call to \inlinecode{max\_subset\_sum\_at\_most()}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(n \cdot W)$ per call to \inlinecode{max\_subset\_sum\_bounded()}, where $n$ is the number of items and $W$ is the largest weight.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(2^{n/2})$ auxiliary heap space for \inlinecode{max\_subset\_sum\_at\_most()}.
  \item $O(W)$ auxiliary heap space for \inlinecode{max\_subset\_sum\_bounded()}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstdint>
#include <utility>
#include <vector>

template<typename It>
std::pair<int64_t, std::vector<int>> max_subset_sum_at_most(It lo, It hi, int64_t v) {
  int n = static_cast<int>(hi - lo);
  int64_t llen = 1LL << (n / 2), hlen = 1LL << (n - n / 2);
  std::vector<int64_t> lsum(llen);
  std::vector<std::pair<int64_t, int64_t>> hsum(hlen);
  for (int64_t mask = 1; mask < llen; mask++) {
    int bit = __builtin_ctzll(mask);
    lsum[mask] = lsum[mask ^ (1LL << bit)] + *(lo + bit);
  }
  for (int64_t mask = 1; mask < hlen; mask++) {
    int bit = __builtin_ctzll(mask);
    hsum[mask].first = hsum[mask ^ (1LL << bit)].first + *(lo + n / 2 + bit);
    hsum[mask].second = mask;
  }
  std::sort(hsum.begin(), hsum.end());
  int64_t best = INT64_MIN, lmask = 0, hmask = 0;
  for (int64_t mask = 0; mask < llen; mask++) {
    auto it = std::upper_bound(hsum.begin(), hsum.end(), std::make_pair(v - lsum[mask], INT64_MAX));
    if (it != hsum.begin()) {
      --it;
      int64_t candidate = lsum[mask] + it->first;
      if (best < candidate) {
        best = candidate;
        lmask = mask;
        hmask = it->second;
      }
    }
  }
  std::vector<int> items;
  for (int i = 0; i < n / 2; i++) {
    if ((lmask >> i) & 1) {
      items.push_back(i);
    }
  }
  for (int i = 0; i < n - n / 2; i++) {
    if ((hmask >> i) & 1) {
      items.push_back(n / 2 + i);
    }
  }
  return {best, items};
}

template<typename It>
int64_t max_subset_sum_bounded(It lo, It hi, int target) {
  int n = static_cast<int>(hi - lo);
  // Greedily take a prefix while it still fits; the optimum then lies within one weight of target.
  int sum = 0, b = 0;
  while (b < n && sum + static_cast<int>(*(lo + b)) <= target) {
    sum += static_cast<int>(*(lo + b++));
  }
  if (b == n) {
    return sum;  // Every item fits, so the whole collection is the best subset.
  }
  int m = 0;
  for (It it = lo; it != hi; ++it) {
    m = std::max(m, static_cast<int>(*it));
  }
  // reach[s - target + m] = largest prefix boundary that can realize a balanced subset of sum s.
  // Only sums within m of target are tracked, which is where the optimum must lie.
  std::vector<int> reach(2 * m, -1);
  reach[sum - target + m] = b;
  for (int i = b; i < n; i++) {
    std::vector<int> prev = reach;
    int wi = static_cast<int>(*(lo + i));
    for (int x = 0; x < m; x++) {  // Add item i to sums currently below target.
      reach[x + wi] = std::max(reach[x + wi], prev[x]);
    }
    for (int x = 2 * m - 1; x > m; x--) {  // Drop an earlier item from sums above target.
      for (int j = std::max(0, prev[x]); j < reach[x]; j++) {
        int wj = static_cast<int>(*(lo + j));
        reach[x - wj] = std::max(reach[x - wj], j);
      }
    }
  }
  int s = target;
  while (reach[s - target + m] < 0) {
    s--;
  }
  return s;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  std::vector<int> a{9, 1, 5, 0, 1, 11, 5};
  auto [sum, items] = max_subset_sum_at_most(a.begin(), a.end(), 8);
  int64_t selected_sum = 0;
  for (int i : items) {
    selected_sum += a[i];
  }
  assert(sum == 7 && selected_sum == sum);
  std::vector<int> b{-7, -3, -2, 5, 8};
  assert(max_subset_sum_at_most(b.begin(), b.end(), 0).first == 0);

  // The bounded-weight method agrees with meet-in-the-middle on nonnegative inputs.
  assert(max_subset_sum_bounded(a.begin(), a.end(), 8) == 7);
  std::vector<int> c{3, 34, 4, 12, 5, 2};
  assert(max_subset_sum_bounded(c.begin(), c.end(), 9) == 9);     // 4 + 5.
  assert(max_subset_sum_bounded(c.begin(), c.end(), 10) == 10);   // 3 + 5 + 2.
  assert(max_subset_sum_bounded(c.begin(), c.end(), 100) == 60);  // All items fit.
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Greedy and Scheduling}{1.3 Greedy and Scheduling}}
\setcounter{section}{3}
\setcounter{subsection}{0}
\subsection{Interval Scheduling}
Selects the maximum number of non-overlapping intervals using the classic earliest-finish-time greedy algorithm. This is the unweighted interval scheduling problem, also known as activity selection: every interval has the same value, so the objective is to choose as many compatible intervals as possible.

The greedy choice is safe because among all intervals that could be chosen first, taking one with the earliest finish time leaves the most room for the remaining intervals. In any optimal solution, the first chosen interval can be exchanged for an earliest-finishing compatible interval without reducing the number of intervals selected. Repeating this argument after each choice proves the greedy algorithm optimal. The weighted version in 1.3.2 needs dynamic programming instead.

Intervals are represented as half-open ranges $[{\inlinecodemath{start}}, {\inlinecodemath{finish}})$, so two intervals are compatible if the next interval's \inlinecode{start} is at least the previous interval's \inlinecode{finish}.

\begin{compactitem}
  \item \inlinecode{schedule\_intervals(intervals)} returns one maximum-size compatible subset of intervals, in the order they are selected, from an input vector of \inlinecode{Interval} with fields \inlinecode{start}, \inlinecode{finish}, and the original \inlinecode{id}.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call due to sorting, where $n$ is the number of intervals.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space for sorting and the answer.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <climits>
#include <vector>

struct Interval {
  int start, finish, id;
};

std::vector<Interval> schedule_intervals(std::vector<Interval> intervals) {
  std::sort(intervals.begin(), intervals.end(), [](const Interval &a, const Interval &b) {
    return a.finish != b.finish ? a.finish < b.finish : a.start < b.start;
  });
  std::vector<Interval> res;
  int last_finish = INT_MIN;
  for (const auto &iv : intervals) {
    if (iv.start >= last_finish) {
      res.push_back(iv);
      last_finish = iv.finish;
    }
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<Interval> intervals{{1, 4, 0}, {3, 5, 1}, {0, 6, 2}, {5, 7, 3}, {8, 9, 4}};
  vector<Interval> chosen = schedule_intervals(intervals);
  // Earliest-finish greedy chooses ids 0, 3, 4.
  assert(chosen.size() == 3);
  assert(chosen[0].id == 0);
  assert(chosen[1].id == 3);
  assert(chosen[2].id == 4);

  vector<Interval> touching{{0, 2, 0}, {2, 4, 1}, {4, 5, 2}};
  assert(schedule_intervals(touching).size() == 3);
  return 0;
}
\end{lstlisting}
\subsection{Weighted Interval Scheduling}
Selects a maximum-weight subset of non-overlapping intervals using dynamic programming and binary search. Unlike unweighted interval scheduling, the earliest finish-time greedy choice is not sufficient when intervals have weights. With intervals sorted by finish time, the best total for the first $i$ intervals either skips interval $i$ or adds its weight to the best total over intervals finishing no later than its start, with that predecessor located by binary search.

Intervals are represented as half-open ranges $[{\inlinecodemath{start}}, {\inlinecodemath{finish}})$, so two intervals are compatible if the next interval's \inlinecode{start} is at least the previous interval's \inlinecode{finish}.

\begin{compactitem}
  \item \inlinecode{select\_weighted\_intervals(intervals)} returns a pair $({\inlinecodemath{weight}}, {\inlinecodemath{selected}})$ containing that maximum weight and the selected intervals as original input indices in execution order, from an input vector of \inlinecode{WeightedInterval} with fields \inlinecode{start}, \inlinecode{finish}, and \inlinecode{weight}. \inlinecode{dp[i]} stores the best answer using the first \inlinecode{i} intervals after sorting by finish time.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call due to sorting and binary searching compatible intervals.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstdint>
#include <numeric>
#include <utility>
#include <vector>

struct WeightedInterval {
  int start, finish;
  int64_t weight;
};

std::pair<int64_t, std::vector<int>> select_weighted_intervals(
    const std::vector<WeightedInterval> &intervals
) {
  int n = static_cast<int>(intervals.size());
  std::vector<int> order(n), finish(n), prev(n);
  std::iota(order.begin(), order.end(), 0);
  std::sort(order.begin(), order.end(), [&](int i, int j) {
    return intervals[i].finish != intervals[j].finish ? intervals[i].finish < intervals[j].finish
                                                      : intervals[i].start < intervals[j].start;
  });
  for (int i = 0; i < n; i++) {
    finish[i] = intervals[order[i]].finish;
  }
  std::vector<int64_t> dp(n + 1, 0);
  std::vector<char> take(n + 1, false);
  for (int i = 1; i <= n; i++) {
    int j =
        std::upper_bound(finish.begin(), finish.begin() + i - 1, intervals[order[i - 1]].start) -
        finish.begin();
    prev[i - 1] = j;
    int64_t candidate = dp[j] + intervals[order[i - 1]].weight;
    if (dp[i - 1] < candidate) {
      dp[i] = candidate;
      take[i] = true;
    } else {
      dp[i] = dp[i - 1];
    }
  }
  std::vector<int> selected;
  for (int i = n; i > 0;) {
    if (take[i]) {
      selected.push_back(order[i - 1]);
      i = prev[i - 1];
    } else {
      i--;
    }
  }
  std::reverse(selected.begin(), selected.end());
  return {dp[n], selected};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<WeightedInterval> intervals{{1, 3, 5}, {2, 5, 6}, {4, 6, 5}, {6, 7, 4}, {5, 8, 11}};
  auto [weight, selected] = select_weighted_intervals(intervals);
  // Taking ids 1 and 4 beats the earliest-finish unweighted-looking choices.
  assert(weight == 17);
  assert((selected == vector<int>{1, 4}));

  vector<WeightedInterval> touching{{0, 2, 5}, {2, 4, 6}, {1, 3, 7}};
  auto [touching_weight, touching_selected] = select_weighted_intervals(touching);
  assert(touching_weight == 11);
  assert((touching_selected == vector<int>{0, 1}));
  return 0;
}
\end{lstlisting}
\subsection{Interval Partitioning}
Assigns intervals to the minimum number of groups so that no two intervals in the same group overlap. This is the classic interval partitioning problem, also known as finding the minimum number of rooms needed for meetings.

The greedy algorithm sorts intervals by start time and reuses the group whose current finish time is earliest whenever possible. Otherwise, it creates a new group. This is optimal because the number of groups must be at least the maximum number of intervals overlapping at any time. When the earliest-finishing active group still overlaps the next interval, every active group overlaps it, so creating a new group is unavoidable. If that group is free, reusing it preserves as many later options as possible.

Intervals are treated as half-open ranges $[{\inlinecodemath{start}}, {\inlinecodemath{finish}})$, so one interval may reuse a room that another interval vacates at the same time.

\begin{compactitem}
  \item \inlinecode{partition\_intervals(intervals)} returns a vector \inlinecode{room}, where \inlinecode{room[id]} is the assigned room for each input interval given as a vector of \inlinecode{PartitionInterval} with fields \inlinecode{start}, \inlinecode{finish}, and \inlinecode{id}. The number of rooms used in the returned assignment is $1 + \max({\inlinecodemath{room[id]}})$, or $0$ if there are no intervals.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call due to sorting and priority queue operations.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <functional>
#include <queue>
#include <utility>
#include <vector>

struct PartitionInterval {
  int start, finish, id;
};

std::vector<int> partition_intervals(std::vector<PartitionInterval> intervals) {
  std::sort(
      intervals.begin(), intervals.end(),
      [](const PartitionInterval &a, const PartitionInterval &b) {
        return a.start != b.start ? a.start < b.start : a.finish < b.finish;
      }
  );
  std::vector<int> room(intervals.size());
  std::priority_queue<std::pair<int, int>, std::vector<std::pair<int, int>>, std::greater<>> pq;
  int rooms = 0;
  for (const auto &[start, finish, id] : intervals) {
    if (!pq.empty() && pq.top().first <= start) {
      int r = pq.top().second;
      pq.pop();
      room[id] = r;
    } else {
      room[id] = rooms++;
    }
    pq.emplace(finish, room[id]);
  }
  return room;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
using namespace std;

int main() {
  vector<PartitionInterval> intervals{{0, 30, 0}, {5, 10, 1}, {15, 20, 2}};
  vector<int> room = partition_intervals(intervals);
  // The long interval overlaps both short intervals, but the short intervals can share a room.
  assert(1 + *max_element(room.begin(), room.end()) == 2);
  assert(room[1] == room[2]);
  assert(room[0] != room[1]);

  vector<PartitionInterval> touching{{0, 2, 0}, {2, 4, 1}, {4, 5, 2}};
  vector<int> touching_room = partition_intervals(touching);
  assert(*max_element(touching_room.begin(), touching_room.end()) == 0);
  assert(partition_intervals({}).empty());
  return 0;
}
\end{lstlisting}
\subsection{Deadline Scheduling}
Schedules unit-time jobs with deadlines and profits to maximize total profit. Each job takes one time slot and must be scheduled at or before its deadline. The greedy algorithm processes jobs in decreasing profit order and places each chosen job in the latest available slot before its deadline.

The latest-slot choice is safe because it uses the current high-profit job while leaving earlier slots available for jobs with tighter deadlines. A disjoint-set forest over time slots makes this efficient: \inlinecode{find\_root(t)} returns the latest still-free slot at or before \inlinecode{t}, and occupying slot \inlinecode{t} links it to the next candidate slot \inlinecode{t - 1}.

\begin{compactitem}
  \item \inlinecode{select\_deadline\_jobs(jobs)} returns a pair $({\inlinecodemath{profit}}, {\inlinecodemath{slot}})$ containing that maximum profit and \inlinecode{slot[i]}, the assigned time slot of input job \inlinecode{i}, or $-1$ if that job is not selected, for a vector of \inlinecode{Job} with fields \inlinecode{deadline} and \inlinecode{profit}. Deadlines are positive integer time slots.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call due to sorting, with near-constant disjoint-set operations.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstdint>
#include <numeric>
#include <utility>
#include <vector>

struct Job {
  int deadline;
  int64_t profit;
};

class SlotDSU {
  std::vector<int> root;

 public:
  explicit SlotDSU(int n) : root(n + 1) { std::iota(root.begin(), root.end(), 0); }

  int find_root(int u) {
    if (root[u] != u) {
      root[u] = find_root(root[u]);
    }
    return root[u];
  }

  void occupy(int u) { root[u] = find_root(u - 1); }
};

std::pair<int64_t, std::vector<int>> select_deadline_jobs(const std::vector<Job> &jobs) {
  int n = static_cast<int>(jobs.size());
  std::vector<int> order(n);
  std::iota(order.begin(), order.end(), 0);
  std::sort(order.begin(), order.end(), [&](int i, int j) {
    return jobs[i].profit != jobs[j].profit ? jobs[i].profit > jobs[j].profit
                                            : jobs[i].deadline < jobs[j].deadline;
  });
  int max_deadline = 0;
  for (const auto &j : jobs) {
    max_deadline = std::max(max_deadline, j.deadline);
  }
  max_deadline = std::min(max_deadline, n);
  SlotDSU slots(max_deadline);
  std::vector<int> assigned(n, -1);
  int64_t res = 0;
  for (int i : order) {
    if (jobs[i].profit <= 0) {
      continue;
    }
    int slot = slots.find_root(std::min(jobs[i].deadline, max_deadline));
    if (slot > 0) {
      res += jobs[i].profit;
      assigned[i] = slot;
      slots.occupy(slot);
    }
  }
  return {res, assigned};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<Job> jobs{{2, 100}, {1, 19}, {2, 27}, {1, 25}, {3, 15}};
  auto [profit, slot] = select_deadline_jobs(jobs);
  // Jobs 0, 2, 4 fill slots 2, 1, 3 for total profit 100 + 27 + 15.
  assert(profit == 142);
  assert((slot == vector<int>{2, -1, 1, -1, 3}));
  return 0;
}
\end{lstlisting}
\subsection{Minimizing Late Jobs (Moore-Hodgson)}
Selects a maximum-size subset of jobs that can all be completed by their deadlines. Each job has a processing time and deadline, and all jobs are available at time $0$. The Moore-Hodgson algorithm sorts by deadline and keeps the accepted jobs in a max-heap by processing time; whenever the schedule becomes late, it removes the longest accepted job.

After considering jobs up to some deadline, if the accepted set no longer fits, at least one accepted job must be removed. Removing the longest one leaves the most remaining time while keeping the same number of removed jobs, so it is never worse than removing a shorter accepted job. Repeating this repair after each deadline leaves a largest feasible accepted set.

\begin{compactitem}
  \item \inlinecode{select\_on\_time\_jobs(jobs)} returns the selected jobs as original input indices in a feasible earliest-deadline-first execution order, for an input vector of \inlinecode{TimedJob} with fields \inlinecode{duration} and \inlinecode{deadline}.
\end{compactitem}

\textbf{Time Complexity}: $O(n \log n)$ per call due to sorting and heap operations.

\textbf{Space Complexity}: $O(n)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <numeric>
#include <queue>
#include <utility>
#include <vector>

struct TimedJob {
  int duration, deadline;
};

std::vector<int> select_on_time_jobs(const std::vector<TimedJob> &jobs) {
  int n = static_cast<int>(jobs.size());
  std::vector<int> order(n);
  std::iota(order.begin(), order.end(), 0);
  std::sort(order.begin(), order.end(), [&](int i, int j) {
    return jobs[i].deadline != jobs[j].deadline ? jobs[i].deadline < jobs[j].deadline
                                                : jobs[i].duration < jobs[j].duration;
  });
  std::priority_queue<std::pair<int, int>> accepted;
  std::vector<char> selected(n, false);
  int time = 0;
  for (int i : order) {
    time += jobs[i].duration;
    accepted.push({jobs[i].duration, i});
    selected[i] = true;
    if (time > jobs[i].deadline) {
      time -= accepted.top().first;
      selected[accepted.top().second] = false;
      accepted.pop();
    }
  }
  std::vector<int> res;
  for (int i : order) {
    if (selected[i]) {
      res.push_back(i);
    }
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<TimedJob> jobs{{3, 4}, {2, 3}, {1, 2}, {2, 7}};
  assert((select_on_time_jobs(jobs) == vector<int>{2, 1, 3}));
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Streaming and Randomized Algorithms}{1.4 Streaming and Randomized Algorithms}}
\setcounter{section}{4}
\setcounter{subsection}{0}
\subsection{Fisher-Yates Shuffle}
Randomly permutes a range in-place using the Fisher-Yates shuffle. Each possible permutation is produced with equal probability, assuming the random number source is uniform.

The algorithm scans from right to left. At each position $i$, it chooses a random position $j$ from $[0, i]$ and swaps the two elements. This fixes one uniformly random remaining element at a time.

\begin{compactitem}
  \item \inlinecode{fisher\_yates\_shuffle(lo, hi)} randomly shuffles the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ per call, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <random>

template<typename It>
void fisher_yates_shuffle(It lo, It hi) {
  static std::mt19937 rng(std::random_device{}());
  int n = static_cast<int>(hi - lo);
  for (int i = n - 1; i > 0; i--) {
    std::uniform_int_distribution<int> dist(0, i);
    int j = dist(rng);
    std::iter_swap(lo + i, lo + j);
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <vector>
using namespace std;

int main() {
  vector<int> a{1, 2, 3, 4, 5}, original(a);
  fisher_yates_shuffle(a.begin(), a.end());
  sort(a.begin(), a.end());
  assert(a == original);
  return 0;
}
\end{lstlisting}
\subsection{Reservoir Sampling}
Selects uniformly random samples from a stream whose length may be unknown in advance. This is useful when values arrive online or when storing the full dataset is unnecessary.

Reservoir sampling keeps only the chosen sample or samples in memory while processing each element once. The first arrivals fill the reservoir, and the $i$-th element then replaces a uniformly chosen occupant with probability $1/i$ (or $k/i$ when keeping $k$ samples), which by induction leaves every element seen so far equally likely to be in the sample.

Each class maintains its reservoir incrementally; call \inlinecode{add(x)} once per stream element in any order, then \inlinecode{get()} to retrieve the result.

\begin{compactitem}
  \item \inlinecode{ReservoirSampleOne\textless{}T\textgreater{}()} constructs a single-element sampler.
  \item \inlinecode{ReservoirSampleK\textless{}T\textgreater{}(k)} constructs a \inlinecode{k}-element sampler.
  \item \inlinecode{add(x)} incorporates one more stream element.
  \item \inlinecode{get()} returns the current sample. For \inlinecode{ReservoirSampleOne}, \inlinecode{get()} requires at least one \inlinecode{add()} call; for \inlinecode{ReservoirSampleK}, returns the full reservoir (may be fewer than \inlinecode{k} elements if the stream was shorter).
  \item \inlinecode{count()} returns the number of stream elements seen so far.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ per call to \inlinecode{add(x)}.
  \item $O(n)$ to process a range of $n$ elements.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for \inlinecode{ReservoirSampleOne}.
  \item $O(k)$ auxiliary heap space for \inlinecode{ReservoirSampleK}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cassert>
#include <random>
#include <vector>

int rand_int(int lo, int hi) {
  static std::mt19937 rng(std::random_device{}());
  return std::uniform_int_distribution<int>(lo, hi)(rng);
}

template<typename T>
class ReservoirSampleOne {
  T sample{};
  int seen;

 public:
  ReservoirSampleOne() : seen(0) {}

  void add(const T &x) {
    if (rand_int(0, seen++) == 0) {
      sample = x;
    }
  }

  const T &get() const {
    assert(seen > 0);
    return sample;
  }

  int count() const { return seen; }
};

template<typename T>
class ReservoirSampleK {
  int k, seen;
  std::vector<T> reservoir;

 public:
  explicit ReservoirSampleK(int k) : k(k), seen(0) {}

  void add(const T &x) {
    ++seen;
    if (static_cast<int>(reservoir.size()) < k) {
      reservoir.push_back(x);
    } else {
      int j = rand_int(0, seen - 1);
      if (j < k) {
        reservoir[j] = x;
      }
    }
  }

  const std::vector<T> &get() const { return reservoir; }
  int count() const { return seen; }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> a{10, 20, 30, 40, 50};

  // Streaming interface: feed elements one at a time.
  ReservoirSampleOne<int> s1;
  for (int x : a) {
    s1.add(x);
  }
  assert(s1.count() == 5);
  int one = s1.get();
  assert(one == 10 || one == 20 || one == 30 || one == 40 || one == 50);

  ReservoirSampleK<int> sk(3);
  for (int x : a) {
    sk.add(x);
  }
  assert(sk.count() == 5);
  assert(sk.get().size() == 3);

  ReservoirSampleK<int> large(10);
  for (int x : a) {
    large.add(x);
  }
  assert(large.count() == 5);
  assert(large.get().size() == 5);
  return 0;
}
\end{lstlisting}
\subsection{Majority Element (Boyer-Moore)}
Given a range of $n$ elements, find the majority element: an element that occurs strictly more than $\lfloor n/2 \rfloor$ times in the range.

The Boyer-Moore voting algorithm performs a first pass while maintaining a single candidate and a counter, incrementing it on a match and decrementing on a mismatch. Each decrement can be viewed as canceling one candidate occurrence against one different occurrence. Removing pairs of different values cannot eliminate a true majority, so if one exists, it survives all cancellations as the final candidate. The second pass counts the candidate's occurrences to confirm it truly holds a majority. The range is therefore traversed twice, so it only needs to be supplied as ForwardIterators rather than random-access.

If a majority element is guaranteed in advance to exist, the verification pass can be dropped and the candidate returned directly. The candidate phase alone is single-pass, so that variant could be written as a ``true'' streaming API with one element per call.

\begin{compactitem}
  \item \inlinecode{majority(lo, hi)} returns an iterator to the first occurrence of the majority element of the range $[{\inlinecodemath{lo}}, {\inlinecodemath{hi}})$, or \inlinecode{hi} if no majority element exists. Requires \inlinecode{operator==} on the value type.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ per call, where $n$ is the size of the array.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
template<typename It>
It majority(It lo, It hi) {
  int count = 0, n = 0;
  It candidate = lo;
  for (It it = lo; it != hi; ++it) {
    n++;  // Tally the size to avoid a O(n) std::distance() pass in case of ForwardIterators.
    if (count == 0) {
      candidate = it;
      count = 1;
    } else if (*it == *candidate) {
      count++;
    } else {
      count--;
    }
  }
  // Second pass: verify the candidate. Omit this if a majority is guaranteed to exist.
  count = 0;
  for (It it = lo; it != hi; ++it) {
    if (*it == *candidate) {
      count++;
    }
  }
  if (count <= n / 2) {
    return hi;
  }
  return candidate;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <vector>

int main() {
  std::vector<int> a{3, 2, 3, 1, 3};
  assert(*majority(a.begin(), a.end()) == 3);
  std::vector<int> b{2, 3, 3, 3, 2, 1};
  assert(majority(b.begin(), b.end()) == b.end());
  return 0;
}
\end{lstlisting}
\subsection{Online Statistics (Welford)}
Maintains the mean and variance of a stream of numbers in one pass using Welford's algorithm. The naive formula $\operatorname{Var}(X) = E[X^2] - E[X]^2$ can lose precision when the two terms are large and close together. Welford instead keeps the current mean and \inlinecode{m2}, the sum of squared deviations from that mean. Each new value shifts the running mean by its deviation divided by the new count, then adds the product of its deviations from the old and new means to \inlinecode{m2}. Population and sample variance follow by dividing \inlinecode{m2} by $n$ or $n - 1$.

\begin{compactitem}
  \item \inlinecode{OnlineStatistics()} constructs an empty summary.
  \item \inlinecode{add(x)} incorporates one more value \inlinecode{x} into the summary.
  \item \inlinecode{count()} returns the number of values seen so far.
  \item \inlinecode{mean()} returns the current arithmetic mean.
  \item \inlinecode{variance\_population()} returns population variance, dividing by $n$.
  \item \inlinecode{variance\_sample()} returns sample variance, dividing by $n - 1$.
\end{compactitem}

\textbf{Time Complexity}: $O(1)$ per call to \inlinecode{add(x)} and each query.

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
class OnlineStatistics {
  int n;
  double avg, m2;

 public:
  OnlineStatistics() : n(0), avg(0), m2(0) {}

  void add(double x) {
    n++;
    double delta = x - avg;
    avg += delta / n;
    double delta2 = x - avg;
    m2 += delta * delta2;
  }

  int count() const { return n; }
  double mean() const { return avg; }
  double variance_population() const { return n == 0 ? 0 : m2 / n; }
  double variance_sample() const { return n <= 1 ? 0 : m2 / (n - 1); }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cmath>

bool EQ(double a, double b) {
  return fabs(a - b) < 1e-9;
}

int main() {
  OnlineStatistics empty;
  assert(empty.count() == 0);
  assert(EQ(empty.mean(), 0.0));
  assert(EQ(empty.variance_population(), 0.0));
  assert(EQ(empty.variance_sample(), 0.0));

  OnlineStatistics singleton;
  singleton.add(7);
  assert(singleton.count() == 1);
  assert(EQ(singleton.mean(), 7.0));
  assert(EQ(singleton.variance_population(), 0.0));
  assert(EQ(singleton.variance_sample(), 0.0));

  OnlineStatistics stats;
  for (int x = 1; x <= 5; x++) {
    stats.add(x);
  }
  assert(stats.count() == 5);
  assert(EQ(stats.mean(), 3.0));
  assert(EQ(stats.variance_population(), 2.0));
  assert(EQ(stats.variance_sample(), 2.5));
  return 0;
}
\end{lstlisting}
\subsection{Frequent Elements (Misra-Gries)}
Finds candidate frequent elements in a stream using the Misra-Gries algorithm. With parameter $k$, the algorithm keeps at most $k - 1$ counters and guarantees that every value occurring more than $\lfloor n/k \rfloor$ times appears among the returned candidates. Each element increments its counter if already tracked, claims a counter if one of the $k - 1$ slots is free, and otherwise decrements every counter, discarding those that reach zero. Since a decrement step cancels $k$ distinct occurrences at once, it can happen at most $\lfloor n/k \rfloor$ times, so any value occurring more often must survive.

The candidates are not automatically verified, since the algorithm intentionally uses sublinear memory and does not retain the stream. If exact frequencies are needed, make a second pass over the input and count only the returned candidates.

\begin{compactitem}
  \item \inlinecode{misra\_gries(lo, hi, k)} returns a hash table of candidate values to their residual counters. Use \inlinecode{k = 2} for the Boyer-Moore majority-candidate special case.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ amortized per call on average: each element either hits an existing counter, taking $O(1)$, or triggers a full-table decrement, taking $O(k)$, but the table-full event can occur at most $n/(k-1)$ times total, so the amortized cost per element is $O(1 + k/(k-1))$ = $O(1)$. In the worst case, a single call is $O(n \cdot k)$, but that cannot be sustained across a long stream.

\textbf{Space Complexity}: $O(k)$ auxiliary heap space.

\begin{lstlisting}[language=C++]
#include <unordered_map>
#include <vector>

template<typename T, typename It>
std::unordered_map<T, int> misra_gries(It lo, It hi, int k) {
  std::unordered_map<T, int> count;
  for (It it = lo; it != hi; ++it) {
    if (auto found = count.find(*it); found != count.end()) {
      found->second++;
    } else if (static_cast<int>(count.size()) < k - 1) {
      count[*it] = 1;
    } else {
      std::vector<T> erase;
      for (auto &[key, val] : count) {
        if (--val == 0) {
          erase.push_back(key);
        }
      }
      for (const auto &key : erase) {
        count.erase(key);
      }
    }
  }
  return count;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> a{1, 2, 1, 3, 1, 2, 1, 4, 2, 2, 2};
  unordered_map<int, int> candidates = misra_gries<int>(a.begin(), a.end(), 3);
  assert(candidates.count(1));
  assert(candidates.count(2));
  assert(!candidates.count(3));
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Bit Manipulation}{1.5 Bit Manipulation}}
\setcounter{section}{5}
\setcounter{subsection}{0}
\subsection{Bit Operations and Masks}
Integers are all fixed-size sets of bits, where the $i$-th bit (counting from the least significant bit where $i = 0$) is ``present'' when set to $1$. Treating an \inlinecode{int} as a bitmask makes set membership, insertion, deletion, and iteration into single machine instructions, which is why bitmasks are the backbone of subset dynamic programming and many low-level tricks. The helpers below operate on \inlinecode{uint32\_t} masks; to use 64-bit masks, replace \inlinecode{uint32\_t} with \inlinecode{uint64\_t}, the literal \inlinecode{1u} with \inlinecode{1ull}, and the \inlinecode{1u \textless{}\textless{} 31} inside \inlinecode{clz()} with \inlinecode{1ull \textless{}\textless{} 63}. The loop implementations below exist for education only; production code should prefer the GCC built-ins named in the inline comments whenever available, or on C++20 systems the \inlinecode{\textless{}bit\textgreater{}} header, whose generic \inlinecode{constexpr} equivalents (also named in the comments) are additionally well-defined at $0$.

\begin{compactitem}
  \item \inlinecode{test\_bit(x, i)} returns whether the \inlinecode{i}-th bit of \inlinecode{x} is set.
  \item \inlinecode{set\_bit(x, i)}, \inlinecode{clear\_bit(x, i)}, and \inlinecode{toggle\_bit(x, i)} return \inlinecode{x} with the \inlinecode{i}-th respectively forced to $1$, forced to $0$, or flipped.
  \item \inlinecode{lowest\_set\_bit(x)} returns the value of the lowest set bit of \inlinecode{x} (a power of two), or $0$ if \inlinecode{x} is $0$. \inlinecode{clear\_lowest\_set\_bit(x)} returns \inlinecode{x} with its lowest set bit removed.
  \item \inlinecode{popcount(x)} returns the number of set bits, analogous to \inlinecode{\_\_builtin\_popcount()} and C++20's \inlinecode{std::popcount()}.
  \item \inlinecode{parity(x)} returns $1$ if the number of set bits is odd and $0$ otherwise, analogous to \inlinecode{\_\_builtin\_parity()}.
  \item \inlinecode{ctz(x)} returns the number of trailing 0-bits (undefined for $0$), analogous to \inlinecode{\_\_builtin\_ctz()} and C++20's \inlinecode{std::countr\_zero()}.
  \item \inlinecode{ffs(x)} returns one plus the position of the lowest 1-bit of \inlinecode{x}, or $0$ if \inlinecode{x} is $0$, analogous to \inlinecode{\_\_builtin\_ffs()}.
  \item \inlinecode{clz(x)} returns the number of leading 0-bits (undefined for $0$), analogous to \inlinecode{\_\_builtin\_clz()} and C++20's \inlinecode{std::countl\_zero()}.
  \item \inlinecode{is\_pow2(x)} returns whether \inlinecode{x} has exactly one set bit, analogous to C++20's \inlinecode{std::has\_single\_bit()}.
  \item \inlinecode{floor\_pow2(x)} returns the largest power of two that is $\leq$ \inlinecode{x} (for \inlinecode{x} $> 0$), analogous to C++20's \inlinecode{std::bit\_floor()}.
  \item \inlinecode{ceil\_pow2(x)} returns the smallest power of two that is $\geq$ \inlinecode{x} (for \inlinecode{x} $> 0$), analogous to C++20's \inlinecode{std::bit\_ceil()}.
  \item \inlinecode{for\_each\_set\_bit(x, f)} calls \inlinecode{f(i)} once for each set bit position \inlinecode{i} of \inlinecode{x}, in increasing order.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(b)$ worst case per call to \inlinecode{popcount(x)}, \inlinecode{parity(x)}, \inlinecode{ctz(x)}, \inlinecode{ffs(x)}, \inlinecode{clz(x)}, and \inlinecode{for\_each\_set\_bit(x, f)}, where $b$ is the bit width of the mask (32 here).
  \item $O(1)$ for all other operations.
\end{compactitem}

\textbf{Space Complexity}: $O(1)$ auxiliary for all operations.

\begin{lstlisting}[language=C++]
#include <cstdint>

bool test_bit(uint32_t x, int i) {  return (x >> i) & 1u; }
uint32_t set_bit(uint32_t x, int i) { return x | (1u << i); }
uint32_t clear_bit(uint32_t x, int i) { return x & ~(1u << i); }
uint32_t toggle_bit(uint32_t x, int i) { return x ^ (1u << i); }
uint32_t lowest_set_bit(uint32_t x) { return x & (0u - x); }
uint32_t clear_lowest_set_bit(uint32_t x) { return x & (x - 1); }

// popcount(), parity(), ctz(), ffs(), and clz() are spelled out here for educational purposes.
// Production code should use the compiler intrinsics named in the comments below.

int popcount(uint32_t x) {  // __builtin_popcount(x); std::popcount(x) in C++20.
  int count = 0;
  for (; x != 0; x = clear_lowest_set_bit(x)) {
    count++;
  }
  return count;
}

int parity(uint32_t x) {  // __builtin_parity(x)
  return popcount(x) & 1;
}

int ctz(uint32_t x) {  // __builtin_ctz(x); undefined when x == 0. std::countr_zero(x) in C++20.
  int count = 0;
  for (; (x & 1u) == 0; x >>= 1) {
    count++;
  }
  return count;
}

int ffs(uint32_t x) {  // __builtin_ffs(x); returns 0 when x == 0.
  return x == 0 ? 0 : ctz(x) + 1;
}

int clz(uint32_t x) {  // __builtin_clz(x); undefined when x == 0. std::countl_zero(x) in C++20.
  int count = 0;
  for (uint32_t mask = 1u << 31; (x & mask) == 0; mask >>= 1) {
    count++;
  }
  return count;
}

bool is_pow2(uint32_t x) {  // std::has_single_bit(x) in C++20.
  return x != 0 && (x & (x - 1)) == 0;
}

uint32_t floor_pow2(uint32_t x) {  // std::bit_floor(x) in C++20.
  return 1u << (31 - __builtin_clz(x));
}

uint32_t ceil_pow2(uint32_t x) {  // std::bit_ceil(x) in C++20.
  return is_pow2(x) ? x : 1u << (32 - __builtin_clz(x));
}

template<typename Fn>
void for_each_set_bit(uint32_t x, Fn f) {
  while (x != 0) {
    f(ctz(x));
    x = clear_lowest_set_bit(x);
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <vector>
using namespace std;

int main() {
  uint32_t x = 0b101100;
  assert(test_bit(x, 2) == true);
  assert(test_bit(x, 0) == false);
  assert(set_bit(x, 0) == 0b101101u);
  assert(clear_bit(x, 2) == 0b101000u);
  assert(toggle_bit(x, 3) == 0b100100u);

  assert(lowest_set_bit(x) == 0b100u);
  assert(clear_lowest_set_bit(x) == 0b101000u);
  assert(lowest_set_bit(0) == 0u);
  assert(clear_lowest_set_bit(0) == 0u);
  assert(popcount(x) == 3);
  assert(popcount(0) == 0);
  assert(parity(x) == 1);
  assert(parity(0b101101u) == 0);
  assert(parity(0) == 0);
  assert(ctz(x) == 2);
  assert(ffs(x) == 3);
  assert(ffs(0) == 0);
  assert(clz(x) == 26);

  assert(is_pow2(16) == true);
  assert(is_pow2(24) == false);
  assert(floor_pow2(20) == 16u);
  assert(ceil_pow2(20) == 32u);
  assert(ceil_pow2(16) == 16u);

  vector<int> bits;
  for_each_set_bit(x, [&](int b) { bits.push_back(b); });
  assert((bits == vector<int>{2, 3, 5}));
  return 0;
}
\end{lstlisting}
\subsection{Subset and Submask Iteration}
Many bitmask algorithms must visit related masks in a structured order: every submask of a fixed mask, every mask with a fixed number of set bits, or every mask while accumulating contributions from its submasks. Each pattern has a short, well-known idiom.

The classic submask loop \inlinecode{for (int s = m; s \textgreater{} 0; s = (s - 1) \& m)} walks every nonzero submask of \inlinecode{m} in decreasing order; subtracting 1 borrows through the zero bits of \inlinecode{m}, and the \inlinecode{\& m} masks them back off. Summed over all masks \inlinecode{m} of \inlinecode{n} bits, the total work is $\sum_k \binom{n}{k} 2^k = 3^n$, since each of the \inlinecode{n} bit positions is independently absent, present in \inlinecode{m} only, or present in both \inlinecode{m} and \inlinecode{s}.

Gosper's hack advances to the next mask with the same popcount. It isolates the lowest set bit and adds it to move the lowest movable 1 one position left, clearing the block of 1-bits beneath it. The remaining expression counts those cleared bits and packs them into the least-significant positions, producing the smallest larger integer with the same number of set bits.

\begin{compactitem}
  \item \inlinecode{submasks(m)} returns all submasks of \inlinecode{m}, including \inlinecode{m} itself and 0, in decreasing order.
  \item \inlinecode{next\_popcount(x)} returns the smallest integer greater than \inlinecode{x} with the same number of set bits (Gosper's hack), for \inlinecode{x \textgreater{} 0}.
  \item \inlinecode{masks\_with\_popcount(n, k)} returns all \inlinecode{k}-bit subsets of an \inlinecode{n}-bit universe as masks, in increasing order.
  \item \inlinecode{subset\_sum\_transform(f)} overwrites \inlinecode{f} (indexed by mask over \inlinecode{n} bits) so that \inlinecode{f[m]} becomes the sum of the original \inlinecode{f[s]} over all submasks \inlinecode{s} of \inlinecode{m}. This ``sum over subsets'' (SOS) DP is the bitmask analog of a prefix sum, accumulating one bit dimension at a time.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(2^{p})$ per call to \inlinecode{submasks(m)} where $p$ = \inlinecode{popcount(m)}.
  \item $O(1)$ per call to \inlinecode{next\_popcount(x)}.
  \item $O(\binom{n}{k})$ per call to \inlinecode{masks\_with\_popcount(n, k)}.
  \item $O(n \cdot 2^{n})$ per call to \inlinecode{subset\_sum\_transform(f)}, where \inlinecode{f} has $2^n$ entries.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for \inlinecode{next\_popcount(x)} and \inlinecode{subset\_sum\_transform(f)}.
  \item $O(s)$ auxiliary for \inlinecode{submasks(m)} and \inlinecode{masks\_with\_popcount(n, k)}, where $s$ is the result size.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cassert>
#include <cstdint>
#include <vector>

std::vector<uint32_t> submasks(uint32_t m) {
  std::vector<uint32_t> res;
  uint32_t s = m;
  while (true) {
    res.push_back(s);
    if (s == 0) {
      break;
    }
    s = (s - 1) & m;
  }
  return res;
}

uint32_t next_popcount(uint32_t x) {
  uint32_t c = x & (0u - x), r = x + c;
  return r | (((x ^ r) >> 2) / c);
}

std::vector<uint32_t> masks_with_popcount(int n, int k) {
  std::vector<uint32_t> res;
  if (k == 0) {
    return {0u};
  }
  uint32_t limit = 1u << n;
  for (uint32_t x = (1u << k) - 1; x < limit; x = next_popcount(x)) {
    res.push_back(x);
  }
  return res;
}

template<typename T>
void subset_sum_transform(std::vector<T> &f) {
  assert(!f.empty() && (f.size() & (f.size() - 1)) == 0);
  int n = __builtin_ctz(f.size());  // f.size() must be a power of two, 2^n.
  for (int b = 0; b < n; b++) {
    for (uint32_t m = 0; m < static_cast<uint32_t>(f.size()); m++) {
      if (m & (1u << b)) {
        f[m] += f[m ^ (1u << b)];
      }
    }
  }
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
using namespace std;

int main() {
  assert((submasks(0b101) == vector<uint32_t>{0b101, 0b100, 0b001, 0b000}));

  assert(next_popcount(0b0110) == 0b1001u);
  assert((
      masks_with_popcount(4, 2) == vector<uint32_t>{0b0011, 0b0101, 0b0110, 0b1001, 0b1010, 0b1100}
  ));

  // f[m] = 1 for every mask; after the transform f[m] counts submasks of m = 2^popcount(m).
  vector<int> f(1 << 3, 1);
  subset_sum_transform(f);
  assert(f[0b000] == 1);
  assert(f[0b101] == 4);
  assert(f[0b111] == 8);
  return 0;
}
\end{lstlisting}
\subsection{Gray Code}
The binary reflected Gray code is an ordering of all $2^n$ bitmasks in which consecutive masks differ in exactly one bit. It is useful whenever the cost of moving between adjacent states depends on a single bit flip: enumerating subsets while incrementally maintaining a value, hardware where only one bit may toggle at a time, and constructions like Hamiltonian paths on the hypercube.

The \inlinecode{k}-th Gray code is obtained from the ordinary binary value \inlinecode{k} by \inlinecode{k \textasciicircum{} (k \textgreater{}\textgreater{} 1)}. Incrementing \inlinecode{k} flips one binary bit and every lower bit, but XORing each bit with its higher neighbor cancels all but the highest of those changes, so consecutive Gray codes differ in exactly one bit. The map is a bijection: each binary bit is the prefix XOR of the Gray bits at and above it. The inverse loop computes those prefix XORs in doubling steps to recover the original rank.

\begin{compactitem}
  \item \inlinecode{gray\_code(k)} returns the \inlinecode{k}-th mask in Gray code order.
  \item \inlinecode{inverse\_gray\_code(g)} returns the rank \inlinecode{k} such that \inlinecode{gray\_code(k) == g}.
  \item \inlinecode{gray\_sequence(n)} returns all $2^n$ masks in Gray code order; consecutive entries (and the last with the first) differ in exactly one bit.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ per call to \inlinecode{gray\_code(k)}.
  \item $O(\log g)$ per call to \inlinecode{inverse\_gray\_code(g)}.
  \item $O(2^{n})$ per call to \inlinecode{gray\_sequence(n)}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for \inlinecode{gray\_code(k)} and \inlinecode{inverse\_gray\_code(g)}.
  \item $O(2^{n})$ auxiliary for \inlinecode{gray\_sequence(n)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <vector>

uint32_t gray_code(uint32_t k) {
  return k ^ (k >> 1);
}

uint32_t inverse_gray_code(uint32_t g) {
  for (uint32_t shift = 1; (g >> shift) != 0; shift <<= 1) {
    g ^= g >> shift;
  }
  return g;
}

std::vector<uint32_t> gray_sequence(int n) {
  std::vector<uint32_t> res(1u << n);
  for (uint32_t k = 0; k < static_cast<uint32_t>(res.size()); k++) {
    res[k] = gray_code(k);
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  assert(gray_code(0) == 0b000u);
  assert(gray_code(1) == 0b001u);
  assert(gray_code(2) == 0b011u);
  assert(gray_code(3) == 0b010u);

  for (uint32_t k = 0; k < 256; k++) {
    assert(inverse_gray_code(gray_code(k)) == k);
  }

  vector<uint32_t> seq = gray_sequence(3);
  assert(seq.size() == 8);
  for (int i = 0; i < static_cast<int>(seq.size()); i++) {
    uint32_t diff = seq[i] ^ seq[(i + 1) % seq.size()];
    assert(__builtin_popcount(diff) == 1);  // Cyclically, one bit changes per step.
  }
  return 0;
}
\end{lstlisting}
\subsection{XOR Basis}
An XOR basis (also called a linear basis over the field GF(2)) is the analog of a vector space basis for integers under the bitwise XOR operation. Each integer is treated as a vector of bits, where XOR plays the role of vector addition. Inserting a set of integers one at a time and keeping only those that are linearly independent yields a basis whose XOR-combinations (subset XORs) span exactly the same set of values as the original integers. From a basis of size $r$, exactly $2^r$ distinct values are reachable, and queries such as the maximum attainable XOR become a simple greedy scan.

The basis is kept in reduced form: the $i$-th basis is either $0$ or a value whose highest set bit is bit $i$, and no two basis elements share the same highest bit. Reducing a value means XORing it against the basis elements from the highest bit down, eliminating each set bit that a basis element covers.

\begin{compactitem}
  \item \inlinecode{insert(x)} adds \inlinecode{x} to the basis, returning \inlinecode{true} if \inlinecode{x} was linearly independent of the current basis (increasing its size), or \inlinecode{false} if \inlinecode{x} was already representable as a subset XOR.
  \item \inlinecode{contains(x)} returns whether \inlinecode{x} is representable as the XOR of some subset of the inserted values, i.e. whether it reduces to $0$ against the basis.
  \item \inlinecode{max\_xor(base)} returns the maximum value of \inlinecode{base} XORed with some subset XOR of the basis. With the default \inlinecode{base} of 0, this is the maximum subset XOR.
  \item \inlinecode{size()} returns the number of elements in the basis (its rank).
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(b)$ per call to \inlinecode{insert()}, \inlinecode{contains()}, and \inlinecode{max\_xor()}, where $b$ is the number of bits.
  \item $O(1)$ per call to \inlinecode{size()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(b)$ for storage of the basis.
  \item $O(1)$ auxiliary for all operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <array>
#include <climits>
#include <cstdint>

class XorBasis {
  static const int BITS = sizeof(uint64_t) * CHAR_BIT;

  std::array<uint64_t, BITS> basis{};  // basis[i] != 0 has its highest set bit at bit i.
  int sz = 0;

 public:
  bool insert(uint64_t x) {
    for (int i = BITS - 1; i >= 0; i--) {
      if (!((x >> i) & 1)) {
        continue;
      }
      if (basis[i] == 0) {
        basis[i] = x;
        sz++;
        return true;
      }
      x ^= basis[i];
    }
    return false;
  }

  bool contains(uint64_t x) const {
    for (int i = BITS - 1; i >= 0; i--) {
      if (((x >> i) & 1) && basis[i] != 0) {
        x ^= basis[i];
      }
    }
    return x == 0;
  }

  uint64_t max_xor(uint64_t base = 0) const {
    for (int i = BITS - 1; i >= 0; i--) {
      if ((base ^ basis[i]) > base) {
        base ^= basis[i];
      }
    }
    return base;
  }

  int size() const { return sz; }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  XorBasis b;
  assert(b.insert(1));   // 001
  assert(b.insert(2));   // 010
  assert(!b.insert(3));  // 011 = 1 ^ 2, already representable.
  assert(b.size() == 2);

  assert(b.contains(3));
  assert(b.contains(0));
  assert(!b.contains(4));

  assert(b.max_xor() == 3);   // Best subset XOR over {0, 1, 2, 3}.
  assert(b.max_xor(4) == 7);  // 4 ^ 3.
  return 0;
}
\end{lstlisting}
\subsection{Binary Trie}
Maintain a multiset of nonnegative $b$-bit integers as a binary trie, storing each value as a root-to-leaf path of its bits from the most significant (bit $b - 1$) down to the least significant. Every node keeps a count of the stored values passing through it, which supports deletion and counting queries, and lets the trie answer XOR-extremal queries by a greedy walk down the bits.

The classic application is the maximum-XOR query: to maximize \inlinecode{x \textasciicircum{} y} over all stored \inlinecode{y}, walk from the most significant bit. At each step, descend toward the child whose bit differs from that of \inlinecode{x} whenever such a branch exists, since setting a higher bit of the result always dominates any combination of lower bits. The minimum-XOR query is the mirror image, preferring the matching bit.

This is a multiset (a value may be inserted more than once), and is distinct from the XOR basis: the basis spans subset XORs of a fixed set, whereas this answers XOR queries against the stored elements themselves with support for insertion and deletion.

Values must lie in $[0, 2^b)$ for bit width $b$; the template parameters select the unsigned word type \inlinecode{U} and the bit width \inlinecode{BITS}, which must be less than the width of \inlinecode{U} (for example, $30$ with \inlinecode{uint32\_t} or $62$ with \inlinecode{uint64\_t}). Nodes are kept in a pool indexed by integer, where index $0$ is the root and also serves as the null child.

\begin{compactitem}
  \item \inlinecode{BinaryTrie\textless{}U, BITS\textgreater{}()} constructs an empty multiset.
  \item \inlinecode{size()} returns the number of stored values, and \inlinecode{empty()} returns whether there are none.
  \item \inlinecode{insert(x)} adds one occurrence of \inlinecode{x}.
  \item \inlinecode{erase(x)} removes one occurrence of \inlinecode{x}, returning whether one was present.
  \item \inlinecode{count(x)} returns the number of occurrences of \inlinecode{x}.
  \item \inlinecode{max\_xor(x)} and \inlinecode{min\_xor(x)} return the maximum and minimum of \inlinecode{x XOR y} over all stored \inlinecode{y}. The trie must be non-empty.
  \item \inlinecode{count\_xor\_less(x, k)} returns the number of stored \inlinecode{y} (with multiplicity) such that \inlinecode{x XOR y} is strictly less than \inlinecode{k}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(b)$ per call to \inlinecode{insert()}, \inlinecode{erase()}, \inlinecode{count()}, \inlinecode{max\_xor()}, \inlinecode{min\_xor()}, and \inlinecode{count\_xor\_less()}, where $b$ is the bit width \inlinecode{BITS}.
  \item $O(1)$ per call to \inlinecode{size()} and \inlinecode{empty()}.
\end{compactitem}

\textbf{Space Complexity}: $O(n \cdot b)$ for storage, where $n$ is the number of distinct values inserted so far.

\begin{lstlisting}[language=C++]
#include <array>
#include <cstdint>
#include <vector>

template<typename U = uint32_t, int BITS = 30>
class BinaryTrie {
  std::vector<std::array<int, 2>> child;  // child[node][bit], where 0 means no child.
  std::vector<int> cnt;                   // Number of stored values passing through each node.

  int new_node() {
    child.push_back({0, 0});
    cnt.push_back(0);
    return static_cast<int>(cnt.size()) - 1;
  }

 public:
  BinaryTrie() { new_node(); }  // Node 0 is the root (and doubles as the null child).

  int size() const { return cnt[0]; }
  bool empty() const { return cnt[0] == 0; }

  void insert(U x) {
    int node = 0;
    cnt[node]++;
    for (int b = BITS - 1; b >= 0; b--) {
      int bit = (x >> b) & 1;
      if (child[node][bit] == 0) {
        int next = new_node();
        child[node][bit] = next;
      }
      node = child[node][bit];
      cnt[node]++;
    }
  }

  int count(U x) const {
    int node = 0;
    for (int b = BITS - 1; b >= 0; b--) {
      int next = child[node][(x >> b) & 1];
      if (next == 0) {
        return 0;
      }
      node = next;
    }
    return cnt[node];
  }

  bool erase(U x) {
    if (count(x) == 0) {
      return false;
    }
    int node = 0;
    cnt[node]--;
    for (int b = BITS - 1; b >= 0; b--) {
      node = child[node][(x >> b) & 1];
      cnt[node]--;
    }
    return true;
  }

  U max_xor(U x) const {
    int node = 0;
    U res = 0;
    for (int b = BITS - 1; b >= 0; b--) {
      int bit = (x >> b) & 1;
      int opp = child[node][bit ^ 1];
      if (opp != 0 && cnt[opp] > 0) {
        res |= U(1) << b;
        node = opp;
      } else {
        node = child[node][bit];
      }
    }
    return res;
  }

  U min_xor(U x) const {
    int node = 0;
    U res = 0;
    for (int b = BITS - 1; b >= 0; b--) {
      int bit = (x >> b) & 1;
      int same = child[node][bit];
      if (same != 0 && cnt[same] > 0) {
        node = same;
      } else {
        res |= U(1) << b;
        node = child[node][bit ^ 1];
      }
    }
    return res;
  }

  int count_xor_less(U x, U k) const {
    if ((k >> BITS) != 0) {
      return cnt[0];  // Every XOR result is below 2^BITS, which is at most k.
    }
    int node = 0, res = 0;
    for (int b = BITS - 1; b >= 0; b--) {
      int xb = (x >> b) & 1, kb = (k >> b) & 1;
      if (kb == 1) {
        // Stored values matching x at bit b give XOR bit 0 here, hence are already below k.
        int below = child[node][xb];
        if (below != 0) {
          res += cnt[below];
        }
        node = child[node][xb ^ 1];  // Stay tied with k by taking XOR bit 1.
      } else {
        node = child[node][xb];  // XOR bit must be 0 to remain below k.
      }
      if (node == 0) {
        break;
      }
    }
    return res;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  BinaryTrie<> trie;
  assert(trie.empty());
  for (int x : {3, 10, 5, 25, 2}) {
    trie.insert(x);
  }
  assert(!trie.empty());
  assert(trie.size() == 5);

  // x XOR y over the stored y: {26, 19, 28, 0, 27} for x = 25.
  assert(trie.max_xor(25) == 28);                           // 25 XOR 5.
  assert(trie.min_xor(25) == 0);                            // 25 XOR 25.
  assert(trie.count_xor_less(25, 20) == 2);                 // XOR values 0 and 19 are below 20.
  assert(trie.count_xor_less(25, uint32_t(1) << 30) == 5);  // Bound at/above 2^BITS counts all.

  assert(trie.count(10) == 1);
  trie.insert(10);
  assert(trie.count(10) == 2);  // Multiset: two copies of 10.
  assert(trie.erase(10) && trie.count(10) == 1);
  assert(trie.erase(25) && trie.count(25) == 0);
  assert(!trie.erase(25));        // Already gone.
  assert(trie.max_xor(0) == 10);  // Largest remaining value is 10.
  return 0;
}
\end{lstlisting}
\subsection{Bitwise Convolution (FWHT)}
Bitwise convolution combines two arrays indexed by masks, grouping pairs of masks by a bitwise operation instead of by ordinary addition. For arrays \inlinecode{a} and \inlinecode{b} over an $n$-bit universe, the XOR convolution \inlinecode{c[m]} is the sum of \inlinecode{a[x]*b[y]} over all \inlinecode{x \textasciicircum{} y = m}. OR and AND convolution replace \inlinecode{x \textasciicircum{} y} with \inlinecode{x | y} or \inlinecode{x \& y}. These appear in subset DP, counting pairs by bitwise result, and algebra over the Boolean hypercube.

The fast Walsh-Hadamard transform (FWHT) diagonalizes XOR convolution in the same way that the FFT diagonalizes polynomial convolution: transform both arrays, multiply pointwise, and transform back. The OR and AND versions are Zeta/Mobius transforms over the subset lattice, but they have the same butterfly-shaped implementation and the same $O(n \log n)$ runtime for arrays of length $n = 2^k$.

\begin{compactitem}
  \item \inlinecode{xor\_transform(f, invert)} applies the XOR FWHT to \inlinecode{f} in place. The inverse transform uses \inlinecode{invert = true} and divides every coefficient by \inlinecode{f.size()}.
  \item \inlinecode{or\_transform(f, invert)} applies the OR Zeta transform in place; the inverse is Mobius inversion.
  \item \inlinecode{and\_transform(f, invert)} applies the AND Zeta transform in place; the inverse is Mobius inversion.
  \item \inlinecode{xor\_convolve(a, b)} returns the XOR convolution of \inlinecode{a} and \inlinecode{b}.
  \item \inlinecode{or\_convolve(a, b)} returns the OR convolution of \inlinecode{a} and \inlinecode{b}.
  \item \inlinecode{and\_convolve(a, b)} returns the AND convolution of \inlinecode{a} and \inlinecode{b}.
\end{compactitem}

Input arrays are padded with zeros to the next power of two. For exact integer XOR convolution, the inverse divisions are exact; for modular arithmetic, replace the division by multiplication with the modular inverse of the transform length.

\textbf{Time Complexity}: $O(n \log n)$ per transform or convolution, where $n$ is the padded power-of-two length.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(1)$ auxiliary for the in-place transforms.
  \item $O(n)$ auxiliary heap space for each convolution.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cassert>
#include <cstdint>
#include <vector>

int next_power_of_two(int n) {
  int res = 1;
  while (res < n) {
    res <<= 1;
  }
  return res;
}

template<typename T>
void xor_transform(std::vector<T> &f, bool invert) {
  int n = static_cast<int>(f.size());
  assert((n & (n - 1)) == 0);
  for (int len = 1; len < n; len <<= 1) {
    for (int i = 0; i < n; i += len << 1) {
      for (int j = 0; j < len; j++) {
        T u = f[i + j], v = f[i + j + len];
        f[i + j] = u + v;
        f[i + j + len] = u - v;
      }
    }
  }
  if (invert) {
    for (T &x : f) {
      x /= n;
    }
  }
}

template<typename T>
void or_transform(std::vector<T> &f, bool invert) {
  int n = static_cast<int>(f.size());
  assert((n & (n - 1)) == 0);
  for (int bit = 1; bit < n; bit <<= 1) {
    for (int mask = 0; mask < n; mask++) {
      if (mask & bit) {
        if (invert) {
          f[mask] -= f[mask ^ bit];
        } else {
          f[mask] += f[mask ^ bit];
        }
      }
    }
  }
}

template<typename T>
void and_transform(std::vector<T> &f, bool invert) {
  int n = static_cast<int>(f.size());
  assert((n & (n - 1)) == 0);
  for (int bit = 1; bit < n; bit <<= 1) {
    for (int mask = 0; mask < n; mask++) {
      if ((mask & bit) == 0) {
        if (invert) {
          f[mask] -= f[mask ^ bit];
        } else {
          f[mask] += f[mask ^ bit];
        }
      }
    }
  }
}

template<typename T, typename Transform>
std::vector<T> bitwise_convolve(std::vector<T> a, std::vector<T> b, Transform transform) {
  if (a.empty() || b.empty()) {
    return {};
  }
  int n = next_power_of_two(static_cast<int>(std::max(a.size(), b.size())));
  a.resize(n);
  b.resize(n);
  transform(a, false);
  transform(b, false);
  for (int i = 0; i < n; i++) {
    a[i] *= b[i];
  }
  transform(a, true);
  return a;
}

template<typename T>
std::vector<T> xor_convolve(std::vector<T> a, std::vector<T> b) {
  return bitwise_convolve(a, b, xor_transform<T>);
}

template<typename T>
std::vector<T> or_convolve(std::vector<T> a, std::vector<T> b) {
  return bitwise_convolve(a, b, or_transform<T>);
}

template<typename T>
std::vector<T> and_convolve(std::vector<T> a, std::vector<T> b) {
  return bitwise_convolve(a, b, and_transform<T>);
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int64_t> a{1, 2, 3, 4};
  vector<int64_t> b{5, 6, 7, 8};

  {
    vector<int64_t> f{1, 2, 3, 4}, original(f);
    xor_transform(f, false);
    xor_transform(f, true);
    assert(f == original);
  }
  {
    vector<int64_t> f{1, 2, 3, 4}, original(f);
    or_transform(f, false);
    or_transform(f, true);
    assert(f == original);
  }
  {
    vector<int64_t> f{1, 2, 3, 4}, original(f);
    and_transform(f, false);
    and_transform(f, true);
    assert(f == original);
  }

  // XOR: c[m] sums a[x] * b[y] over pairs with x ^ y == m.
  assert((xor_convolve(a, b) == vector<int64_t>{70, 68, 62, 60}));

  // OR: c[m] sums a[x] * b[y] over pairs with x | y == m.
  assert((or_convolve(a, b) == vector<int64_t>{5, 28, 43, 184}));

  // AND: c[m] sums a[x] * b[y] over pairs with x & y == m.
  assert((and_convolve(a, b) == vector<int64_t>{103, 52, 73, 32}));

  vector<int64_t> one_mask{0, 1, 0, 0};
  vector<int64_t> two_mask{0, 0, 1};
  assert((xor_convolve(one_mask, two_mask) == vector<int64_t>{0, 0, 0, 1}));
  assert((or_convolve(one_mask, two_mask) == vector<int64_t>{0, 0, 0, 1}));
  assert((and_convolve(one_mask, two_mask) == vector<int64_t>{1, 0, 0, 0}));
  return 0;
}
\end{lstlisting}
