\chapter{\texorpdfstring{Strings}{3. Strings}}

\section{\texorpdfstring{String Utilities}{3.1 String Utilities}}
\setcounter{section}{1}
Common string functions, many of which already have standard STL equivalents. These are written for educational purposes, without heavy optimizations (often depending on certain \inlinecode{std::string} functions that have unspecified complexity).

\textbf{Time Complexity}: $O(n)$ per call to most operations, where $n$ is the length of the input string or total length of all input strings. Exceptions are noted with the individual helpers below.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary heap space per call to operations that return a new string or vector of strings.
  \item $O(1)$ auxiliary space per call to in-place operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cctype>
#include <regex>
#include <sstream>
#include <stdexcept>
#include <string>
#include <vector>
using std::string;
\end{lstlisting}
Integer Conversion:

\begin{compactitem}
  \item \inlinecode{to\_str(i)} returns the string representation of integer \inlinecode{i}, much like \inlinecode{std::to\_string()}.
  \item \inlinecode{to\_int(s)} returns the integer representation of string \inlinecode{s}, much like \inlinecode{std::atoi()}, except here we handle special cases of overflow by throwing an exception.
  \item \inlinecode{itoa(value, \&str, base)} implements the non-standard C function which converts \inlinecode{value} into a C string, storing it into pointer \inlinecode{str} in the given \inlinecode{base}. For more generalized base conversion, see the math utilities section.
  \item \inlinecode{is\_integer(s)} returns whether \inlinecode{s} is a valid base-10 integer literal with an optional sign.
  \item \inlinecode{is\_number(s)} returns whether \inlinecode{s} is a decimal number literal with optional sign, decimal point, and exponent.
\end{compactitem}

\begin{lstlisting}[language=C++]
template<class Int>
string to_str(Int i) {
  std::ostringstream oss;
  oss << i;
  return oss.str();
}

int to_int(const string &s) {
  std::istringstream iss(s);
  int res;
  if (!(iss >> res)) {
    throw std::runtime_error("to_int failed");
  }
  return res;
}

char *itoa(int value, char *str, int base = 10) {
  if (base < 2 || base > 36) {
    *str = '\0';
    return str;
  }
  char *ptr = str, *ptr1 = str, tmp_c;
  int tmp_v;
  do {
    tmp_v = value;
    value /= base;
    *ptr++ =
        "zyxwvutsrqponmlkjihgfedcba9876543210123456789"
        "abcdefghijklmnopqrstuvwxyz"[35 + (tmp_v - value * base)];
  } while (value);
  if (tmp_v < 0) {
    *ptr++ = '-';
  }
  for (*ptr-- = '\0'; ptr1 < ptr; *ptr1++ = tmp_c) {
    tmp_c = *ptr;
    *ptr-- = *ptr1;
  }
  return str;
}

bool is_integer(const string &s) {
  int i = (s.empty() || (s[0] != '-' && s[0] != '+')) ? 0 : 1;
  if (i == static_cast<int>(s.size())) {
    return false;
  }
  for (; i < static_cast<int>(s.size()); i++) {
    if (!isdigit(static_cast<unsigned char>(s[i]))) {
      return false;
    }
  }
  return true;
}

bool is_number(const string &s) {
  int i = (s.empty() || (s[0] != '-' && s[0] != '+')) ? 0 : 1;
  bool seen_digit = false, seen_dot = false;
  for (; i < static_cast<int>(s.size()); i++) {
    unsigned char c = s[i];
    if (isdigit(c)) {
      seen_digit = true;
    } else if (c == '.' && !seen_dot) {
      seen_dot = true;
    } else {
      break;
    }
  }
  if (!seen_digit) {
    return false;
  }
  if (i < static_cast<int>(s.size()) && (s[i] == 'e' || s[i] == 'E')) {
    i++;
    if (i < static_cast<int>(s.size()) && (s[i] == '-' || s[i] == '+')) {
      i++;
    }
    bool seen_exp_digit = false;
    for (; i < static_cast<int>(s.size()); i++) {
      if (!isdigit(static_cast<unsigned char>(s[i]))) {
        return false;
      }
      seen_exp_digit = true;
    }
    return seen_exp_digit;
  }
  return i == static_cast<int>(s.size());
}
\end{lstlisting}
Case Conversion:

\begin{compactitem}
  \item \inlinecode{to\_upper(s)} returns \inlinecode{s} with all alphabetical characters converted to uppercase.
  \item \inlinecode{to\_lower(s)} returns \inlinecode{s} with all alphabetical characters converted to lowercase.
  \item \inlinecode{to\_title(s)} returns the title case representation of string \inlinecode{s}, where the first letter of every word (consecutive alphabetical characters) is capitalized.
\end{compactitem}

\begin{lstlisting}[language=C++]
string to_upper(const string &s) {
  string res;
  auto upper = [](unsigned char c) { return static_cast<char>(toupper(c)); };
  for (char c : s) {
    res.push_back(upper(c));
  }
  return res;
}

string to_lower(const string &s) {
  string res;
  auto lower = [](unsigned char c) { return static_cast<char>(tolower(c)); };
  for (char c : s) {
    res.push_back(lower(c));
  }
  return res;
}

string to_title(const string &s) {
  string res;
  auto lower = [](unsigned char c) { return static_cast<char>(tolower(c)); };
  auto upper = [](unsigned char c) { return static_cast<char>(toupper(c)); };
  char prev = '\0';
  for (char c : s) {
    res.push_back(isalpha(static_cast<unsigned char>(prev)) ? lower(c) : upper(c));
    prev = res.back();
  }
  return res;
}
\end{lstlisting}
Stripping:

\begin{compactitem}
  \item \inlinecode{lstrip(s)} strips the left side of \inlinecode{s} in-place (that is, the input is modified) using the given delimiters and returns a reference to the stripped string.
  \item \inlinecode{rstrip(s)} strips the right side of \inlinecode{s} in-place using the given delimiters and returns a reference to the stripped string.
  \item \inlinecode{strip(s)} strips both sides of \inlinecode{s} in-place and returns a reference to the stripped string.
  \item \inlinecode{ltrimmed(s)}, \inlinecode{rtrimmed(s)}, and \inlinecode{trimmed(s)} do not modify \inlinecode{s} and returns stripped copies.
\end{compactitem}

\begin{lstlisting}[language=C++]
string &lstrip(string &s, const string &delim = " \n\t\v\f\r") {
  size_t pos = s.find_first_not_of(delim);
  if (pos != string::npos) {
    s.erase(0, pos);
  }
  return s;
}

string &rstrip(string &s, const string &delim = " \n\t\v\f\r") {
  size_t pos = s.find_last_not_of(delim);
  if (pos != string::npos) {
    s.erase(pos + 1);
  }
  return s;
}

string &strip(string &s, const string &delim = " \n\t\v\f\r") {
  return lstrip(rstrip(s));
}

string ltrimmed(string s, const string &delim = " \n\t\v\f\r") {
  return lstrip(s, delim);
}

string rtrimmed(string s, const string &delim = " \n\t\v\f\r") {
  return rstrip(s, delim);
}

string trimmed(string s, const string &delim = " \n\t\v\f\r") {
  return strip(s, delim);
}
\end{lstlisting}
Find and Replace:

\begin{compactitem}
  \item \inlinecode{starts\_with(s, prefix)} returns whether \inlinecode{s} begins with \inlinecode{prefix}.
  \item \inlinecode{ends\_with(s, suffix)} returns whether \inlinecode{s} ends with \inlinecode{suffix}.
  \item \inlinecode{contains(s, needle)} returns whether \inlinecode{needle} occurs in \inlinecode{s}.
  \item \inlinecode{find\_all(haystack, needle)} returns a vector of all positions where the string \inlinecode{needle} appears in the string \inlinecode{haystack}.
  \item \inlinecode{replace(s, old, replacement)} returns a copy of \inlinecode{s} with all occurrences of the string \inlinecode{old} replaced with the given \inlinecode{replacement}.
  \item \inlinecode{regex\_find\_all(s, pattern)} returns every substring of \inlinecode{s} matched by \inlinecode{pattern}.
  \item \inlinecode{regex\_groups(s, pattern)} returns the capture groups of the first match of \inlinecode{pattern} in \inlinecode{s}, excluding the full match. If there is no match, returns an empty vector.
  \item \inlinecode{regex\_replace\_all(s, pattern, replacement)} returns a copy of \inlinecode{s} with every regex match replaced by \inlinecode{replacement}.
\end{compactitem}

\begin{lstlisting}[language=C++]
bool starts_with(const string &s, const string &prefix) {
  return s.size() >= prefix.size() && s.compare(0, prefix.size(), prefix) == 0;
}

bool ends_with(const string &s, const string &suffix) {
  return s.size() >= suffix.size() &&
         s.compare(s.size() - suffix.size(), suffix.size(), suffix) == 0;
}

bool contains(const string &s, const string &needle) {
  return s.find(needle) != string::npos;
}

std::vector<int> find_all(const string &haystack, const string &needle) {
  std::vector<int> res;
  size_t pos = haystack.find(needle, 0);
  while (pos != string::npos) {
    res.push_back(pos);
    pos = haystack.find(needle, pos + 1);
  }
  return res;
}

string replace(const string &s, const string &old, const string &replacement) {
  if (old.empty()) {
    return s;
  }
  string res(s);
  size_t pos = 0;
  while ((pos = res.find(old, pos)) != string::npos) {
    res.replace(pos, old.length(), replacement);
    pos += replacement.length();
  }
  return res;
}

std::vector<string> regex_find_all(const string &s, const string &pattern) {
  std::regex re(pattern);
  return {std::sregex_token_iterator(s.begin(), s.end(), re, 0), std::sregex_token_iterator()};
}

std::vector<string> regex_groups(const string &s, const string &pattern) {
  std::smatch match;
  if (!std::regex_search(s, match, std::regex(pattern))) {
    return {};
  }
  std::vector<string> res;
  for (int i = 1; i < static_cast<int>(match.size()); i++) {
    res.push_back(match[i].str());
  }
  return res;
}

string regex_replace_all(const string &s, const string &pattern, const string &replacement) {
  return std::regex_replace(s, std::regex(pattern), replacement);
}
\end{lstlisting}
Joining and Splitting:

\begin{compactitem}
  \item \inlinecode{join(v, delim)} returns the strings in vector \inlinecode{v} concatenated, separated by the given delimiter.
  \item \inlinecode{pad\_left(s, width, ch)} returns \inlinecode{s} left-padded with \inlinecode{ch} until its length is at least \inlinecode{width}.
  \item \inlinecode{pad\_right(s, width, ch)} returns \inlinecode{s} right-padded with \inlinecode{ch} until its length is at least \inlinecode{width}.
  \item \inlinecode{split(s, char delim)} returns a vector of tokens of \inlinecode{s}, split on a single character delimiter. Empty tokens are skipped, so \inlinecode{split("a::b", \textquotesingle{}:\textquotesingle{})} returns \inlinecode{\{"a", "b"\}}, not \inlinecode{\{"a", "", "b"\}}.
  \item \inlinecode{split(s, string delim)} returns a vector of tokens of \inlinecode{s}, split on a set of many possible single character delimiters. All characters of \inlinecode{delim} will be removed from \inlinecode{s}, and the remaining token(s) of \inlinecode{s} will be added sequentially to a vector and returned. As with the first version, empty tokens are skipped. For example, \inlinecode{split("a::b", ":")} returns \inlinecode{\{"a", "b"\}}, not \inlinecode{\{"a", "", "b"\}}.
  \item \inlinecode{explode(s, delim)} returns a vector of tokens of \inlinecode{s}, split on the entire delimiter string \inlinecode{delim}. Unlike the \inlinecode{split()} functions above, \inlinecode{delim} is treated as a contiguous boundary string, not merely a set of possible boundary characters. This will not skip empty tokens. For example, \inlinecode{explode("a::::b", "::")} yields \inlinecode{\{"a", "", "b"\}}, not \inlinecode{\{"a", "b"\}}.
  \item \inlinecode{explode(s, char delim)} is the single-character delimiter version that also preserves empty tokens.
\end{compactitem}

\begin{lstlisting}[language=C++]
string join(const std::vector<string> &v, const string &delim = " ") {
  string res;
  for (int i = 0; i < static_cast<int>(v.size()); i++) {
    if (i > 0) res += delim;
    res += v[i];
  }
  return res;
}

string pad_left(const string &s, int width, char ch = ' ') {
  int pad = width - static_cast<int>(s.size());
  return pad > 0 ? string(pad, ch) + s : s;
}

string pad_right(const string &s, int width, char ch = ' ') {
  int pad = width - static_cast<int>(s.size());
  return pad > 0 ? s + string(pad, ch) : s;
}

std::vector<string> split(const string &s, char delim) {
  std::vector<string> res;
  std::stringstream ss(s);
  string curr;
  while (std::getline(ss, curr, delim)) {
    if (!curr.empty()) {
      res.push_back(curr);
    }
  }
  return res;
}

std::vector<string> split(const string &s, const string &delim = " \n\t\v\f\r") {
  std::vector<string> res;
  string curr;
  for (char c : s) {
    if (delim.find(c) == string::npos) {
      curr += c;
    } else if (!curr.empty()) {
      res.push_back(curr);
      curr = "";
    }
  }
  if (!curr.empty()) {
    res.push_back(curr);
  }
  return res;
}

std::vector<string> explode(const string &s, const string &delim) {
  std::vector<string> res;
  std::size_t last = 0, next = 0;
  while ((next = s.find(delim, last)) != string::npos) {
    res.push_back(s.substr(last, next - last));
    last = next + static_cast<int>(delim.size());
  }
  res.push_back(s.substr(last));
  return res;
}

std::vector<string> explode(const string &s, char delim) {
  std::vector<string> res;
  std::size_t last = 0, next = 0;
  while ((next = s.find(delim, last)) != string::npos) {
    res.push_back(s.substr(last, next - last));
    last = next + 1;
  }
  res.push_back(s.substr(last));
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  assert(to_str(123) + "4" == "1234");
  assert(to_int("1234") == 1234);
  vector<char> buffer(50);
  assert(string(itoa(1750, buffer.data(), 10)) == "1750");
  assert(string(itoa(1750, buffer.data(), 16)) == "6d6");
  assert(string(itoa(1750, buffer.data(), 2)) == "11011010110");
  assert(is_integer("-123") && !is_integer("12.3"));
  assert(is_number("-.5e+2") && is_number("123.") && !is_number("1e"));

  assert(to_upper("Hello world") == "HELLO WORLD");
  assert(to_lower("Hello World") == "hello world");
  assert(to_title("hello world") == "Hello World");

  string s("   abc \n");
  string t = s;
  assert(lstrip(s) == "abc \n");
  assert(rstrip(s) == strip(t));
  assert(trimmed(" \t abc \n") == "abc");

  vector<int> pos;
  pos.push_back(0);
  pos.push_back(7);
  assert(starts_with("abracadabra", "abra"));
  assert(ends_with("abracadabra", "dabra"));
  assert(contains("abracadabra", "cad"));
  assert(find_all("abracadabra", "ab") == pos);
  assert(replace("abcdabba", "ab", "00") == "00cd00ba");
  assert(join(regex_find_all("a12b345", "[0-9]+"), "|") == "12|345");
  assert(join(regex_groups("abc-123", "([a-z]+)-([0-9]+)"), "|") == "abc|123");
  assert(regex_replace_all("a12b345", "[0-9]+", "#") == "a#b#");

  assert(pad_left("42", 5, '0') == "00042");
  assert(pad_right("hi", 4, '.') == "hi..");
  assert(join(split("a\nb\ncde\nf", '\n'), "|") == "a|b|cde|f");  // split v1
  assert(join(split("a::b", ':'), "|") == "a|b");                 // split v1 skips empty tokens
  assert(join(split("a::b,cde:,f", ":,"), "|") == "a|b|cde|f");   // split v2
  assert(join(explode("a..b.cde....f", ".."), "|") == "a|b.cde||f");
  assert(join(explode("a::b:", ':'), "|") == "a||b|");
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Hashing}{3.2 Hashing}}
\setcounter{section}{2}
\setcounter{subsection}{0}
\subsection{Hash Functions}
Provides a small library of non-cryptographic hash functions and hash functors for common contest keys. These helpers are useful for fingerprinting values, combining keys, seeding randomized algorithms, and defining \inlinecode{std::unordered\_map} keys for pairs, tuples, and vectors. They are not suitable for passwords, signatures, or adversarial security.

The standalone functions below are deterministic. The \inlinecode{IntHasher} functor adds a per-run random seed before mixing, which is useful for making integer-keyed hash tables harder to hack in open-test contests.

The composite hashers separate the mixing primitive (\inlinecode{mix64}/\inlinecode{hash\_combine64}) from each type's traversal of its parts, so swapping in a different mixing function only requires editing one place.

FNV-1a (Fowler-Noll-Vo) is a tiny hash for variable-length byte sequences. Starting from a fixed offset basis, it XORs each byte into the accumulator and then multiplies by a fixed prime; the \inlinecode{1a} variant XORs before multiplying, which mixes slightly better than the original FNV-1. Reach for it when the key is a string or byte buffer, whereas \inlinecode{mix32} and \inlinecode{mix64} are finalizers for a single fixed-width integer. Its appeal is simplicity: a few lines, no lookup tables, and no seed, with distribution good enough for the short keys common in contests. For very long inputs a block hash such as MurmurHash or xxHash is faster and mixes more thoroughly, and because FNV-1a is unseeded it offers no protection against adversarial inputs.

\begin{compactitem}
  \item \inlinecode{mix32(x)} mixes a 32-bit unsigned integer \inlinecode{x}, using the MurmurHash3 finalizer.
  \item \inlinecode{mix64(x)} mixes a 64-bit unsigned integer \inlinecode{x}, using the SplitMix64 finalizer.
  \item \inlinecode{hash32(x)} and \inlinecode{hash64(x)} hash any integer type that can be converted to \inlinecode{uint32\_t} or \inlinecode{uint64\_t}, respectively.
  \item \inlinecode{hash\_combine32(h, x)} and \inlinecode{hash\_combine64(h, x)} fold another already-hashed value into an existing hash accumulator.
  \item \inlinecode{hash\_float(x)} and \inlinecode{hash\_double(x)} hash floating-point bit patterns, normalizing \inlinecode{-0.0} to \inlinecode{0.0}. Different NaN representations may still hash differently.
  \item \inlinecode{hash\_string\_fnv1a32(s)} and \inlinecode{hash\_string\_fnv1a64(s)} compute FNV-1a hashes of string \inlinecode{s}.
  \item \inlinecode{hash\_range32(first, last)} and \inlinecode{hash\_range64(first, last)} hash a sequence of integer-like values.
  \item \inlinecode{PairIntHasher} is a self-contained hasher for \inlinecode{std::pair\textless{}int, int\textgreater{}} keys, written without any dependency on the templates below so that it can be copy-pasted on its own.
  \item \inlinecode{IntHasher\textless{}Int\textgreater{}}, \inlinecode{PairHasher\textless{}A, B\textgreater{}}, \inlinecode{TupleHasher\textless{}Ts...\textgreater{}}, and \inlinecode{VectorHasher\textless{}T\textgreater{}} are hash functors passed as the third template argument of \inlinecode{std::unordered\_map} or \inlinecode{std::unordered\_set}.
  \item \inlinecode{GenericHasher\textless{}T\textgreater{}} recursively handles integer, pair, tuple, and vector keys, and falls back to \inlinecode{std::hash\textless{}T\textgreater{}} for other hashable types.
  \item Specializations of \inlinecode{std::hash} for \inlinecode{std::pair} and \inlinecode{std::tuple} are also provided, letting those keys be used in \inlinecode{std::unordered\_map}/\inlinecode{std::unordered\_set} without an explicit hasher argument. Specializing \inlinecode{std::hash} for purely standard types is technically nonstandard, so the functor forms above are the portable alternative. (Vector keys are left to the functors to avoid clashing with the standard \inlinecode{std::hash\textless{}std::vector\textless{}bool\textgreater{}\textgreater{}}.)
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ per call to the integer, combiner, and floating-point hash functions.
  \item $O(n)$ per call to string and range hashing functions, where $n$ is the number of elements processed.
  \item $O(n)$ per call to \inlinecode{VectorHasher\textless{}T\textgreater{}::operator()}, where $n$ is the vector size.
\end{compactitem}

\textbf{Space Complexity}: $O(1)$ auxiliary.

\begin{lstlisting}[language=C++]
#include <chrono>
#include <cstddef>
#include <cstdint>
#include <cstring>
#include <functional>
#include <string>
#include <tuple>
#include <type_traits>
#include <unordered_map>
#include <utility>
#include <vector>

uint32_t mix32(uint32_t x) {
  x ^= x >> 16;
  x *= 0x85ebca6bU;
  x ^= x >> 13;
  x *= 0xc2b2ae35U;
  return x ^ (x >> 16);
}

uint64_t mix64(uint64_t x) {
  x += 0x9e3779b97f4a7c15ULL;
  x = (x ^ (x >> 30)) * 0xbf58476d1ce4e5b9ULL;
  x = (x ^ (x >> 27)) * 0x94d049bb133111ebULL;
  return x ^ (x >> 31);
}

template<class Int>
uint32_t hash32(Int x) {
  return mix32(static_cast<uint32_t>(x));
}

template<class Int>
uint64_t hash64(Int x) {
  return mix64(static_cast<uint64_t>(x));
}

uint32_t hash_combine32(uint32_t h, uint32_t x) {
  return mix32(h ^ (x + 0x9e3779b9U + (h << 6) + (h >> 2)));
}

uint64_t hash_combine64(uint64_t h, uint64_t x) {
  return mix64(h ^ (x + 0x9e3779b97f4a7c15ULL + (h << 6) + (h >> 2)));
}

uint32_t hash_float(float x) {
  if (x == 0.0f) {
    x = 0.0f;  // Normalize -0.0.
  }
  uint32_t bits = 0;
  std::memcpy(&bits, &x, sizeof(bits));
  return mix32(bits);
}

uint64_t hash_double(double x) {
  if (x == 0.0) {
    x = 0.0;  // Normalize -0.0.
  }
  uint64_t bits = 0;
  std::memcpy(&bits, &x, sizeof(bits));
  return mix64(bits);
}

uint32_t hash_string_fnv1a32(const std::string &s) {
  uint32_t h = 2166136261U;
  for (unsigned char c : s) {
    h ^= c;
    h *= 16777619U;
  }
  return h;
}

uint64_t hash_string_fnv1a64(const std::string &s) {
  uint64_t h = 14695981039346656037ULL;
  for (unsigned char c : s) {
    h ^= c;
    h *= 1099511628211ULL;
  }
  return h;
}

template<class It>
uint32_t hash_range32(It first, It last) {
  uint32_t h = 0;
  for (It it = first; it != last; ++it) {
    h = hash_combine32(h, hash32(*it));
  }
  return h;
}

template<class It>
uint64_t hash_range64(It first, It last) {
  uint64_t h = 0;
  for (It it = first; it != last; ++it) {
    h = hash_combine64(h, hash64(*it));
  }
  return h;
}

template<class Int>
struct IntHasher {
  std::size_t operator()(Int x) const {
    static const uint64_t RAND_SEED = std::chrono::steady_clock::now().time_since_epoch().count();
    return static_cast<std::size_t>(mix64(static_cast<uint64_t>(x) + RAND_SEED));
  }
};

template<class T>
struct GenericHasher;

// Use std::hash by default.
template<class T, bool IsInteger>
struct ScalarHasher {
  std::size_t operator()(const T &x) const { return std::hash<T>()(x); }
};

// Optional: Use our seeded hasher for ints in open-hacking environments.
template<class T>
struct ScalarHasher<T, true> {
  std::size_t operator()(T x) const { return IntHasher<T>{}(x); }
};

// A self-contained hasher for std::pair<int, int> keys: copy just this struct when you do not need
// the rest of this file. The two 32-bit halves pack losslessly into one 64-bit word, which the
// SplitMix64 mixer (the same one used by mix64 above) then scrambles. Add a random seed to `key`
// before mixing, as IntHasher does, if you need anti-hacking protection.
struct PairIntHasher {
  std::size_t operator()(const std::pair<int, int> &p) const {
    uint64_t key = (static_cast<uint64_t>(static_cast<uint32_t>(p.first)) << 32) |
                   static_cast<uint32_t>(p.second);
    key += 0x9e3779b97f4a7c15ULL;
    key = (key ^ (key >> 30)) * 0xbf58476d1ce4e5b9ULL;
    key = (key ^ (key >> 27)) * 0x94d049bb133111ebULL;
    return static_cast<std::size_t>(key ^ (key >> 31));
  }
};

template<class A, class B>
struct PairHasher {
  std::size_t operator()(const std::pair<A, B> &p) const {
    uint64_t h = 0;
    h = hash_combine64(h, static_cast<uint64_t>(GenericHasher<A>()(p.first)));
    h = hash_combine64(h, static_cast<uint64_t>(GenericHasher<B>()(p.second)));
    return static_cast<std::size_t>(h);
  }
};

template<class T>
struct VectorHasher {
  std::size_t operator()(const std::vector<T> &v) const {
    uint64_t h = 0;
    for (const auto &x : v) {
      h = hash_combine64(h, static_cast<uint64_t>(GenericHasher<T>()(x)));
    }
    h = hash_combine64(h, v.size());  // Mix in the length so different sizes (e.g. empty) separate.
    return static_cast<std::size_t>(h);
  }
};

template<class... Ts>
struct TupleHasher {
  std::size_t operator()(const std::tuple<Ts...> &t) const {
    uint64_t h = 0;
    std::apply(
        [&](const Ts &...xs) {
          ((h = hash_combine64(h, static_cast<uint64_t>(GenericHasher<Ts>()(xs)))), ...);
        },
        t
    );
    return static_cast<std::size_t>(h);
  }
};

template<class T>
struct GenericHasher : ScalarHasher<T, std::is_integral<T>::value> {};

template<class A, class B>
struct GenericHasher<std::pair<A, B>> : PairHasher<A, B> {};

template<class... Ts>
struct GenericHasher<std::tuple<Ts...>> : TupleHasher<Ts...> {};

template<class T>
struct GenericHasher<std::vector<T>> : VectorHasher<T> {};

// Specializing std::hash lets pair and tuple keys be used directly in std::unordered_map and
// std::unordered_set, with no explicit hasher template argument. The C++ standard only sanctions
// specializing std::hash when a user-defined type is involved, so the functors above remain the
// portable choice; these specializations are a widely supported contest convenience. Vector keys
// are intentionally left out here to avoid clashing with the standard library's own
// std::hash<std::vector<bool>>; pass VectorHasher or GenericHasher explicitly for those.
namespace std {

template<class A, class B>
struct hash<pair<A, B>> {
  size_t operator()(const pair<A, B> &p) const { return ::GenericHasher<pair<A, B>>()(p); }
};

template<class... Ts>
struct hash<tuple<Ts...>> {
  size_t operator()(const tuple<Ts...> &t) const { return ::GenericHasher<tuple<Ts...>>()(t); }
};

}  // namespace std
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  assert(mix32(123U) == mix32(123U));
  assert(mix32(123U) != mix32(124U));
  assert(hash32(-1) == hash32(-1));
  assert(hash64(123) == mix64(123ULL));
  assert(hash_float(-0.0f) == hash_float(0.0f));
  assert(hash_double(-0.0) == hash_double(0.0));
  assert(hash_string_fnv1a32("abc") == hash_string_fnv1a32("abc"));
  assert(hash_string_fnv1a64("abc") != hash_string_fnv1a64("acb"));

  vector<int> v;
  v.push_back(1);
  v.push_back(2);
  v.push_back(3);
  assert(hash_range32(v.begin(), v.end()) == hash_range32(v.begin(), v.end()));
  assert(hash_range64(v.begin(), v.end()) == hash_range64(v.begin(), v.end()));

  unordered_map<int64_t, int, IntHasher<int64_t>> count;
  count[1000000000000LL] = 7;
  assert(count[1000000000000LL] == 7);

  using Point = pair<int, int>;
  unordered_map<Point, int, PairHasher<int, int>> dist;
  dist[Point(3, 4)] = 5;
  assert(dist[Point(3, 4)] == 5);

  // The standalone PairIntHasher is the quick choice for pair<int, int> keys.
  unordered_map<Point, int, PairIntHasher> dist_fast;
  dist_fast[Point(3, 4)] = 5;
  assert(dist_fast[Point(3, 4)] == 5);

  unordered_map<vector<int>, int, VectorHasher<int>> seen;
  seen[v] = 1;
  assert(seen[v] == 1);

  using State = pair<vector<int>, double>;
  unordered_map<State, int, GenericHasher<State>> dp;
  dp[State(v, 4.5)] = 9;
  assert(dp[State(v, 4.5)] == 9);

  // Tuple keys via the explicit functor.
  using Triple = tuple<int, char, double>;
  unordered_map<Triple, int, GenericHasher<Triple>> grid;
  grid[Triple(1, 'a', 3.14)] = 6;
  assert(grid[Triple(1, 'a', 3.14)] == 6);

  // With the std::hash specializations, pair and tuple keys need no hasher argument.
  unordered_map<Point, int> dist_default;
  dist_default[Point(3, 4)] = 5;
  assert(dist_default[Point(3, 4)] == 5);

  unordered_map<Triple, int> grid_default;
  grid_default[Triple(1, 'a', 3.14)] = 6;
  assert(grid_default[Triple(1, 'a', 3.14)] == 6);
  return 0;
}
\end{lstlisting}
\subsection{Rolling Hash}
Computes polynomial rolling hashes for sequences, supporting $O(1)$ contiguous subsequence hash queries after preprocessing. This is useful for probabilistic string matching, substring equality checks, repeated subarray detection, and binary-searching over candidate lengths.

Given a sequence $a_0, a_1, \ldots, a_{n-1}$ and a value hash $h(a_i)$ for each element, the hash is the polynomial $H(a) = \sum_{i=0}^{n-1} h(a_i) B^{n-1-i} \pmod M$, where $B$ is \inlinecode{HASH\_BASE} and $M$ is \inlinecode{HASH\_MOD}. Equivalently, it is built left-to-right by the recurrence \inlinecode{pref[i + 1] = pref[i] * HASH\_BASE + h(a[i])}. A subsequence hash for \inlinecode{[l, r)} is obtained by subtracting away the prefix before \inlinecode{l}: \inlinecode{pref[r] - pref[l] * pow(HASH\_BASE, r - l)}.

The implementation works modulo the Mersenne prime $2^{61} - 1$, using \inlinecode{\_\_uint128\_t} for the multiplication where available and falling back to a double-and-add modular multiply otherwise. The base \inlinecode{HASH\_BASE} should be changed or chosen randomly for open-hacking environments. With a fixed base, hashing remains fast and practical, but it is still probabilistic and should not be used as proof of equality when exact verification is required.

By default, each sequence value is cast to \inlinecode{uint64\_t} and mixed. For non-integer element types, pass a custom value hasher that maps each element to a stable nonzero value in \inlinecode{[1, HASH\_MOD)}.

\begin{compactitem}
  \item \inlinecode{RollingHash(first, last)} constructs prefix hashes for any iterator range of values accepted by the value hasher.
  \item \inlinecode{RollingHash(v)} constructs prefix hashes for vector \inlinecode{v}.
  \item \inlinecode{get(l, r)} returns the hash of the half-open subsequence \inlinecode{[l, r)}.
  \item \inlinecode{hash(first, last)} returns the hash of an iterator range.
  \item \inlinecode{hash(v)} returns the hash of vector \inlinecode{v}.
  \item \inlinecode{concat(left, right, right\_len)} returns the hash of the concatenation of a sequence with hash \inlinecode{left} and a sequence with hash \inlinecode{right} and length \inlinecode{right\_len}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per constructor call and whole-sequence hash, where $n$ is the sequence length.
  \item $O(1)$ per call to \inlinecode{get(l, r)} and \inlinecode{concat(left, right, right\_len)}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for storage of prefix hashes and powers, where $n$ is the maximum sequence length processed so far.
  \item $O(1)$ auxiliary per query.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <string>
#include <utility>
#include <vector>

const uint64_t HASH_MOD = (1ULL << 61) - 1;
const uint64_t HASH_BASE = 911382323ULL;  // Change or randomize.

uint64_t hash_add(uint64_t a, uint64_t b) {
  uint64_t c = a + b;
  return c >= HASH_MOD ? c - HASH_MOD : c;
}

uint64_t hash_sub(uint64_t a, uint64_t b) {
  return a >= b ? a - b : a + HASH_MOD - b;
}

uint64_t hash_mul(uint64_t a, uint64_t b) {
#if defined(__SIZEOF_INT128__)
  __uint128_t p = static_cast<__uint128_t>(a) * b;
  return hash_add(static_cast<uint64_t>(p) & HASH_MOD, static_cast<uint64_t>(p >> 61));
#else
  // Portable fallback when no 128-bit type exists: double-and-add modular multiply.
  uint64_t res = 0;
  for (a %= HASH_MOD; b > 0; b >>= 1) {
    if (b & 1) {
      res = hash_add(res, a);
    }
    a = hash_add(a, a);
  }
  return res;
#endif
}

uint64_t mix64(uint64_t x) {
  x += 0x9e3779b97f4a7c15ULL;
  x = (x ^ (x >> 30)) * 0xbf58476d1ce4e5b9ULL;
  x = (x ^ (x >> 27)) * 0x94d049bb133111ebULL;
  return x ^ (x >> 31);
}

template<class T>
struct RollingValueHasher {
  // +1 ensures the result is in [1, HASH_MOD) and never zero; a zero element in a polynomial
  // hash is invisible and enables trivial collisions.
  uint64_t operator()(const T &x) const { return mix64((uint64_t)x) % (HASH_MOD - 1) + 1; }
};

template<class T, class ValueHasher = RollingValueHasher<T>>
class RollingHash {
  static std::vector<uint64_t> pow_base;
  std::vector<uint64_t> pref;
  ValueHasher value_hasher;

  static void ensure_powers(int n) {
    while (static_cast<int>(pow_base.size()) <= n) {
      pow_base.push_back(hash_mul(pow_base.back(), HASH_BASE));
    }
  }

  template<class It>
  void build(It first, It last) {
    pref.clear();
    pref.push_back(0);
    for (; first != last; ++first) {
      pref.push_back(hash_add(hash_mul(pref.back(), HASH_BASE), value_hasher(*first)));
    }
    ensure_powers(static_cast<int>(pref.size()) - 1);
  }

 public:
  explicit RollingHash(const ValueHasher &hasher = ValueHasher()) : value_hasher(hasher) {
    ensure_powers(0);
    pref.push_back(0);
  }

  template<class It>
  RollingHash(It first, It last, const ValueHasher &hasher = ValueHasher()) : value_hasher(hasher) {
    build(first, last);
  }

  explicit RollingHash(const std::vector<T> &v, const ValueHasher &hasher = ValueHasher())
      : value_hasher(hasher) {
    build(v.begin(), v.end());
  }

  int size() const { return static_cast<int>(pref.size()) - 1; }

  uint64_t get(int lo, int hi) const {
    return hash_sub(pref[hi], hash_mul(pref[lo], pow_base[hi - lo]));
  }

  template<class It>
  static uint64_t hash(It first, It last, const ValueHasher &hasher = ValueHasher()) {
    RollingHash<T, ValueHasher> h(first, last, hasher);
    return h.get(0, h.size());
  }

  static uint64_t hash(const std::vector<T> &v, const ValueHasher &hasher = ValueHasher()) {
    RollingHash<T, ValueHasher> h(v, hasher);
    return h.get(0, h.size());
  }

  static uint64_t concat(uint64_t left, uint64_t right, int right_len) {
    ensure_powers(right_len);
    return hash_add(hash_mul(left, pow_base[right_len]), right);
  }
};

template<class T, class ValueHasher>
std::vector<uint64_t> RollingHash<T, ValueHasher>::pow_base(1, 1);
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

using PointI = pair<int, int>;

struct PointHasher {
  uint64_t operator()(const PointI &p) const {
    // Pack the two 32-bit coordinates losslessly into one 64-bit key before mixing, so distinct
    // points never collide pre-mix (a small-multiplier combine like a * C + b would).
    uint64_t key = (static_cast<uint64_t>(static_cast<uint32_t>(p.first)) << 32) |
                   static_cast<uint32_t>(p.second);
    return mix64(key) % (HASH_MOD - 1) + 1;
  }
};

int main() {
  string s = "abracadabra";
  RollingHash<char> hs(s.begin(), s.end());
  assert(hs.get(0, 4) == hs.get(7, 11));  // "abra" == "abra"
  assert(hs.get(0, 4) != hs.get(3, 7));   // "abra" != "acad"

  string a = "abc", b = "def";
  uint64_t ab = RollingHash<char>::concat(
      RollingHash<char>::hash(a.begin(), a.end()), RollingHash<char>::hash(b.begin(), b.end()),
      b.size()
  );
  string c = a + b;
  assert(ab == RollingHash<char>::hash(c.begin(), c.end()));

  vector<int> v;
  v.push_back(1);
  v.push_back(2);
  v.push_back(3);
  v.push_back(1);
  v.push_back(2);
  RollingHash<int> hv(v);
  assert(hv.get(0, 2) == hv.get(3, 5));  // [1, 2] == [1, 2]
  assert(hv.get(0, 3) == RollingHash<int>::hash(v.begin(), v.begin() + 3));

  vector<PointI> poly{{1, 2}, {3, 4}};
  RollingHash<PointI, PointHasher> hp(poly);
  assert((hp.get(0, 2) == RollingHash<PointI, PointHasher>::hash(poly)));
  return 0;
}
\end{lstlisting}
\subsection{Dynamic Rolling Hash}
Maintain a polynomial rolling hash over a mutable sequence, supporting point updates and $O(\log n)$ subsequence hash queries. Unlike the static rolling hash, which fixes its prefix hashes at construction, this structure allows any element to change between queries, which is useful for problems that compare substrings while characters are edited.

The key idea is to weight each element by a power of the base tied to its absolute position: storing $a_i B^i$ at index \inlinecode{i} turns the hash of any prefix into an ordinary prefix sum, which a Fenwick tree maintains under point updates. The raw sum over a half-open range \inlinecode{[lo, hi)} is then $\sum_i a_i B^i$, and multiplying by $B^{-{\inlinecodemath{lo}}}$ shifts it to the position-independent value $\sum_i a_i B^{i - {\inlinecodemath{lo}}}$, so two ranges with equal contents receive equal hashes (the leftmost element carries the lowest power). A segment tree could replace the Fenwick tree if range operations (such as lazy range assignment) were needed.

\inlinecode{MOD1} and \inlinecode{MOD2} are two distinct primes, each kept below $2^{31}$ so that the product of any two residues fits in a 64-bit integer; hashing modulo both and comparing the resulting pair makes a collision astronomically unlikely. \inlinecode{BASE1} and \inlinecode{BASE2} are the corresponding polynomial bases, which should exceed the largest element value and may be randomized per run to resist adversarial inputs.

Element values must be non-negative. As with any polynomial hash, comparing subsequences of different lengths is only safe when values are positive (for example, map an alphabet to $[1, B)$) so that trailing zeros cannot make a shorter sequence collide with a longer one.

\begin{compactitem}
  \item \inlinecode{DynamicHash(a)} builds prefix hashes over the integer sequence \inlinecode{a}.
  \item \inlinecode{size()} returns the length of the sequence.
  \item \inlinecode{set(i, x)} assigns the value at index \inlinecode{i} to \inlinecode{x}.
  \item \inlinecode{hash(lo, hi)} returns the pair of hashes of the half-open subsequence \inlinecode{[lo, hi)}. Two ranges are equal with high probability if and only if their \inlinecode{hash()} pairs are equal.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log n)$ for the constructor, where $n$ is the length of the sequence.
  \item $O(\log n)$ per call to \inlinecode{set()} and \inlinecode{hash()}.
  \item $O(1)$ per call to \inlinecode{size()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for storage.
  \item $O(1)$ auxiliary per query.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <utility>
#include <vector>

class DynamicHash {
  static const uint64_t MOD1 = 1000000007, MOD2 = 1000000009;  // Distinct primes below 2^31.
  static const uint64_t BASE1 = 131, BASE2 = 137;  // Should exceed the maximum element value.

  int len;
  std::vector<int> cur;
  std::vector<uint64_t> pw1, pw2, ip1, ip2;  // Base powers and inverse base powers.
  std::vector<uint64_t> ft1, ft2;            // Fenwick trees, 1-indexed of size len + 1.

  // Products and sums of residues below 2^30 stay under 2^60, so 64-bit arithmetic never overflows.
  static uint64_t power(uint64_t b, uint64_t e, uint64_t m) {
    uint64_t res = 1;
    for (b %= m; e > 0; e >>= 1) {
      if (e & 1) {
        res = res * b % m;
      }
      b = b * b % m;
    }
    return res;
  }

  void add(std::vector<uint64_t> &ft, int i, uint64_t delta, uint64_t m) {
    for (i++; i <= len; i += i & -i) {
      ft[i] = (ft[i] + delta) % m;
    }
  }

  uint64_t sum(const std::vector<uint64_t> &ft, int i, uint64_t m) const {
    uint64_t res = 0;
    for (i++; i > 0; i -= i & -i) {
      res = (res + ft[i]) % m;
    }
    return res;
  }

 public:
  explicit DynamicHash(const std::vector<int> &a) : len(static_cast<int>(a.size())), cur(len, 0) {
    pw1.resize(len + 1);
    pw2.resize(len + 1);
    ip1.resize(len + 1);
    ip2.resize(len + 1);
    pw1[0] = pw2[0] = ip1[0] = ip2[0] = 1;
    uint64_t invb1 = power(BASE1, MOD1 - 2, MOD1), invb2 = power(BASE2, MOD2 - 2, MOD2);
    for (int i = 1; i <= len; i++) {
      pw1[i] = pw1[i - 1] * BASE1 % MOD1;
      pw2[i] = pw2[i - 1] * BASE2 % MOD2;
      ip1[i] = ip1[i - 1] * invb1 % MOD1;
      ip2[i] = ip2[i - 1] * invb2 % MOD2;
    }
    ft1.assign(len + 1, 0);
    ft2.assign(len + 1, 0);
    for (int i = 0; i < len; i++) {
      set(i, a[i]);
    }
  }

  int size() const { return len; }

  void set(int i, int x) {
    uint64_t old1 = static_cast<uint64_t>(cur[i]) % MOD1 * pw1[i] % MOD1;
    uint64_t old2 = static_cast<uint64_t>(cur[i]) % MOD2 * pw2[i] % MOD2;
    uint64_t new1 = static_cast<uint64_t>(x) % MOD1 * pw1[i] % MOD1;
    uint64_t new2 = static_cast<uint64_t>(x) % MOD2 * pw2[i] % MOD2;
    add(ft1, i, (new1 + MOD1 - old1) % MOD1, MOD1);
    add(ft2, i, (new2 + MOD2 - old2) % MOD2, MOD2);
    cur[i] = x;
  }

  std::pair<uint64_t, uint64_t> hash(int lo, int hi) const {
    uint64_t raw1 = (sum(ft1, hi - 1, MOD1) + MOD1 - sum(ft1, lo - 1, MOD1)) % MOD1;
    uint64_t raw2 = (sum(ft2, hi - 1, MOD2) + MOD2 - sum(ft2, lo - 1, MOD2)) % MOD2;
    return {raw1 * ip1[lo] % MOD1, raw2 * ip2[lo] % MOD2};
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <string>
using namespace std;

int main() {
  string s = "abcabc";
  vector<int> a(s.begin(), s.end());
  DynamicHash h(a);
  assert(h.size() == 6);
  assert(h.hash(0, 3) == h.hash(3, 6));  // "abc" == "abc".

  h.set(0, 'x');  // "xbcabc".
  assert(!(h.hash(0, 3) == h.hash(3, 6)));
  h.set(3, 'x');                         // "xbcxbc".
  assert(h.hash(0, 3) == h.hash(3, 6));  // "xbc" == "xbc".

  // A single element's hash is invariant under where an equal value sits.
  assert(h.hash(1, 2) == h.hash(4, 5));  // both 'b'.
  return 0;
}
\end{lstlisting}
\subsection{Zobrist Hashing}
Computes randomized XOR fingerprints for keys and sets of keys using Zobrist hashing. Each distinct key is assigned a pseudorandom 64-bit token, and a set is represented by the XOR of its members' tokens. This makes insertion and deletion the same operation: XOR the key's token once.

This technique is useful for probabilistic equality checks on sets, distinct-prefix queries, board-game states, and other unordered collections. It is not cryptographic, and collisions are possible with probability roughly $q^2 / 2^{64}$ over $q$ compared fingerprints. For multisets, plain XOR only records element parity; use a summed hash or pair it with counts if multiplicity matters.

\begin{compactitem}
  \item \inlinecode{ZobristHash(seed)} constructs a token generator with initial value \inlinecode{seed}.
  \item \inlinecode{get(x)} returns the stable token assigned to key \inlinecode{x}, creating it if needed.
  \item \inlinecode{toggle(h, x)} returns the hash obtained by inserting \inlinecode{x} into set hash \inlinecode{h} if absent, or removing \inlinecode{x} from \inlinecode{h} if present.
  \item \inlinecode{distinct\_prefix\_hashes(lo, hi)} returns prefix hashes where each distinct key in the range \inlinecode{[lo, hi)} contributes only on its first occurrence.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(1)$ expected per call to \inlinecode{get(x)} and \inlinecode{toggle(h, x)}.
  \item $O(n)$ expected per call to \inlinecode{distinct\_prefix\_hashes(lo, hi)}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for stored key tokens, where $n$ is the number of distinct keys seen so far.
  \item $O(n)$ auxiliary for \inlinecode{distinct\_prefix\_hashes(lo, hi)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <unordered_map>
#include <unordered_set>
#include <vector>

template<class T>
class ZobristHash {
  std::unordered_map<T, uint64_t> token;
  uint64_t state;

  static uint64_t splitmix64(uint64_t x) {
    x += 0x9e3779b97f4a7c15ULL;
    x = (x ^ (x >> 30)) * 0xbf58476d1ce4e5b9ULL;
    x = (x ^ (x >> 27)) * 0x94d049bb133111ebULL;
    return x ^ (x >> 31);
  }

 public:
  explicit ZobristHash(uint64_t seed = 0) : state(seed) {}

  uint64_t get(const T &x) {
    if (auto it = token.find(x); it != token.end()) {
      return it->second;
    }
    state = splitmix64(state);
    token[x] = state;
    return state;
  }

  uint64_t toggle(uint64_t h, const T &x) { return h ^ get(x); }

  template<class It>
  std::vector<uint64_t> distinct_prefix_hashes(It lo, It hi) {
    std::unordered_set<T> seen;
    std::vector<uint64_t> res(1, 0);
    uint64_t h = 0;
    for (It it = lo; it != hi; ++it) {
      if (seen.insert(*it).second) {
        h ^= get(*it);
      }
      res.push_back(h);
    }
    return res;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  vector<int> a{1, 2, 1, 3};
  vector<int> b{2, 1, 3};
  ZobristHash<int> zh(123);
  vector<uint64_t> pa = zh.distinct_prefix_hashes(a.begin(), a.end());
  vector<uint64_t> pb = zh.distinct_prefix_hashes(b.begin(), b.end());

  assert(pa[2] == pb[2]);  // Distinct sets {1, 2} and {2, 1}.
  assert(pa[4] == pb[3]);  // Distinct sets {1, 2, 3} and {2, 1, 3}.

  uint64_t h = 0;
  h = zh.toggle(h, 5);
  assert(h != 0);
  h = zh.toggle(h, 5);
  assert(h == 0);
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Pattern Matching}{3.3 Pattern Matching}}
\setcounter{section}{3}
\setcounter{subsection}{0}
\subsection{String Searching (KMP)}
Given a single string (needle) and subsequent queries of texts (haystacks) to be searched, determine the first positions in which the needle occurs within the given haystacks in linear time using the Knuth-Morris-Pratt algorithm. In comparison, \inlinecode{std::string::find} runs in $O(n^{2})$ in the worst case. KMP precomputes, for every prefix of the needle, the length of its longest proper border (a prefix that is also a suffix). On a mismatch, the needle falls back to that border instead of restarting, so the position in the haystack never moves backward.

\begin{compactitem}
  \item \inlinecode{KMP(needle)} constructs the partial match table for a string \inlinecode{needle} that is to be searched for subsequently in \inlinecode{haystack} queries.
  \item \inlinecode{find\_in(haystack)} returns the first position that \inlinecode{needle} occurs in \inlinecode{haystack}, or \inlinecode{std::string::npos} if it cannot be found. Note that the function can be modified to return all matches by simply letting the loop run and storing the results instead of returning early.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(m)$ per call to the constructor, where $m$ is the length of \inlinecode{needle}.
  \item $O(n)$ per call to \inlinecode{find\_in(haystack)}, where $n$ is the length of \inlinecode{haystack}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m)$ for storage of the partial match table, where $m$ is the length of \inlinecode{needle}.
  \item $O(1)$ auxiliary space per call to \inlinecode{find\_in(haystack)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <string>
#include <vector>
using std::string;

class KMP {
  string needle;
  std::vector<int> table;

 public:
  explicit KMP(const string &needle) : needle(needle), table(needle.size()) {
    for (int i = 1, j = 0; i < static_cast<int>(needle.size()); i++) {
      while (j > 0 && needle[i] != needle[j]) {
        j = table[j - 1];
      }
      if (needle[i] == needle[j]) {
        j++;
      }
      table[i] = j;
    }
  }

  size_t find_in(const string &haystack) {
    int m = static_cast<int>(needle.size());
    if (m == 0) {
      return 0;
    }
    for (int i = 0, j = 0; i < static_cast<int>(haystack.size()); i++) {
      while (j > 0 && needle[j] != haystack[i]) {
        j = table[j - 1];
      }
      if (needle[j] == haystack[i]) {
        j++;
      }
      if (j == m) {
        return i + 1 - m;
      }
    }
    return string::npos;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(15 == KMP("ABCDABD").find_in("ABC ABCDAB ABCDABCDABDE"));
  return 0;
}
\end{lstlisting}
\subsection{String Searching (Z Algorithm)}
Given a single string (needle) and a single text (haystack) to be searched, determine the first position in which the needle occurs within the haystack in linear time using the Z algorithm. In comparison, \inlinecode{std::string::find} runs in quadratic time.

The Z array stores, for each position of a string, the length of the longest substring starting there that matches a prefix of the whole string. It is computed in linear time by maintaining the rightmost matched window and seeding each position from its mirror earlier in that window.

The \inlinecode{find()} function below calls the Z algorithm on the concatenation of \inlinecode{needle} and \inlinecode{haystack}, separated by a sentinel value that is guaranteed not to collide with any byte in either input. Any position whose Z value reaches the needle's length marks an occurrence.

\begin{compactitem}
  \item \inlinecode{z\_array(s)} constructs the Z array for a needle string \inlinecode{s} that can be used for string searching. Each resulting value \inlinecode{z[i]} stores the length of the longest substring starting from \inlinecode{s[i]} which is also a prefix of the needle \inlinecode{s}.
  \item \inlinecode{find(haystack, needle)} returns the first position that \inlinecode{needle} occurs in \inlinecode{haystack}, or \inlinecode{std::string::npos} if it cannot be found. Note that the function can be modified to return all matches by simply letting the loop run and storing the results instead of returning early.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to \inlinecode{z\_array(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(n + m)$ per call to \inlinecode{find(haystack, needle)}, where $n$ is the length of \inlinecode{haystack} and $m$ is the length of \inlinecode{needle}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary heap space for \inlinecode{z\_array(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(n + m)$ auxiliary heap space for \inlinecode{find(haystack, needle)} where $n$ is the length of \inlinecode{haystack} and $m$ is the length of \inlinecode{needle}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <string>
#include <vector>
using std::string;

template<class Seq>
std::vector<int> z_array(const Seq &s) {
  std::vector<int> z(s.size());
  for (int i = 1, l = 0, r = 0; i < static_cast<int>(z.size()); i++) {
    if (i <= r) {
      z[i] = std::min(r - i + 1, z[i - l]);
    }
    while (i + z[i] < static_cast<int>(z.size()) && s[z[i]] == s[i + z[i]]) {
      z[i]++;
    }
    if (r < i + z[i] - 1) {
      l = i;
      r = i + z[i] - 1;
    }
  }
  return z;
}

size_t find(const string &haystack, const string &needle) {
  if (needle.empty()) {
    return 0;
  }
  std::vector<int> s;
  s.reserve(needle.size() + haystack.size() + 1);
  for (unsigned char c : needle) {
    s.push_back(c + 1);
  }
  s.push_back(0);
  for (unsigned char c : haystack) {
    s.push_back(c + 1);
  }
  auto z = z_array(s);
  int m = static_cast<int>(needle.size());
  for (int i = m + 1; i < static_cast<int>(z.size()); i++) {
    if (z[i] == m) {
      return i - m - 1;
    }
  }
  return string::npos;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(15 == find("ABC ABCDAB ABCDABCDABDE", "ABCDABD"));
  return 0;
}
\end{lstlisting}
\subsection{String Searching (Aho-Corasick)}
Given a set of strings (needles) and subsequent queries of texts (haystacks) to be searched, determine all positions in which needles occur within the given haystacks in linear time using the Aho-Corasick algorithm. The needles are arranged in a trie whose nodes carry failure links to the longest proper suffix that is also a trie node. Scanning the haystack once while following transitions and failure links then reports every match through each node's output set.

This implementation stores transitions in hash tables, which keeps lookups expected constant time without assuming a fixed alphabet size. The output tables are stored as ordered sets so matches ending at the same position are reported in deterministic needle order.

\begin{compactitem}
  \item \inlinecode{AhoCorasick(needles)} constructs the finite-state automaton for a set of needle strings that are to be searched for subsequently in haystack queries. Empty needles are ignored.
  \item \inlinecode{find\_all\_in(haystack, report\_match)} calls the function \inlinecode{report\_match(s, pos)} once on each occurrence of each needle that occurs in the \inlinecode{haystack}, where \inlinecode{pos} is the starting position in the \inlinecode{haystack} at which string \inlinecode{s} (a matched needle) occurs. The matches will be reported in increasing order of their ending positions within the \inlinecode{haystack}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(m \cdot l + z \log m)$ expected per call to the constructor, where $m$ is the number of needles, $l$ is the maximum length of any needle, and $z$ is the total number of inherited output entries.
  \item $O(n + z)$ expected per call to \inlinecode{find\_all\_in(haystack, report\_match)}, where $n$ is the length of \inlinecode{haystack} and $z$ is the number of matches.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m \cdot l)$ for storage of the automaton, where $m$ is the number of needles and $l$ is the maximum length for any needle.
  \item $O(1)$ auxiliary space per call to \inlinecode{find\_all\_in(haystack, report\_match)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <queue>
#include <set>
#include <string>
#include <unordered_map>
#include <vector>
using std::string;

class AhoCorasick {
  std::vector<string> needles;
  std::vector<int> fail;
  std::vector<std::unordered_map<char, int>> adj;
  std::vector<std::set<int>> out;

  int next_state(int curr, char c) {
    int next = curr;
    while (next != 0 && adj[next].find(c) == adj[next].end()) {
      next = fail[next];
    }
    auto it = adj[next].find(c);
    return (it != adj[next].end()) ? it->second : 0;
  }

 public:
  explicit AhoCorasick(const std::vector<string> &needles) : needles(needles) {
    int total_len = 0;
    for (const auto &needle : needles) {
      total_len += static_cast<int>(needle.size());
    }
    // The trie has at most total_len + 1 states: the root plus one per character when no prefixes
    // are shared. Sizing to total_len alone overflows for low-sharing needle sets (e.g. a single
    // needle, which still needs the root plus one node).
    int max_states = total_len + 1;
    fail.resize(max_states, -1);
    adj.resize(max_states);
    out.resize(max_states);
    int states = 1;
    for (int i = 0; i < static_cast<int>(needles.size()); i++) {
      if (needles[i].empty()) {
        continue;
      }
      int curr = 0;
      for (char c : needles[i]) {
        if (auto it = adj[curr].find(c); it != adj[curr].end()) {
          curr = it->second;
        } else {
          curr = adj[curr][c] = states++;
        }
      }
      out[curr].insert(i);
    }
    std::queue<int> q;
    for (auto &[c, v] : adj[0]) {
      if (v != 0) {
        fail[v] = 0;
        q.push(v);
      }
    }
    while (!q.empty()) {
      int u = q.front();
      q.pop();
      for (auto &[c, v] : adj[u]) {
        int f = fail[u];
        while (f != 0 && adj[f].find(c) == adj[f].end()) {
          f = fail[f];
        }
        auto fit = adj[f].find(c);
        f = (fit != adj[f].end()) ? fit->second : 0;
        fail[v] = f;
        out[v].insert(out[f].begin(), out[f].end());
        q.push(v);
      }
    }
  }

  template<class Fn>
  void find_all_in(const string &haystack, Fn report_match) {
    int state = 0;
    for (int i = 0; i < static_cast<int>(haystack.size()); i++) {
      state = next_state(state, haystack[i]);
      for (int idx : out[state]) {
        report_match(needles[idx], i - static_cast<int>(needles[idx].size()) + 1);
      }
    }
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <iostream>
using namespace std;

void report_match(const string &needle, int pos) {
  cout << "Matched \"" << needle << "\" at position " << pos << "." << endl;
}

int main() {
  vector<string> needles;
  needles.push_back("a");
  needles.push_back("ab");
  needles.push_back("bab");
  needles.push_back("bc");
  needles.push_back("bca");
  needles.push_back("c");
  needles.push_back("caa");
  needles.push_back("abccab");

  AhoCorasick(needles).find_all_in("abccab", report_match);
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
Matched "a" at position 0.
Matched "ab" at position 0.
Matched "bc" at position 1.
Matched "c" at position 2.
Matched "c" at position 3.
Matched "a" at position 4.
Matched "ab" at position 4.
Matched "abccab" at position 0.
\end{exampleoutput}
\subsection{Palindrome Searching (Manacher)}
Computes all odd- and even-length palindromic radii of a string in linear time using Manacher's algorithm. These radii can be used to test whether any substring is a palindrome in $O(1)$, find the longest palindromic substring, or count all palindromic substrings. The algorithm tracks the rightmost known palindrome and seeds each new center with the radius of its mirror inside that window, so characters are only ever compared when a palindrome is extended past the window.

For odd palindromes, \inlinecode{odd[i]} is the radius centered at character \inlinecode{s[i]}, including the center character. Thus the palindrome spans \inlinecode{[i - odd[i] + 1, i + odd[i])}. For even palindromes, \inlinecode{even[i]} is the radius centered between \inlinecode{s[i - 1]} and \inlinecode{s[i]}. Thus the palindrome spans \inlinecode{[i - even[i], i + even[i])}.

\begin{compactitem}
  \item \inlinecode{Manacher(s)} constructs the palindromic radii arrays for string \inlinecode{s}.
  \item \inlinecode{is\_palindrome(lo, hi)} returns whether substring \inlinecode{[lo, hi)} is a palindrome.
  \item \inlinecode{longest\_palindrome()} returns one longest palindromic substring of \inlinecode{s}.
  \item \inlinecode{count\_palindromes()} returns the total number of palindromic substrings of \inlinecode{s}, counted with multiplicity by position.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per constructor call, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ per call to \inlinecode{is\_palindrome(lo, r)}.
  \item $O(n)$ per call to \inlinecode{longest\_palindrome()} and \inlinecode{count\_palindromes()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ for storage of the radii arrays.
  \item $O(1)$ auxiliary per query.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstdint>
#include <string>
#include <vector>
using std::string;

class Manacher {
  string s;
  std::vector<int> odd, even;

 public:
  explicit Manacher(const string &s) : s(s), odd(s.size()), even(s.size()) {
    int n = static_cast<int>(s.size());
    for (int i = 0, l = 0, r = -1; i < n; i++) {
      int k = (i > r) ? 1 : std::min(odd[l + r - i], r - i + 1);
      while (0 <= i - k && i + k < n && s[i - k] == s[i + k]) {
        k++;
      }
      odd[i] = k;
      if (i + k - 1 > r) {
        l = i - k + 1;
        r = i + k - 1;
      }
    }
    for (int i = 0, l = 0, r = -1; i < n; i++) {
      int k = (i > r) ? 0 : std::min(even[l + r - i + 1], r - i + 1);
      while (0 <= i - k - 1 && i + k < n && s[i - k - 1] == s[i + k]) {
        k++;
      }
      even[i] = k;
      if (i + k - 1 > r) {
        l = i - k;
        r = i + k - 1;
      }
    }
  }

  bool is_palindrome(int lo, int hi) const {
    int len = hi - lo;
    if (len <= 0) {
      return true;
    }
    int mid = lo + (hi - lo) / 2;
    return (len % 2) ? odd[mid] >= len / 2 + 1 : even[mid] >= len / 2;
  }

  string longest_palindrome() const {
    int best_l = 0, best_len = 0;
    int n = static_cast<int>(s.size());
    for (int i = 0; i < n; i++) {
      int len = 2 * odd[i] - 1;
      if (len > best_len) {
        best_len = len;
        best_l = i - odd[i] + 1;
      }
      len = 2 * even[i];
      if (len > best_len) {
        best_len = len;
        best_l = i - even[i];
      }
    }
    return s.substr(best_l, best_len);
  }

  int64_t count_palindromes() const {
    int64_t res = 0;
    for (int i = 0; i < static_cast<int>(s.size()); i++) {
      res += odd[i] + even[i];
    }
    return res;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  Manacher m("abacaba");
  assert(m.is_palindrome(0, 7));
  assert(m.is_palindrome(1, 6));
  assert(!m.is_palindrome(0, 6));
  assert(m.longest_palindrome() == "abacaba");
  assert(m.count_palindromes() == 12);

  Manacher e("cabbad");
  assert(e.is_palindrome(1, 5));
  assert(e.longest_palindrome() == "abba");
  return 0;
}
\end{lstlisting}
\subsection{Lyndon Factorization (Duval)}
A Lyndon word is a nonempty string that is strictly smaller than all of its proper suffixes, or equivalently, strictly smaller than all of its nontrivial cyclic rotations. By the Chen-Fox-Lyndon theorem, every string has a unique factorization into a sequence of Lyndon words whose values are non-increasing: $s = w_1 w_2 \cdots w_k$ with $w_1 \geq w_2 \geq \cdots \geq w_k$. Duval's algorithm computes this factorization in linear time and constant auxiliary space by scanning the string while tracking the current candidate Lyndon prefix and the position it last matched.

The same scan, applied to the string viewed as a cycle, yields the starting position of its lexicographically least rotation. This is the standard way to canonicalize a necklace, for example to test whether two strings are cyclic rotations of one another.

\begin{compactitem}
  \item \inlinecode{lyndon\_factorization(s)} returns a vector of the starting indices of the Lyndon factors of \inlinecode{s}, in order. The factor at index \inlinecode{i} of the result spans the half-open range from \inlinecode{res[i]} to \inlinecode{res[i + 1]} (or to \inlinecode{s.size()} for the final factor).
  \item \inlinecode{min\_rotation(s)} returns the 0-indexed position \inlinecode{p} such that the rotation beginning at \inlinecode{p}, that is, \inlinecode{s[p], s[(p + 1) \% n], ..., s[(p + n - 1) \% n]}, is lexicographically smallest. If several rotations tie, the smallest such \inlinecode{p} is returned.
\end{compactitem}

Both functions operate on any random-access sequence whose elements support \inlinecode{operator\textless{}}.

\textbf{Time Complexity}: $O(n)$ per call to \inlinecode{lyndon\_factorization(s)} and \inlinecode{min\_rotation(s)}, where $n$ is the length of \inlinecode{s}.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(k)$ auxiliary heap space for \inlinecode{lyndon\_factorization(s)}, where $k$ is the number of factors.
  \item $O(1)$ auxiliary space for \inlinecode{min\_rotation(s)}.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <vector>

template<class Seq>
std::vector<int> lyndon_factorization(const Seq &s) {
  int n = static_cast<int>(s.size());
  std::vector<int> starts;
  int i = 0;
  while (i < n) {
    int j = i + 1, k = i;
    while (j < n && !(s[j] < s[k])) {
      k = (s[k] < s[j]) ? i : k + 1;
      j++;
    }
    while (i <= k) {
      starts.push_back(i);
      i += j - k;
    }
  }
  return starts;
}

template<class Seq>
int min_rotation(const Seq &s) {
  int n = static_cast<int>(s.size());
  if (n == 0) {
    return 0;
  }
  int i = 0, res = 0;
  while (i < n) {
    res = i;
    int j = i + 1, k = i;
    while (j < 2 * n && !(s[j % n] < s[k % n])) {
      k = (s[k % n] < s[j % n]) ? i : k + 1;
      j++;
    }
    while (i <= k) {
      i += j - k;
    }
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <string>
using std::string;

int main() {
  // "banana" factors as "b" >= "an" >= "an" >= "a".
  assert((lyndon_factorization(string("banana")) == std::vector<int>{0, 1, 3, 5}));

  // Among the rotations of "baca", "abac" (starting at index 3) is smallest.
  assert(min_rotation(string("baca")) == 3);

  // Both routines work on any comparable sequence, not just strings.
  assert(min_rotation(std::vector<int>{3, 1, 2, 1, 3}) == 1);
  return 0;
}
\end{lstlisting}
\subsection{De Bruijn Sequence}
A De Bruijn sequence $B(k, n)$ is a cyclic string over the alphabet $[0, k)$ of length $k^n$ in which every one of the $k^n$ possible length-$n$ strings appears exactly once as a (cyclic) substring. Reading a window of $n$ symbols and advancing one position at a time therefore cycles through all length-$n$ strings without repetition, which makes the sequence a compact way to probe every short pattern: cracking a rotating lock, addressing a shift register, or building the perfect-hash table that maps an isolated set bit to its index.

The construction here is the Fredricksen-Kessler-Maiorana (FKM) algorithm. It concatenates, in lexicographic order, the Lyndon words (see the Lyndon factorization section) whose length divides $n$. The result is the lexicographically smallest De Bruijn sequence. The recursion enumerates these necklace representatives directly, appending the first $p$ symbols of each candidate whenever its period $p$ divides $n$.

\begin{compactitem}
  \item \inlinecode{de\_bruijn(k, n)} returns a vector of length $k^n$ holding one period of $B(k, n)$. Each entry lies in $[0, k)$. Both $k$ and $n$ must be at least 1. Treat the result as cyclic: the length-$n$ windows that wrap past the end are exactly the ones still needed to cover all strings.
\end{compactitem}

\textbf{Time Complexity}: $O(k^{n})$ per call to \inlinecode{de\_bruijn(k, n)}, which is optimal since the output itself has length $k^n$.

\textbf{Space Complexity}: $O(k^{n})$ heap space for the returned sequence, plus $O(n)$ auxiliary space for the recursion.

\begin{lstlisting}[language=C++]
#include <vector>

namespace {

void de_bruijn_gen(int t, int p, int k, int n, std::vector<int> &word, std::vector<int> &seq) {
  if (t > n) {
    if (n % p == 0) {
      for (int i = 1; i <= p; i++) {
        seq.push_back(word[i]);
      }
    }
    return;
  }
  word[t] = word[t - p];
  de_bruijn_gen(t + 1, p, k, n, word, seq);
  for (int j = word[t - p] + 1; j < k; j++) {
    word[t] = j;
    de_bruijn_gen(t + 1, t, k, n, word, seq);
  }
}

}  // namespace

std::vector<int> de_bruijn(int k, int n) {
  std::vector<int> word(n + 1, 0), seq;
  de_bruijn_gen(1, 1, k, n, word, seq);
  return seq;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cmath>
#include <set>
#include <vector>
using namespace std;

int main() {
  // B(2, 3) has length 8 and every 3-bit string appears once cyclically.
  vector<int> seq = de_bruijn(2, 3);
  assert(seq.size() == 8);

  for (int k = 2; k <= 4; k++) {
    for (int n = 1; n <= 4; n++) {
      vector<int> s = de_bruijn(k, n);
      int len = static_cast<int>(std::lround(std::pow(k, n)));
      assert(static_cast<int>(s.size()) == len);

      // Collect every length-n window, wrapping cyclically, and confirm all k^n are distinct.
      set<vector<int>> windows;
      for (int i = 0; i < len; i++) {
        vector<int> w(n);
        for (int j = 0; j < n; j++) {
          w[j] = s[(i + j) % len];
        }
        windows.insert(w);
      }
      assert(static_cast<int>(windows.size()) == len);
    }
  }
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Suffix Arrays and LCP}{3.4 Suffix Arrays and LCP}}
\setcounter{section}{4}
\setcounter{subsection}{0}
\subsection{Suffix Array and LCP (Manber-Myers)}
Given a string \inlinecode{s}, a suffix array is the array of the smallest starting positions for the sorted suffixes of \inlinecode{s}. That is, the $i$-th position of the suffix array stores the starting position of the $i$-th lexicographically smallest suffix of \inlinecode{s}. For example, \inlinecode{s = "cab"} has the suffixes \inlinecode{"cab"}, \inlinecode{"ab"}, and \inlinecode{"b"}. When sorted, the indices of the suffixes are \inlinecode{"ab"}, \inlinecode{"b"}, and \inlinecode{"cab"}, so the suffix array (assuming 0-based indices) is $[1, 2, 0]$.

For a string \inlinecode{s} of length $n$, the longest common prefix (LCP) array of length $n - 1$ stores the lengths of the longest common prefixes between all pairs of lexicographically adjacent suffixes in \inlinecode{s}. For example, \inlinecode{"baa"} has the sorted suffixes \inlinecode{"a"}, \inlinecode{"aa"}, and \inlinecode{"baa"}, with an LCP array of $[1, 0]$.

The doubling algorithm sorts the suffixes by their first $2^k$ characters for increasing $k$: each round orders suffixes by their pair of ranks from the previous round, so a comparison costs $O(1)$ and the full order emerges after $O(\log n)$ rounds.

\begin{compactitem}
  \item \inlinecode{SuffixArrayManberMyers(s)} constructs a suffix array from the given string \inlinecode{s} using the original Manber-Myers doubling algorithm with a comparison-based sort.
  \item \inlinecode{get\_sa()} returns the constructed suffix array.
  \item \inlinecode{get\_lcp()} returns the corresponding LCP array for the suffix array.
  \item \inlinecode{find(needle)} returns one position that \inlinecode{needle} occurs in \inlinecode{s} (not necessarily the first), or \inlinecode{std::string::npos} if it cannot be found. For a \inlinecode{needle} of length $m$, this implementation uses an $O(m \log n)$ binary search, but can be optimized to $O(m + \log n)$ by first computing the LCP-LR array using the LCP array.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log^{2} n)$ per call to the constructor, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ per call to \inlinecode{get\_sa()}.
  \item $O(n)$ per call to \inlinecode{get\_lcp()}, where $n$ is the length of \inlinecode{s}.
  \item $O(m \log n)$ per call to \inlinecode{find(needle)}, where $m$ is the length of \inlinecode{needle} and $n$ is the length of \inlinecode{s}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary for storage of the suffix and LCP arrays, where $n$ is the length of \inlinecode{s}.
  \item $O(n)$ auxiliary heap space for the constructor.
  \item $O(1)$ auxiliary space for all other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <numeric>
#include <string>
#include <utility>
#include <vector>
using std::string;

class SuffixArrayManberMyers {
  string s;
  std::vector<int> sa, rk;

 public:
  explicit SuffixArrayManberMyers(const string &s) : s(s), sa(s.size()), rk(s.size()) {
    int n = static_cast<int>(s.size());
    std::iota(sa.begin(), sa.end(), 0);
    for (int i = 0; i < n; i++) {
      rk[i] = static_cast<unsigned char>(s[i]);
    }
    std::vector<std::pair<int, int>> rk2(n);
    for (int gap = 1; gap < n; gap *= 2) {
      for (int i = 0; i < n; i++) {
        rk2[i] = {rk[i], i + gap < n ? rk[i + gap] + 1 : 0};
      }
      std::sort(sa.begin(), sa.end(), [&](int i, int j) { return rk2[i] < rk2[j]; });
      for (int i = 0; i < n; i++) {
        rk[sa[i]] = (i > 0 && rk2[sa[i - 1]] == rk2[sa[i]]) ? rk[sa[i - 1]] : i;
      }
    }
  }

  const std::vector<int> &get_sa() const { return sa; }

  std::vector<int> get_lcp() const {
    int n = static_cast<int>(s.size());
    if (n == 0) {
      return {};  // Avoid constructing a vector of size (size_t)(-1).
    }
    std::vector<int> lcp(n - 1);
    for (int i = 0, k = 0; i < n; i++) {
      if (rk[i] < n - 1) {
        int j = sa[rk[i] + 1];
        while (std::max(i, j) + k < n && s[i + k] == s[j + k]) {
          k++;
        }
        lcp[rk[i]] = k;
        if (k > 0) {
          k--;
        }
      }
    }
    return lcp;
  }

  size_t find(const string &needle) {
    if (needle.empty()) {
      return 0;
    }
    int lo = 0, hi = static_cast<int>(s.size()) - 1;
    while (lo <= hi) {
      int mid = lo + (hi - lo) / 2;
      int cmp = s.compare(sa[mid], needle.size(), needle);
      if (cmp < 0) {
        lo = mid + 1;
      } else if (cmp > 0) {
        hi = mid - 1;
      } else {
        return sa[mid];
      }
    }
    return string::npos;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  SuffixArrayManberMyers sa("banana");
  vector<int> sarr = sa.get_sa(), lcp = sa.get_lcp();
  vector<int> sarr_expected{5, 3, 1, 0, 4, 2};
  vector<int> lcp_expected{1, 3, 0, 0, 2};
  assert(sarr == sarr_expected);
  assert(lcp == lcp_expected);
  assert(sa.find("ana") == 1);
  assert(sa.find("x") == string::npos);
  return 0;
}
\end{lstlisting}
\subsection{Suffix Array and LCP (Counting Sort)}
Given a string \inlinecode{s}, a suffix array is the array of the smallest starting positions for the sorted suffixes of \inlinecode{s}. That is, the $i$-th position of the suffix array stores the starting position of the $i$-th lexicographically smallest suffix of \inlinecode{s}. For example, \inlinecode{s = "cab"} has the suffixes \inlinecode{"cab"}, \inlinecode{"ab"}, and \inlinecode{"b"}. When sorted, the indices of the suffixes are \inlinecode{"ab"}, \inlinecode{"b"}, and \inlinecode{"cab"}, so the suffix array (assuming 0-based indices) is $[1, 2, 0]$.

For a string \inlinecode{s} of length $n$, the longest common prefix (LCP) array of length $n - 1$ stores the lengths of the longest common prefixes between all pairs of lexicographically adjacent suffixes in \inlinecode{s}. For example, \inlinecode{"baa"} has the sorted suffixes \inlinecode{"a"}, \inlinecode{"aa"}, and \inlinecode{"baa"}, with an LCP array of \inlinecode{[1, 0]}.

The doubling algorithm sorts the suffixes by their first $2^k$ characters for increasing $k$: each round orders suffixes by their pair of ranks from the previous round, so a comparison costs $O(1)$ and the full order emerges after $O(\log n)$ rounds. Replacing the comparison sort of each round with counting-sort-style rank updates removes a logarithmic factor.

\begin{compactitem}
  \item \inlinecode{SuffixArrayCountingSort(s)} constructs a suffix array from the given string \inlinecode{s} using the original Manber-Myers doubling algorithm with counting-sort-style rank updates to reduce the running time to $O(n \log n)$.
  \item \inlinecode{get\_sa()} returns the constructed suffix array.
  \item \inlinecode{get\_lcp()} returns the corresponding LCP array for the suffix array.
  \item \inlinecode{find(needle)} returns one position that \inlinecode{needle} occurs in \inlinecode{s} (not necessarily the first), or \inlinecode{std::string::npos} if it cannot be found. For a \inlinecode{needle} of length $m$, this implementation uses an $O(m \log n)$ binary search, but can be optimized to $O(m + \log n)$ by first computing the LCP-LR array using the LCP array.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log n)$ per call to the constructor, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ per call to \inlinecode{get\_sa()}.
  \item $O(n)$ per call to \inlinecode{get\_lcp()}, where $n$ is the length of \inlinecode{s}.
  \item $O(m \log n)$ per call to \inlinecode{find(needle)}, where $m$ is the length of \inlinecode{needle} and $n$ is the length of \inlinecode{s}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary for storage of the suffix and LCP arrays, where $n$ is the length of \inlinecode{s}.
  \item $O(n)$ auxiliary heap space for the constructor.
  \item $O(1)$ auxiliary space for all other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <numeric>
#include <string>
#include <utility>
#include <vector>
using std::string;

class SuffixArrayCountingSort {
  string s;
  std::vector<int> sa, rk;

 public:
  explicit SuffixArrayCountingSort(const string &s) : s(s), sa(s.size()), rk(s.size()) {
    int n = static_cast<int>(s.size());
    std::iota(sa.rbegin(), sa.rend(), 0);
    for (int i = 0; i < n; i++) {
      rk[i] = static_cast<unsigned char>(s[i]);
    }
    std::stable_sort(sa.begin(), sa.end(), [&](int i, int j) {
      return static_cast<unsigned char>(s[i]) < static_cast<unsigned char>(s[j]);
    });
    for (int gap = 1; gap < n; gap *= 2) {
      std::vector<int> prev_rk(rk), prev_sa(sa), cnt(n);
      std::iota(cnt.begin(), cnt.end(), 0);
      for (int i = 0; i < n; i++) {
        rk[sa[i]] = (i > 0 && prev_rk[sa[i - 1]] == prev_rk[sa[i]] && sa[i - 1] + gap < n &&
                     prev_rk[sa[i - 1] + gap / 2] == prev_rk[sa[i] + gap / 2])
                        ? rk[sa[i - 1]]
                        : i;
      }
      for (int i = 0; i < n; i++) {
        int s1 = prev_sa[i] - gap;
        if (s1 >= 0) {
          sa[cnt[rk[s1]]++] = s1;
        }
      }
    }
  }

  const std::vector<int> &get_sa() const { return sa; }

  std::vector<int> get_lcp() const {
    int n = static_cast<int>(s.size());
    if (n == 0) {
      return {};  // Avoid constructing a vector of size (size_t)(-1).
    }
    std::vector<int> lcp(n - 1);
    for (int i = 0, k = 0; i < n; i++) {
      if (rk[i] < n - 1) {
        int j = sa[rk[i] + 1];
        while (std::max(i, j) + k < n && s[i + k] == s[j + k]) {
          k++;
        }
        lcp[rk[i]] = k;
        if (k > 0) {
          k--;
        }
      }
    }
    return lcp;
  }

  size_t find(const string &needle) {
    if (needle.empty()) {
      return 0;
    }
    int lo = 0, hi = static_cast<int>(s.size()) - 1;
    while (lo <= hi) {
      int mid = lo + (hi - lo) / 2;
      int cmp = s.compare(sa[mid], needle.size(), needle);
      if (cmp < 0) {
        lo = mid + 1;
      } else if (cmp > 0) {
        hi = mid - 1;
      } else {
        return sa[mid];
      }
    }
    return string::npos;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  SuffixArrayCountingSort sa("banana");
  vector<int> sarr = sa.get_sa(), lcp = sa.get_lcp();
  vector<int> sarr_expected{5, 3, 1, 0, 4, 2};
  vector<int> lcp_expected{1, 3, 0, 0, 2};
  assert(sarr == sarr_expected);
  assert(lcp == lcp_expected);
  assert(sa.find("ana") == 1);
  assert(sa.find("x") == string::npos);
  return 0;
}
\end{lstlisting}
\subsection{Suffix Array and LCP (DC3)}
Given a string \inlinecode{s}, a suffix array is the array of the smallest starting positions for the sorted suffixes of \inlinecode{s}. That is, the $i$-th position of the suffix array stores the starting position of the $i$-th lexicographically smallest suffix of \inlinecode{s}. For example, \inlinecode{s = "cab"} has the suffixes \inlinecode{"cab"}, \inlinecode{"ab"}, and \inlinecode{"b"}. When sorted, the indices of the suffixes are \inlinecode{"ab"}, \inlinecode{"b"}, and \inlinecode{"cab"}, so the suffix array (assuming 0-based indices) is $[1, 2, 0]$.

For a string \inlinecode{s} of length $n$, the longest common prefix (LCP) array of length $n - 1$ stores the lengths of the longest common prefixes between all pairs of lexicographically adjacent suffixes in \inlinecode{s}. For example, \inlinecode{"baa"} has the sorted suffixes \inlinecode{"a"}, \inlinecode{"aa"}, and \inlinecode{"baa"}, with an LCP array of $[1, 0]$.

The DC3/skew algorithm recursively constructs the suffix array of the suffixes starting at positions not divisible by 3, uses that order to sort the remaining suffixes with radix sort, and merges the two sorted groups in linear time.

\begin{compactitem}
  \item \inlinecode{SuffixArrayDC3(s)} constructs a suffix array from the given string \inlinecode{s} using the linear time DC3/skew algorithm by Karkkainen \& Sanders (2003) with radix sort.
  \item \inlinecode{get\_sa()} returns the constructed suffix array.
  \item \inlinecode{get\_lcp()} returns the corresponding LCP array for the suffix array.
  \item \inlinecode{find(needle)} returns one position that \inlinecode{needle} occurs in \inlinecode{s} (not necessarily the first), or \inlinecode{std::string::npos} if it cannot be found. For a \inlinecode{needle} of length $m$, this implementation uses an $O(m \log n)$ binary search, but can be optimized to $O(m + \log n)$ by first computing the LCP-LR array using the LCP array.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ per call to the constructor, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ per call to \inlinecode{get\_sa()}.
  \item $O(n)$ per call to \inlinecode{get\_lcp()}, where $n$ is the length of \inlinecode{s}.
  \item $O(m \log n)$ per call to \inlinecode{find(needle)}, where $m$ is the length of \inlinecode{needle} and $n$ is the length of \inlinecode{s}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ auxiliary for storage of the suffix and LCP arrays, where $n$ is the length of \inlinecode{s}.
  \item $O(n)$ auxiliary heap space for the constructor.
  \item $O(1)$ auxiliary space for all other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <string>
#include <utility>
#include <vector>
using std::string;

class SuffixArrayDC3 {
  static bool leq(int a1, int a2, int b1, int b2) { return (a1 < b1) || (a1 == b1 && a2 <= b2); }

  static bool leq(int a1, int a2, int a3, int b1, int b2, int b3) {
    return (a1 < b1) || (a1 == b1 && leq(a2, a3, b2, b3));
  }

  template<class It>
  static void radix_pass(It a, It b, It r, int n, int K) {
    std::vector<int> cnt(K + 1);
    for (int i = 0; i < n; i++) {
      cnt[r[a[i]]]++;
    }
    for (int i = 1; i <= K; i++) {
      cnt[i] += cnt[i - 1];
    }
    for (int i = n - 1; i >= 0; i--) {
      b[--cnt[r[a[i]]]] = a[i];
    }
  }

  template<class It>
  static void suffix_array_dc3(It s, It sa, int n, int K) {
    int n0 = (n + 2) / 3, n1 = (n + 1) / 3, n2 = n / 3, n02 = n0 + n2;
    std::vector<int> s12(n02 + 3), sa12(n02 + 3), s0(n0), sa0(n0);
    s12[n02] = s12[n02 + 1] = s12[n02 + 2] = 0;
    sa12[n02] = sa12[n02 + 1] = sa12[n02 + 2] = 0;
    for (int i = 0, j = 0; i < n + n0 - n1; i++) {
      if (i % 3 != 0) {
        s12[j++] = i;
      }
    }
    radix_pass(s12.begin(), sa12.begin(), s + 2, n02, K);
    radix_pass(sa12.begin(), s12.begin(), s + 1, n02, K);
    radix_pass(s12.begin(), sa12.begin(), s, n02, K);
    int name = 0, c0 = -1, c1 = -1, c2 = -1;
    for (int i = 0; i < n02; i++) {
      if (s[sa12[i]] != c0 || s[sa12[i] + 1] != c1 || s[sa12[i] + 2] != c2) {
        name++;
        c0 = s[sa12[i]];
        c1 = s[sa12[i] + 1];
        c2 = s[sa12[i] + 2];
      }
      (sa12[i] % 3 == 1 ? s12[sa12[i] / 3] : s12[sa12[i] / 3 + n0]) = name;
    }
    if (name < n02) {
      suffix_array_dc3(s12.begin(), sa12.begin(), n02, name);
      for (int i = 0; i < n02; i++) {
        s12[sa12[i]] = i + 1;
      }
    } else {
      for (int i = 0; i < n02; i++) {
        sa12[s12[i] - 1] = i;
      }
    }
    for (int i = 0, j = 0; i < n02; i++) {
      if (sa12[i] < n0) {
        s0[j++] = 3 * sa12[i];
      }
    }
    radix_pass(s0.begin(), sa0.begin(), s, n0, K);
    for (int p = 0, t = n0 - n1, k = 0; k < n; k++) {
      int i = (sa12[t] < n0) ? 3 * sa12[t] + 1 : 3 * (sa12[t] - n0) + 2, j = sa0[p];
      if (sa12[t] < n0
              ? leq(s[i], s12[sa12[t] + n0], s[j], s12[j / 3])
              : leq(s[i], s[i + 1], s12[sa12[t] - n0 + 1], s[j], s[j + 1], s12[j / 3 + n0])) {
        sa[k] = i;
        if (++t == n02) {
          for (k++; p < n0; p++, k++) {
            sa[k] = sa0[p];
          }
        }
      } else {
        sa[k] = j;
        if (++p == n0) {
          for (k++; t < n02; t++, k++) {
            sa[k] = (sa12[t] < n0) ? 3 * sa12[t] + 1 : 3 * (sa12[t] - n0) + 2;
          }
        }
      }
    }
  }

  string s;
  std::vector<int> sa;

 public:
  explicit SuffixArrayDC3(const string &s) : s(s), sa(s.size() + 1) {
    int n = static_cast<int>(s.size());
    std::vector<int> scopy(n);
    for (int i = 0; i < n; i++) {
      scopy[i] = static_cast<unsigned char>(s[i]) + 1;
    }
    scopy.resize(n + 4);
    suffix_array_dc3(scopy.begin(), sa.begin(), n + 1, 256);
    sa.erase(sa.begin());
  }

  const std::vector<int> &get_sa() const { return sa; }

  std::vector<int> get_lcp() const {
    int n = static_cast<int>(s.size());
    if (n == 0) {
      return {};
    }
    std::vector<int> rank(n), lcp(n - 1);
    for (int i = 0; i < n; i++) {
      rank[sa[i]] = i;
    }
    for (int i = 0, k = 0; i < n; i++) {
      if (rank[i] < n - 1) {
        int j = sa[rank[i] + 1];
        while (std::max(i, j) + k < n && s[i + k] == s[j + k]) {
          k++;
        }
        lcp[rank[i]] = k;
        if (k > 0) {
          k--;
        }
      }
    }
    return lcp;
  }

  size_t find(const string &needle) {
    if (needle.empty()) {
      return 0;
    }
    int lo = 0, hi = static_cast<int>(s.size()) - 1;
    while (lo <= hi) {
      int mid = lo + (hi - lo) / 2;
      int cmp = s.compare(sa[mid], needle.size(), needle);
      if (cmp < 0) {
        lo = mid + 1;
      } else if (cmp > 0) {
        hi = mid - 1;
      } else {
        return sa[mid];
      }
    }
    return string::npos;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  SuffixArrayDC3 sa("banana");
  vector<int> sarr = sa.get_sa(), lcp = sa.get_lcp();
  vector<int> sarr_expected{5, 3, 1, 0, 4, 2};
  vector<int> lcp_expected{1, 3, 0, 0, 2};
  assert(sarr == sarr_expected);
  assert(lcp == lcp_expected);
  assert(sa.find("ana") == 1);
  assert(sa.find("x") == string::npos);
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{String Data Structures}{3.5 String Data Structures}}
\setcounter{section}{5}
\setcounter{subsection}{0}
\subsection{Trie}
Maintain a map of strings to values using a tree data structure. Each node corresponds to a character, and each inserted string corresponds to a path from the root to a node that is flagged as a terminal node. Children are stored in hash tables, and \inlinecode{walk()} sorts child labels as needed to preserve lexicographic traversal.

\begin{compactitem}
  \item \inlinecode{Trie()} constructs an empty map.
  \item \inlinecode{size()} returns the size of the map.
  \item \inlinecode{empty()} returns whether the map is empty.
  \item \inlinecode{insert(s, v)} adds an entry with string key \inlinecode{s} and value \inlinecode{v} to the map, returning \inlinecode{true} if a new entry was added or \inlinecode{false} if the string already exists (in which case the map is unchanged and the old value associated with the string key is preserved).
  \item \inlinecode{erase(s)} removes the entry with string key \inlinecode{s} from the map, returning \inlinecode{true} if the removal was successful or \inlinecode{false} if the string to be removed was not found.
  \item \inlinecode{find(s)} returns a pointer to a const value associated with string key \inlinecode{s}, or \inlinecode{nullptr} if the key was not found.
  \item \inlinecode{count\_prefix(s)} returns the number of keys currently in the map that have \inlinecode{s} as a prefix (a key equal to \inlinecode{s} counts as well). \inlinecode{count\_prefix("")} therefore equals \inlinecode{size()}.
  \item \inlinecode{walk(f)} calls the function \inlinecode{f(s, v)} on each entry of the map, in lexicographically ascending order of the string keys.
\end{compactitem}

For a small, fixed alphabet, the \inlinecode{std::unordered\_map} children can be replaced by a fixed-size array of child pointers indexed by character. This removes the per-step hashing and lets \inlinecode{walk()} traverse in sorted order without sorting each node's labels, at the cost of restricting the alphabet and spending more memory per node.

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ expected per call to \inlinecode{insert(s, v)}, \inlinecode{erase(s)}, \inlinecode{find(s)}, and \inlinecode{count\_prefix(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(l)$ per call to \inlinecode{walk()}, where $l$ is the total length of string keys that are currently in the map, treating the character alphabet as constant.
  \item $O(1)$ per call to all other operations.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(l)$ for storage of the Trie, where $l$ is the total length of string keys that are currently in the map.
  \item $O(n)$ auxiliary stack space for construction, destruction, \inlinecode{walk()}, where $n$ is the maximum length of any string that has been inserted so far.
  \item $O(n)$ auxiliary stack space for \inlinecode{erase(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ auxiliary for all other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstddef>
#include <string>
#include <unordered_map>
#include <utility>
#include <vector>
using std::string;

template<class V>
class Trie {
  struct Node {
    V value;
    bool is_terminal;
    int cnt;  // Optional: maintain count of terminal keys in this subtree for count_prefix queries.
    std::unordered_map<char, Node *> children;

    Node() : is_terminal(false), cnt(0) {}
  } *root;

  static bool erase(Node *n, const string &s, int i) {
    if (i == static_cast<int>(s.size())) {
      if (!n->is_terminal) {
        return false;
      }
      n->is_terminal = false;
      n->cnt--;
      return true;
    }
    auto it = n->children.find(s[i]);
    if (it == n->children.end() || !erase(it->second, s, i + 1)) {
      return false;
    }
    n->cnt--;
    // Prune the child only if it is now a non-terminal leaf; a childless terminal is still a key.
    if (it->second->children.empty() && !it->second->is_terminal) {
      delete it->second;
      n->children.erase(it);
    }
    return true;
  }

  template<class Fn>
  static void walk(Node *n, string &s, Fn f) {
    if (n->is_terminal) {
      f(s, n->value);
    }
    std::vector<std::pair<char, Node *>> children(n->children.begin(), n->children.end());
    std::sort(children.begin(), children.end());
    for (auto &[c, child] : children) {
      s += c;
      walk(child, s, f);
      s.pop_back();
    }
  }

  static void clean_up(Node *n) {
    for (auto &[c, child] : n->children) {
      clean_up(child);
    }
    delete n;
  }

  int num_terminals;

 public:
  Trie() : root(new Node()), num_terminals(0) {}

  ~Trie() { clean_up(root); }
  Trie(const Trie &) = delete;
  Trie &operator=(const Trie &) = delete;
  int size() const { return num_terminals; }
  bool empty() const { return num_terminals == 0; }

  bool insert(const string &s, const V &v) {
    Node *n = root;
    for (char c : s) {
      auto [it, inserted] = n->children.try_emplace(c, nullptr);
      if (inserted) {
        it->second = new Node();
      }
      n = it->second;
    }
    if (n->is_terminal) {
      return false;
    }
    num_terminals++;
    n->is_terminal = true;
    n->value = v;
    // Re-walk from the root to bump the subtree count along the inserted path.
    n = root;
    n->cnt++;
    for (char c : s) {
      n = n->children[c];
      n->cnt++;
    }
    return true;
  }

  bool erase(const string &s) {
    if (erase(root, s, 0)) {
      num_terminals--;
      return true;
    }
    return false;
  }

  const V *find(const string &s) const {
    Node *n = root;
    for (char c : s) {
      if (auto it = n->children.find(c); it != n->children.end()) {
        n = it->second;
      } else {
        return nullptr;
      }
    }
    return n->is_terminal ? &(n->value) : nullptr;
  }

  int count_prefix(const string &s) const {
    Node *n = root;
    for (char c : s) {
      auto it = n->children.find(c);
      if (it == n->children.end()) {
        return 0;
      }
      n = it->second;
    }
    return n->cnt;
  }

  template<class Fn>
  void walk(Fn f) const {
    string s;
    walk(root, s, f);
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <iostream>
using namespace std;

int main() {
  vector<string> s{"", "a", "to", "tea", "ted", "ten", "i", "in", "inn"};
  Trie<int> t;
  assert(t.empty());
  for (int i = 0; i < static_cast<int>(s.size()); i++) {
    assert(t.insert(s[i], i));
  }
  t.walk([](string k, int v) { cout << "(\"" << k << "\", " << v << ")" << endl; });
  assert(!t.empty());
  assert(t.size() == 9);
  assert(!t.insert(s[0], 2));
  assert(t.size() == 9);
  assert(*t.find("") == 0);
  assert(*t.find("ten") == 5);
  assert(t.count_prefix("") == 9);    // Every key has the empty prefix.
  assert(t.count_prefix("te") == 3);  // "tea", "ted", "ten".
  assert(t.count_prefix("in") == 2);  // "in", "inn".
  assert(t.count_prefix("z") == 0);
  assert(t.erase("tea"));
  assert(t.size() == 8);
  assert(t.find("tea") == nullptr);
  assert(t.count_prefix("te") == 2);  // "ted", "ten" remain.
  assert(t.erase(""));
  assert(t.find("") == nullptr);
  assert(t.count_prefix("") == 7);

  // Erasing a key must not delete a shorter key sharing its path.
  Trie<int> u;
  u.insert("bc", 1);
  u.insert("bca", 2);
  assert(u.erase("bca"));
  assert(u.find("bc") != nullptr && *u.find("bc") == 1);
  assert(u.size() == 1 && u.count_prefix("bc") == 1);
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
("", 0)
("a", 1)
("i", 6)
("in", 7)
("inn", 8)
("tea", 3)
("ted", 4)
("ten", 5)
("to", 2)
\end{exampleoutput}
\subsection{Radix Tree}
Maintain a map of strings to values using a compressed tree data structure. Each node corresponds to a substring of an inserted string, and each inserted string corresponds to a path from the root to a node that is flagged as a terminal node. Contrary to a regular trie, a radix tree is more space efficient as it combines chains of nodes with only a single child. Children are stored in hash tables, and \inlinecode{walk()} sorts child labels as needed to preserve lexicographic traversal.

\begin{compactitem}
  \item \inlinecode{RadixTree()} constructs an empty map.
  \item \inlinecode{size()} returns the size of the map.
  \item \inlinecode{empty()} returns whether the map is empty.
  \item \inlinecode{insert(s, v)} adds an entry with string key \inlinecode{s} and value \inlinecode{v} to the map, returning \inlinecode{true} if a new entry was added or \inlinecode{false} if the string already exists (in which case the map is unchanged and the old value associated with the string key is preserved).
  \item \inlinecode{erase(s)} removes the entry with string key \inlinecode{s} from the map, returning \inlinecode{true} if the removal was successful or \inlinecode{false} if the string to be removed was not found.
  \item \inlinecode{find(s)} returns a pointer to a const value associated with string key \inlinecode{s}, or \inlinecode{nullptr} if the key was not found.
  \item \inlinecode{count\_prefix(s)} returns the number of keys currently in the map that have \inlinecode{s} as a prefix (a key equal to \inlinecode{s} counts as well), including the case where \inlinecode{s} ends partway along a compressed edge. \inlinecode{count\_prefix("")} therefore equals \inlinecode{size()}.
  \item \inlinecode{walk(f)} calls the function \inlinecode{f(s, v)} on each entry of the map, in lexicographically ascending order of the string keys.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ expected per call to \inlinecode{insert(s, v)}, \inlinecode{erase(s)}, \inlinecode{find(s)}, and \inlinecode{count\_prefix(s)}, where $n$ is the length of \inlinecode{s}, assuming the number of outgoing compressed edges checked at each node is small.
  \item $O(l \log l)$ per call to \inlinecode{walk()}, where $l$ is the total length of string keys that are currently in the map.
  \item $O(1)$ per call to all other operations.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(l)$ for storage of the radix tree, where $l$ is the total length of string keys that are currently in the map.
  \item $O(n)$ auxiliary stack space for construction, destruction, \inlinecode{walk()}, where $n$ is the maximum length of any string that has been inserted so far.
  \item $O(n)$ auxiliary stack space for \inlinecode{erase(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(1)$ auxiliary for all other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cstddef>
#include <string>
#include <unordered_map>
#include <utility>
#include <vector>
using std::string;

template<class V>
class RadixTree {
  struct Node {
    V value;
    bool is_terminal;
    int cnt;  // Optional: maintain count of terminal keys in this subtree for count_prefix queries.
    std::unordered_map<string, Node *> children;

    Node(const V &value = V(), bool is_terminal = false)
        : value(value), is_terminal(is_terminal), cnt(0) {}
  } *root;

  static int lcp_len(const string &s1, const string &s2, int s2start) {
    int i = 0;
    for (int j = s2start; i < static_cast<int>(s1.size()) && j < static_cast<int>(s2.size());
         i++, j++) {
      if (s1[i] != s2[j]) {
        break;
      }
    }
    return i;
  }

  static bool insert(Node *n, const string &s, int i, const V &v) {
    if (i == static_cast<int>(s.size())) {
      if (n->is_terminal) {
        return false;
      }
      n->value = v;
      n->is_terminal = true;
      n->cnt++;
      return true;
    }
    for (auto it = n->children.begin(); it != n->children.end(); ++it) {
      int len = lcp_len(it->first, s, i);
      if (len == 0) {
        continue;
      }
      if (len == static_cast<int>(it->first.size())) {
        if (!insert(it->second, s, i + len, v)) {
          return false;
        }
        n->cnt++;
        return true;
      }
      string left = it->first.substr(0, len);
      string right = it->first.substr(len);
      Node *tmp = new Node();
      tmp->cnt = it->second->cnt;  // The new split node heads the same subtree as the old child.
      tmp->children[right] = it->second;
      n->children.erase(it);
      n->children[left] = tmp;
      if (len == static_cast<int>(s.size()) - i) {
        tmp->value = v;
        tmp->is_terminal = true;
        tmp->cnt++;
      } else {
        insert(tmp, s, i + len, v);
      }
      n->cnt++;
      return true;
    }
    Node *leaf = new Node(v, true);
    leaf->cnt = 1;
    n->children[s.substr(i)] = leaf;
    n->cnt++;
    return true;
  }

  static bool erase(Node *n, const string &s, int i) {
    if (i == static_cast<int>(s.size())) {
      if (!n->is_terminal) {
        return false;
      }
      n->is_terminal = false;
      n->cnt--;
      return true;
    }
    for (auto it = n->children.begin(); it != n->children.end(); ++it) {
      int len = lcp_len(it->first, s, i);
      if (len == 0) {
        continue;
      }
      // The entire edge label must be consumed; a partial match means the key is not present, so
      // erasing must fail rather than descend (otherwise erase("te") would remove "tea").
      if (len < static_cast<int>(it->first.size())) {
        return false;
      }
      Node *child = it->second;
      if (!erase(child, s, i + len)) {
        return false;
      }
      n->cnt--;
      // The merge below leaves counts intact: a non-terminal node with one child already has the
      // same subtree count as that child, so reusing the node keeps its `cnt` correct.
      if (child->children.empty() && !child->is_terminal) {
        delete child;
        n->children.erase(it);
      } else if (child->children.size() == 1) {
        Node *grandchild = child->children.begin()->second;
        if (!child->is_terminal) {
          string merged_key(it->first + child->children.begin()->first);
          child->value = grandchild->value;
          child->is_terminal = grandchild->is_terminal;
          child->children = grandchild->children;
          delete grandchild;
          n->children.erase(it);
          n->children[merged_key] = child;
        }
      }
      return true;
    }
    return false;
  }

  template<class Fn>
  static void walk(Node *n, string &s, Fn f) {
    if (n->is_terminal) {
      f(s, n->value);
    }
    std::vector<std::pair<string, Node *>> children(n->children.begin(), n->children.end());
    std::sort(children.begin(), children.end());
    for (auto &[key, child] : children) {
      s += key;
      walk(child, s, f);
      s.erase(s.size() - key.size());
    }
  }

  static void clean_up(Node *n) {
    for (auto &[key, child] : n->children) {
      clean_up(child);
    }
    delete n;
  }

  int num_terminals;

 public:
  RadixTree() : root(new Node()), num_terminals(0) {}

  ~RadixTree() { clean_up(root); }
  RadixTree(const RadixTree &) = delete;
  RadixTree &operator=(const RadixTree &) = delete;
  int size() const { return num_terminals; }
  bool empty() const { return num_terminals == 0; }

  bool insert(const string &s, const V &v) {
    if (insert(root, s, 0, v)) {
      num_terminals++;
      return true;
    }
    return false;
  }

  bool erase(const string &s) {
    if (erase(root, s, 0)) {
      num_terminals--;
      return true;
    }
    return false;
  }

  const V *find(const string &s) const {
    Node *n = root;
    int i = 0;
    while (i < static_cast<int>(s.size())) {
      bool found = false;
      for (auto &[key, child] : n->children) {
        if (key[0] == s[i]) {
          // The entire edge label must be consumed; a partial match (s ends mid-edge or diverges
          // from it) means the key is not present.
          if (lcp_len(key, s, i) < static_cast<int>(key.size())) {
            return nullptr;
          }
          i += static_cast<int>(key.size());
          n = child;
          found = true;
          break;
        }
      }
      if (!found) {
        return nullptr;
      }
    }
    return n->is_terminal ? &(n->value) : nullptr;
  }

  int count_prefix(const string &s) const {
    Node *n = root;
    int i = 0;
    while (i < static_cast<int>(s.size())) {
      bool found = false;
      for (auto &[key, child] : n->children) {
        if (key[0] == s[i]) {
          int len = lcp_len(key, s, i);
          if (i + len == static_cast<int>(s.size())) {
            // `s` ends within (or exactly at the end of) this edge: every key below qualifies.
            return child->cnt;
          }
          if (len < static_cast<int>(key.size())) {
            return 0;  // `s` diverges partway along the edge.
          }
          i += len;
          n = child;
          found = true;
          break;
        }
      }
      if (!found) {
        return 0;
      }
    }
    return n->cnt;
  }

  template<class Fn>
  void walk(Fn f) const {
    string s;
    walk(root, s, f);
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <iostream>
using namespace std;

void print_entry(const string &k, int v) {
  cout << "(\"" << k << "\", " << v << ")" << endl;
}

int main() {
  vector<string> s{"", "a", "to", "tea", "ted", "ten", "i", "in", "inn"};
  RadixTree<int> t;
  assert(t.empty());
  for (int i = 0; i < static_cast<int>(s.size()); i++) {
    assert(t.insert(s[i], i));
  }
  t.walk([](string k, int v) { cout << "(\"" << k << "\", " << v << ")" << endl; });
  assert(!t.empty());
  assert(t.size() == 9);
  assert(!t.insert(s[0], 2));
  assert(t.size() == 9);
  assert(t.find("") && *t.find("") == 0);
  assert(*t.find("ten") == 5);
  assert(t.count_prefix("") == 9);    // Every key has the empty prefix.
  assert(t.count_prefix("te") == 3);  // "tea", "ted", "ten" (prefix ends mid-edge).
  assert(t.count_prefix("in") == 2);  // "in", "inn".
  assert(t.count_prefix("z") == 0);
  assert(t.erase("tea"));
  assert(t.size() == 8);
  assert(t.find("tea") == nullptr);
  assert(t.count_prefix("te") == 2);  // "ted", "ten" remain.
  assert(t.erase(""));
  assert(t.find("") == nullptr);
  assert(t.count_prefix("") == 7);

  // Erasing a key must not delete a shorter key sharing its path.
  RadixTree<int> u;
  u.insert("bc", 1);
  u.insert("bca", 2);
  assert(u.erase("bca"));
  assert(u.find("bc") != nullptr && *u.find("bc") == 1);
  assert(u.size() == 1 && u.count_prefix("bc") == 1);
  return 0;
}
\end{lstlisting}
\begin{exampleoutput}
("", 0)
("a", 1)
("i", 6)
("in", 7)
("inn", 8)
("tea", 3)
("ted", 4)
("ten", 5)
("to", 2)
\end{exampleoutput}
\subsection{Suffix Tree (Ukkonen)}
A suffix tree is the compressed trie of every suffix of a string: a rooted tree in which each edge is labeled with a substring, no two edges leaving a node start with the same character, and the concatenation of labels from the root to each leaf spells one suffix. Paths from the root therefore spell exactly the distinct substrings of the string, which makes the tree a powerful index for substring queries, repeated-substring problems, and many other tasks. It is the explicit-pointer counterpart of the suffix automaton; the two store the same information and either can be converted into the other.

This implementation is Ukkonen's algorithm, which builds the tree online in linear time by feeding one character at a time. It maintains an ``active point'' (a node, an edge, and a length along that edge) describing where the next suffix must be inserted, together with suffix links that let it hop between successive insertion points in amortized constant time. Edges to leaves are stored with an open-ended length that grows automatically as the string lengthens, so each phase touches only the suffixes that genuinely change. A unique terminal sentinel smaller than every real character is appended internally so that no suffix is a prefix of another and every suffix ends at its own leaf. The input must therefore not contain the character \inlinecode{\textquotesingle{}\textbackslash{}1\textquotesingle{}}.

\begin{compactitem}
  \item \inlinecode{SuffixTree(s)} builds the suffix tree of string \inlinecode{s}.
  \item \inlinecode{contains(t)} returns whether string \inlinecode{t} occurs as a substring of \inlinecode{s}.
  \item \inlinecode{count\_distinct\_substrings()} returns the number of distinct nonempty substrings of \inlinecode{s}.
  \item \inlinecode{longest\_repeated\_substring()} returns one longest substring of \inlinecode{s} that occurs at least twice, or the empty string if every character is distinct.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ to construct the tree for a string of length $n$ (using a per-node ordered map, $O(n \log s)$ where $s$ is the alphabet size).
  \item $O(m \log s)$ per call to \inlinecode{contains(t)}, where $m$ is the length of \inlinecode{t}.
  \item $O(n)$ per call to \inlinecode{count\_distinct\_substrings()} and \inlinecode{longest\_repeated\_substring()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ nodes and edges.
  \item $O(n)$ auxiliary stack space for the traversals.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <map>
#include <string>
#include <vector>

class SuffixTree {
  static const int OPEN = -1;  // Marks a leaf edge whose right end follows the growing string.

  struct Node {
    int start, end;  // Edge into this node labels s[start..end]; end == OPEN means leaf_end.
    int link;
    std::map<char, int> next;
    Node(int start, int end) : start(start), end(end), link(0) {}
  };

  std::string s;
  std::vector<Node> tree;
  int leaf_end;
  // Active point and bookkeeping for Ukkonen's algorithm.
  int active_node, active_edge, active_length, remaining, last_new;

  int edge_end(int v) const { return tree[v].end == OPEN ? leaf_end : tree[v].end; }
  int edge_len(int v) const { return edge_end(v) - tree[v].start + 1; }

  int make_node(int start, int end) {
    tree.emplace_back(start, end);
    return static_cast<int>(tree.size()) - 1;
  }

  // If the active length runs past the edge into v, descend and report that we moved.
  bool walk_down(int v) {
    if (active_length >= edge_len(v)) {
      active_edge += edge_len(v);
      active_length -= edge_len(v);
      active_node = v;
      return true;
    }
    return false;
  }

  void extend(int pos) {
    leaf_end = pos;
    remaining++;
    last_new = -1;
    while (remaining > 0) {
      if (active_length == 0) {
        active_edge = pos;
      }
      char c = s[active_edge];
      auto it = tree[active_node].next.find(c);
      if (it == tree[active_node].next.end()) {
        // Rule 2: no edge starts with c, so grow a new leaf from the active node.
        tree[active_node].next[c] = make_node(pos, OPEN);
        if (last_new != -1) {
          tree[last_new].link = active_node;
          last_new = -1;
        }
      } else {
        int v = it->second;
        if (walk_down(v)) {
          continue;  // Re-evaluate the same suffix from the deeper node.
        }
        if (s[tree[v].start + active_length] == s[pos]) {
          // Rule 3: the character is already present; stop and lengthen the active point.
          if (last_new != -1) {
            tree[last_new].link = active_node;
          }
          active_length++;
          break;
        }
        // Rule 2 with a split: break edge v in two and attach a new leaf at the split point.
        int split = make_node(tree[v].start, tree[v].start + active_length - 1);
        tree[active_node].next[c] = split;
        tree[split].next[s[pos]] = make_node(pos, OPEN);
        tree[v].start += active_length;
        tree[split].next[s[tree[v].start]] = v;
        if (last_new != -1) {
          tree[last_new].link = split;
        }
        last_new = split;
      }
      remaining--;
      if (active_node == 0 && active_length > 0) {
        active_length--;
        active_edge = pos - remaining + 1;
      } else if (active_node != 0) {
        active_node = tree[active_node].link;
      }
    }
  }

  // String depth and right end (one past) of the deepest internal node, for repeated substrings.
  void deepest_internal(int v, int depth, int &best_depth, int &best_end) const {
    bool internal = false;
    for (const auto &kv : tree[v].next) {
      internal = true;
      int w = kv.second;
      deepest_internal(w, depth + edge_len(w), best_depth, best_end);
    }
    if (internal && v != 0 && depth > best_depth) {
      best_depth = depth;
      best_end = edge_end(v) + 1;
    }
  }

 public:
  explicit SuffixTree(const std::string &str) : s(str) {
    s.push_back('\1');  // Unique terminal sentinel, smaller than every real character.
    make_node(-1, -1);  // Root, at index 0; its incoming edge is never read.
    active_node = 0;
    active_edge = 0;
    active_length = 0;
    remaining = 0;
    leaf_end = -1;
    last_new = -1;
    for (int i = 0; i < static_cast<int>(s.size()); i++) {
      extend(i);
    }
  }

  bool contains(const std::string &t) const {
    int v = 0, i = 0, m = static_cast<int>(t.size());
    while (i < m) {
      auto it = tree[v].next.find(t[i]);
      if (it == tree[v].next.end()) {
        return false;
      }
      v = it->second;
      int len = edge_len(v);
      for (int j = 0; j < len && i < m; j++, i++) {
        if (s[tree[v].start + j] != t[i]) {
          return false;
        }
      }
    }
    return true;
  }

  int count_distinct_substrings() const {
    int total = 0;
    for (int v = 1; v < static_cast<int>(tree.size()); v++) {
      total += edge_len(v);
    }
    // Every leaf edge ends in the sentinel; the s.size() suffixes of the sentineled string are the
    // only substrings that contain it, so removing them leaves the count for the original string.
    return total - static_cast<int>(s.size());
  }

  std::string longest_repeated_substring() const {
    int best_depth = 0, best_end = 0;
    deepest_internal(0, 0, best_depth, best_end);
    return s.substr(best_end - best_depth, best_depth);
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <set>
using namespace std;

int main() {
  SuffixTree st("banana");
  assert(st.contains("ana"));
  assert(st.contains("banana"));
  assert(st.contains(""));
  assert(!st.contains("anana_"));
  assert(!st.contains("nb"));
  assert(st.longest_repeated_substring() == "ana");

  // "banana" has 15 distinct substrings:
  // a, an, ana, anan, anana, b, ba, ban, bana, banan, banana, n, na, nan, nana.
  assert(st.count_distinct_substrings() == 15);

  // Cross-check both queries against brute force on a small string.
  string text = "mississippi";
  SuffixTree t(text);
  set<string> subs;
  int n = static_cast<int>(text.size());
  for (int i = 0; i < n; i++) {
    for (int len = 1; i + len <= n; len++) {
      subs.insert(text.substr(i, len));
    }
  }
  assert(t.count_distinct_substrings() == static_cast<int>(subs.size()));
  for (const string &sub : subs) {
    assert(t.contains(sub));
  }
  assert(!t.contains("issa"));
  assert(t.longest_repeated_substring() == "issi");  // "issi" appears at indices 1 and 4.
  return 0;
}
\end{lstlisting}
\subsection{Suffix Automaton}
Maintains a compact deterministic automaton representing all substrings of a string. A suffix automaton is useful for substring membership queries, counting distinct substrings, finding longest common substrings, and many other string dynamic programming tasks.

Each state represents an equivalence class of substrings with the same set of ending positions. The value \inlinecode{st[v].len} is the maximum length represented by state \inlinecode{v}, while \inlinecode{st[v].link} points to the state representing the longest proper suffix of that class.

\begin{compactitem}
  \item \inlinecode{SuffixAutomaton()} constructs an empty automaton.
  \item \inlinecode{SuffixAutomaton(s)} constructs the automaton for string \inlinecode{s}.
  \item \inlinecode{extend(c)} appends character \inlinecode{c} to the current string.
  \item \inlinecode{contains(t)} returns whether string \inlinecode{t} occurs as a substring.
  \item \inlinecode{first\_occurrence(t)} returns the starting index of the first occurrence of \inlinecode{t}, or $-1$ if absent.
  \item \inlinecode{count\_distinct\_substrings()} returns the number of distinct nonempty substrings.
  \item \inlinecode{longest\_common\_substring(t)} returns one longest substring common to the built string and string \inlinecode{t}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ expected to construct the automaton for a string of length $n$.
  \item $O(m)$ expected per call to \inlinecode{contains(t)}, \inlinecode{first\_occurrence(t)}, or \inlinecode{longest\_common\_substring(t)}, where $m$ is the length of \inlinecode{t}.
  \item $O(n)$ per call to \inlinecode{count\_distinct\_substrings()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ transition storage and $O(n)$ states.
  \item $O(1)$ auxiliary per character processed.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cstdint>
#include <string>
#include <unordered_map>
#include <vector>
using std::string;

class SuffixAutomaton {
  struct State {
    int len, link, first_pos;
    std::unordered_map<char, int> next;

    State() : len(0), link(-1), first_pos(-1) {}
  };

  std::vector<State> st;
  int last;

 public:
  SuffixAutomaton() : st(1), last(0) {}

  explicit SuffixAutomaton(const string &s) : st(1), last(0) {
    for (char c : s) {
      extend(c);
    }
  }

  void extend(char c) {
    int cur = static_cast<int>(st.size());
    st.emplace_back();
    st[cur].len = st[last].len + 1;
    st[cur].first_pos = st[cur].len - 1;
    int p = last;
    while (p != -1 && st[p].next.find(c) == st[p].next.end()) {
      st[p].next[c] = cur;
      p = st[p].link;
    }
    if (p == -1) {
      st[cur].link = 0;
    } else {
      int q = st[p].next[c];
      if (st[p].len + 1 == st[q].len) {
        st[cur].link = q;
      } else {
        int clone = static_cast<int>(st.size());
        st.push_back(st[q]);
        st[clone].len = st[p].len + 1;
        while (p != -1 && st[p].next[c] == q) {
          st[p].next[c] = clone;
          p = st[p].link;
        }
        st[q].link = st[cur].link = clone;
      }
    }
    last = cur;
  }

  bool contains(const string &t) const {
    int v = 0;
    for (char c : t) {
      if (auto it = st[v].next.find(c); it != st[v].next.end()) {
        v = it->second;
      } else {
        return false;
      }
    }
    return true;
  }

  // Returns the starting index of the first (leftmost) occurrence of `t` as a substring, or -1 if
  // `t` does not occur. `first_pos` stores the end index of the first occurrence of each state's
  // class, so subtracting the length of `t` recovers its start.
  int first_occurrence(const string &t) const {
    int v = 0;
    for (char c : t) {
      auto it = st[v].next.find(c);
      if (it == st[v].next.end()) {
        return -1;
      }
      v = it->second;
    }
    return st[v].first_pos - static_cast<int>(t.size()) + 1;
  }

  int64_t count_distinct_substrings() const {
    int64_t res = 0;
    for (int v = 1; v < static_cast<int>(st.size()); v++) {
      res += st[v].len - st[st[v].link].len;
    }
    return res;
  }

  string longest_common_substring(const string &t) const {
    int v = 0, len = 0, best = 0, best_pos = -1;
    for (int i = 0; i < static_cast<int>(t.size()); i++) {
      while (v && !st[v].next.count(t[i])) {
        v = st[v].link;
        len = st[v].len;
      }
      if (auto it = st[v].next.find(t[i]); it != st[v].next.end()) {
        v = it->second;
        len++;
      } else {
        v = 0;
        len = 0;
      }
      if (len > best) {
        best = len;
        best_pos = i;
      }
    }
    return best == 0 ? "" : t.substr(best_pos - best + 1, best);
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  SuffixAutomaton sa("ababa");
  assert(sa.contains("aba"));
  assert(sa.contains("bab"));
  assert(!sa.contains("abba"));
  assert(sa.count_distinct_substrings() == 9);
  assert(sa.longest_common_substring("zzbabab") == "baba");
  assert(sa.first_occurrence("aba") == 0);
  assert(sa.first_occurrence("bab") == 1);
  assert(sa.first_occurrence("abba") == -1);

  SuffixAutomaton online;
  online.extend('a');
  online.extend('b');
  online.extend('c');
  assert(online.contains("bc"));
  return 0;
}
\end{lstlisting}
\subsection{Eertree}
Maintains all distinct palindromic substrings of a string online using an Eertree, also known as a palindromic tree. Each non-root node represents one distinct palindrome, and suffix links connect each palindrome to its longest proper palindromic suffix.

The structure has two roots: one of length $-1$, which simplifies boundary cases, and one of length $0$, representing the empty palindrome. After each appended character, \inlinecode{last} points to the node for the longest palindromic suffix of the current string.

\begin{compactitem}
  \item \inlinecode{Eertree()} constructs an empty palindromic tree.
  \item \inlinecode{Eertree(s)} constructs the tree for string \inlinecode{s}.
  \item \inlinecode{add(c)} appends character \inlinecode{c} and returns \inlinecode{true} if this creates a new distinct palindrome.
  \item \inlinecode{count\_distinct\_palindromes()} returns the number of distinct nonempty palindromic substrings.
  \item \inlinecode{longest\_suffix\_length()} returns the length of the current longest palindromic suffix.
  \item \inlinecode{count\_occurrences()} propagates occurrence counts through suffix links and returns \inlinecode{occ[v]} for each node \inlinecode{v}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n)$ expected to build the tree for a string of length $n$.
  \item $O(1)$ expected amortized per call to \inlinecode{add(c)}.
  \item $O(n)$ per call to \inlinecode{count\_occurrences()}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n)$ transition storage and $O(n)$ nodes.
  \item $O(1)$ auxiliary per appended character.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <string>
#include <unordered_map>
#include <vector>
using std::string;

class Eertree {
  struct Node {
    int len, link, occ;
    std::unordered_map<char, int> next;

    Node(int len = 0) : len(len), link(0), occ(0) {}
  };

  std::vector<Node> tree;
  string s;
  int last;

  int get_suffix(int v, int pos, char c) const {
    while (pos - 1 - tree[v].len < 0 || s[pos - 1 - tree[v].len] != c) {
      v = tree[v].link;
    }
    return v;
  }

 public:
  Eertree() : tree(), s(), last(1) {
    tree.emplace_back(-1);
    tree.emplace_back(0);
    tree[0].link = 0;
    tree[1].link = 0;
  }

  explicit Eertree(const string &text) : tree(), s(), last(1) {
    tree.emplace_back(-1);
    tree.emplace_back(0);
    tree[0].link = 0;
    tree[1].link = 0;
    for (char c : text) {
      add(c);
    }
  }

  bool add(char c) {
    s += c;
    int pos = static_cast<int>(s.size()) - 1;
    int cur = get_suffix(last, pos, c);
    auto it = tree[cur].next.find(c);
    if (it != tree[cur].next.end()) {
      last = it->second;
      tree[last].occ++;
      return false;
    }
    last = static_cast<int>(tree.size());
    tree.emplace_back(tree[cur].len + 2);
    tree[last].occ = 1;
    tree[cur].next[c] = last;
    if (tree[last].len == 1) {
      tree[last].link = 1;
      return true;
    }
    int link_parent = get_suffix(tree[cur].link, pos, c);
    tree[last].link = tree[link_parent].next[c];
    return true;
  }

  int count_distinct_palindromes() const { return static_cast<int>(tree.size()) - 2; }
  int longest_suffix_length() const { return tree[last].len; }

  std::vector<int> count_occurrences() const {
    std::vector<int> occ(tree.size());
    for (int i = 0; i < static_cast<int>(tree.size()); i++) {
      occ[i] = tree[i].occ;
    }
    for (int i = static_cast<int>(tree.size()) - 1; i >= 2; i--) {
      occ[tree[i].link] += occ[i];
    }
    return occ;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  Eertree t("abacaba");
  assert(t.count_distinct_palindromes() == 7);
  assert(t.longest_suffix_length() == 7);

  Eertree online;
  assert(online.add('a'));
  assert(online.add('a'));
  assert(online.add('b'));
  assert(online.add('a'));
  assert(online.count_distinct_palindromes() == 4);
  assert(online.longest_suffix_length() == 3);

  Eertree duplicate;
  assert(duplicate.add('a'));
  assert(duplicate.add('b'));
  assert(duplicate.add('c'));
  assert(!duplicate.add('a'));

  std::vector<int> occ = t.count_occurrences();
  assert(static_cast<int>(occ.size()) == t.count_distinct_palindromes() + 2);
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Sequence Dynamic Programming}{3.6 Sequence Dynamic Programming}}
\setcounter{section}{6}
\setcounter{subsection}{0}
\subsection{Longest Common Substring}
Given two strings, determine their longest common substring: a contiguous run of characters that appears in both. A dynamic program computes, for each pair of prefixes, the length of their longest common suffix; the largest value over all pairs locates a longest common substring. Only the previous row of the table is retained, so the auxiliary space is linear in the shorter string.

\begin{compactitem}
  \item \inlinecode{longest\_common\_substring(s1, s2)} returns a longest common substring of \inlinecode{s1} and \inlinecode{s2} (one such substring if several are tied in length), or the empty string if they have none in common.
\end{compactitem}

\textbf{Time Complexity}: $O(n \cdot m)$ per call to \inlinecode{longest\_common\_substring(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.

\textbf{Space Complexity}: $O(\min\left(n, m\right))$ auxiliary heap space, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.

\begin{lstlisting}[language=C++]
#include <string>
#include <vector>
using std::string;

string longest_common_substring(const string &s1, const string &s2) {
  int n = static_cast<int>(s1.size()), m = static_cast<int>(s2.size());
  if (n == 0 || m == 0) {
    return "";
  }
  if (n < m) {
    return longest_common_substring(s2, s1);
  }
  std::vector<int> curr(m), prev(m);
  int pos = 0, len = 0;
  for (int i = 0; i < n; i++) {
    for (int j = 0; j < m; j++) {
      if (s1[i] == s2[j]) {
        curr[j] = (i > 0 && j > 0) ? prev[j - 1] + 1 : 1;
        if (len < curr[j]) {
          len = curr[j];
          pos = i - curr[j] + 1;
        }
      } else {
        curr[j] = 0;
      }
    }
    curr.swap(prev);
  }
  return s1.substr(pos, len);
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(longest_common_substring("bbbabca", "aababcd") == "babc");
  return 0;
}
\end{lstlisting}
\subsection{Longest Common Subsequence}
Given two strings, determine their longest common subsequence. A subsequence is a string that can be derived from the original string by deleting some elements without changing the order of the remaining elements (e.g. \inlinecode{"ACE"} is a subsequence of \inlinecode{"ABCDE"}, but \inlinecode{"BAE"} is not).

\begin{compactitem}
  \item \inlinecode{longest\_common\_subsequence(s1, s2)} returns the longest common subsequence of strings \inlinecode{s1} and \inlinecode{s2} using a classic dynamic programming approach. This implementation computes \inlinecode{dp[i][j]} (the length of the longest common subsequence for the length $i$ prefix of \inlinecode{s1} and the length $j$ prefix of \inlinecode{s2}) before following the path backwards to construct the answer.
  \item \inlinecode{hirschberg\_lcs(s1, s2)} returns the longest common subsequence of strings \inlinecode{s1} and \inlinecode{s2} using the more memory efficient Hirschberg's algorithm.
\end{compactitem}

\textbf{Time Complexity}: $O(n \cdot m)$ per call to \inlinecode{longest\_common\_subsequence(s1, s2)} as well as \inlinecode{hirschberg\_lcs(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n \cdot m)$ auxiliary heap space for \inlinecode{longest\_common\_subsequence(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.
  \item $O(\log\left(\max\left(n, m\right)\right))$ auxiliary stack space and $O(\min\left(n, m\right))$ auxiliary heap space for \inlinecode{hirschberg\_lcs(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <iterator>
#include <string>
#include <vector>
using std::string;

string longest_common_subsequence(const string &s1, const string &s2) {
  int n = static_cast<int>(s1.size()), m = static_cast<int>(s2.size());
  std::vector<std::vector<int>> dp(n + 1, std::vector<int>(m + 1, 0));
  for (int i = 1; i <= n; i++) {
    for (int j = 1; j <= m; j++) {
      if (s1[i - 1] == s2[j - 1]) {
        dp[i][j] = dp[i - 1][j - 1] + 1;
      } else {
        dp[i][j] = std::max(dp[i][j - 1], dp[i - 1][j]);
      }
    }
  }
  string res;
  for (int i = n, j = m; i > 0 && j > 0;) {
    if (s1[i - 1] == s2[j - 1]) {
      res += s1[i - 1];
      i--;
      j--;
    } else if (dp[i - 1][j] >= dp[i][j - 1]) {
      i--;
    } else {
      j--;
    }
  }
  std::reverse(res.begin(), res.end());
  return res;
}

template<class It>
std::vector<int> lcs_len(It lo1, It hi1, It lo2, It hi2) {
  std::vector<int> res(std::distance(lo2, hi2) + 1), prev(res);
  for (It it1 = lo1; it1 != hi1; ++it1) {
    res.swap(prev);
    int i = 0;
    for (It it2 = lo2; it2 != hi2; ++it2) {
      res[i + 1] = (*it1 == *it2) ? prev[i] + 1 : std::max(res[i], prev[i + 1]);
      i++;
    }
  }
  return res;
}

template<class It>
void hirschberg_rec(It lo1, It hi1, It lo2, It hi2, string *res) {
  if (lo1 == hi1) {
    return;
  }
  if (lo1 + 1 == hi1) {
    if (std::find(lo2, hi2, *lo1) != hi2) {
      *res += *lo1;
    }
    return;
  }
  It mid1 = lo1 + (hi1 - lo1) / 2;
  std::reverse_iterator<It> rlo1(hi1), rmid1(mid1), rlo2(hi2), rhi2(lo2);
  std::vector<int> fwd = lcs_len(lo1, mid1, lo2, hi2);
  std::vector<int> rev = lcs_len(rlo1, rmid1, rlo2, rhi2);
  It mid2 = lo2;
  int maxlen = -1;
  for (int i = 0, j = static_cast<int>(rev.size()) - 1; i < static_cast<int>(fwd.size());
       i++, j--) {
    if (fwd[i] + rev[j] > maxlen) {
      maxlen = fwd[i] + rev[j];
      mid2 = lo2 + i;
    }
  }
  hirschberg_rec(lo1, mid1, lo2, mid2, res);
  hirschberg_rec(mid1, hi1, mid2, hi2, res);
}

string hirschberg_lcs(const string &s1, const string &s2) {
  if (s1.size() < s2.size()) {
    return hirschberg_lcs(s2, s1);
  }
  string res;
  hirschberg_rec(s1.begin(), s1.end(), s2.begin(), s2.end(), &res);
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(longest_common_subsequence("xmjyauz", "mzjawxu") == "mjau");
  assert(hirschberg_lcs("xmjyauz", "mzjawxu") == "mjau");
  return 0;
}
\end{lstlisting}
\subsection{Sequence Alignment (Needleman-Wunsch)}
Given two strings, determine their minimum-cost alignment. An alignment of two strings is a transformation of both strings by inserting gap characters \inlinecode{\_} in some way to make the final lengths equal. The total cost of an alignment given a \inlinecode{gap\_cost} (insertion or deletion cost) and a \inlinecode{sub\_cost} (substitution, i.e. mismatch cost) is \inlinecode{gap\_cost} times the number of inserted gaps, plus \inlinecode{sub\_cost} times the number of indices at which the two aligned strings differ.

\begin{compactitem}
  \item \inlinecode{align\_sequences(s1, s2, gap\_cost, sub\_cost)} returns a pair of aligned strings for strings \inlinecode{s1} and \inlinecode{s2}, using a classic dynamic programming approach. This implementation first computes \inlinecode{dp[i][j]} (the cost of aligning the length $i$ prefix of \inlinecode{s1} with the length $j$ prefix of \inlinecode{s2}) before following the path backwards to construct the answer. For \inlinecode{gap\_cost = sub\_cost = 1}, \inlinecode{dp[n][m]} will be the Levenshtein edit distance, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.
  \item \inlinecode{hirschberg\_align\_sequences(s1, s2, gap\_cost, sub\_cost)} returns the sequence alignment of strings \inlinecode{s1} and \inlinecode{s2} using the more memory efficient Hirschberg's algorithm.
\end{compactitem}

\textbf{Time Complexity}: $O(n \cdot m)$ per call to \inlinecode{align\_sequences(s1, s2)} as well as \inlinecode{hirschberg\_align\_sequences(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(n \cdot m)$ auxiliary heap space for \inlinecode{align\_sequences(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.
  \item $O(\log\left(\max\left(n, m\right)\right))$ auxiliary stack space and $O(\min\left(n, m\right))$ auxiliary heap space for \inlinecode{hirschberg\_align\_sequences(s1, s2)}, where $n$ and $m$ are the lengths of \inlinecode{s1} and \inlinecode{s2}, respectively.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <string>
#include <utility>
#include <vector>
using std::string;

std::pair<string, string> align_sequences(
    const string &s1, const string &s2, int gap_cost = 1, int sub_cost = 1
) {
  int n = static_cast<int>(s1.size()), m = static_cast<int>(s2.size());
  std::vector<std::vector<int>> dp(n + 1, std::vector<int>(m + 1, 0));
  for (int i = 0; i <= n; i++) {
    dp[i][0] = i * gap_cost;
  }
  for (int j = 0; j <= m; j++) {
    dp[0][j] = j * gap_cost;
  }
  for (int i = 1; i <= n; i++) {
    for (int j = 1; j <= m; j++) {
      dp[i][j] = (s1[i - 1] == s2[j - 1]) ? dp[i - 1][j - 1]
                                          : std::min(
                                                dp[i - 1][j - 1] + sub_cost,
                                                std::min(dp[i - 1][j], dp[i][j - 1]) + gap_cost
                                            );
    }
  }
  string res1, res2;
  int i = n, j = m;
  while (i > 0 && j > 0) {
    if (s1[i - 1] == s2[j - 1] || dp[i][j] == dp[i - 1][j - 1] + sub_cost) {
      res1 += s1[--i];
      res2 += s2[--j];
    } else if (dp[i][j] == dp[i - 1][j] + gap_cost) {
      res1 += s1[--i];
      res2 += '_';
    } else if (dp[i][j] == dp[i][j - 1] + gap_cost) {
      res1 += '_';
      res2 += s2[--j];
    }
  }
  while (i > 0 || j > 0) {
    res1 += (i > 0) ? s1[--i] : '_';
    res2 += (j > 0) ? s2[--j] : '_';
  }
  std::reverse(res1.begin(), res1.end());
  std::reverse(res2.begin(), res2.end());
  return {res1, res2};
}

template<class It>
std::vector<int> row_cost(It lo1, It hi1, It lo2, It hi2, int gap_cost, int sub_cost) {
  std::vector<int> res(std::distance(lo2, hi2) + 1), prev(res);
  for (It it1 = lo1; it1 != hi1; ++it1) {
    res.swap(prev);
    int i = 0;
    for (It it2 = lo2; it2 != hi2; ++it2) {
      res[i + 1] = (*it1 == *it2) ? prev[i] : std::min(prev[i] + sub_cost, res[i] + gap_cost);
      i++;
    }
  }
  return res;
}

template<class It>
void hirschberg_rec(
    It lo1, It hi1, It lo2, It hi2, string *res1, string *res2, int gap_cost, int sub_cost
) {
  if (lo1 == hi1) {
    for (It it2 = lo2; it2 != hi2; ++it2) {
      *res1 += '_';
      *res2 += *it2;
    }
    return;
  }
  if (lo1 + 1 == hi1) {
    It pos = std::find(lo2, hi2, *lo1);
    bool insert = (pos == hi2) && (gap_cost * (hi2 - lo2 + 1) < sub_cost);
    if (lo2 == hi2 || insert) {
      *res1 += *lo1;
      *res2 += '_';
    }
    for (It it2 = lo2; it2 != hi2; ++it2) {
      *res1 += (pos == it2 || (!insert && it2 == lo2)) ? *lo1 : '_';
      *res2 += *it2;
    }
    return;
  }
  It mid1 = lo1 + (hi1 - lo1) / 2;
  std::reverse_iterator<It> rlo1(hi1), rmid1(mid1), rlo2(hi2), rhi2(lo2);
  std::vector<int> fwd = row_cost(lo1, mid1, lo2, hi2, gap_cost, sub_cost);
  std::vector<int> rev = row_cost(rlo1, rmid1, rlo2, rhi2, gap_cost, sub_cost);
  It mid2 = lo2;
  int mincost = -1;
  for (int i = 0, j = static_cast<int>(rev.size()) - 1; i < static_cast<int>(fwd.size());
       i++, j--) {
    if (mincost < 0 || fwd[i] + rev[j] < mincost) {
      mincost = fwd[i] + rev[j];
      mid2 = lo2 + i;
    }
  }
  hirschberg_rec(lo1, mid1, lo2, mid2, res1, res2, gap_cost, sub_cost);
  hirschberg_rec(mid1, hi1, mid2, hi2, res1, res2, gap_cost, sub_cost);
}

std::pair<string, string> hirschberg_align_sequences(
    const string &s1, const string &s2, int gap_cost = 1, int sub_cost = 1
) {
  if (s1.size() < s2.size()) {
    return hirschberg_align_sequences(s2, s1, gap_cost, sub_cost);
  }
  string res1, res2;
  hirschberg_rec(s1.begin(), s1.end(), s2.begin(), s2.end(), &res1, &res2, gap_cost, sub_cost);
  return {res1, res2};
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(
      align_sequences("AGGGCT", "AGGCA", 2, 3) == (std::pair<string, string>{"AGGGCT", "A_GGCA"})
  );
  assert(
      hirschberg_align_sequences("AGGGCT", "AGGCA", 2, 3) ==
      (std::pair<string, string>{"AGGGCT", "A_GGCA"})
  );
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Expression Parsing}{3.7 Expression Parsing}}
\setcounter{section}{7}
\setcounter{subsection}{0}
\subsection{Recursive Descent Parsing (Simple)}
Evaluate an expression in accordance to the order of operations (parentheses, unary plus and minus signs, multiplication/division, addition/subtraction). The following is a minimalistic recursive descent implementation using iterators. Each precedence level is parsed by its own function that obtains operands by calling the next-tighter level, so the call structure itself enforces the order of operations.

\begin{compactitem}
  \item \inlinecode{eval(s)} returns an evaluation of the arithmetic expression \inlinecode{s}.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ per call to \inlinecode{eval(s)}, where $n$ is the length of \inlinecode{s}.

\textbf{Space Complexity}: $O(n)$ auxiliary stack space for \inlinecode{eval(s)}, where $n$ is the length of \inlinecode{s}.

\begin{lstlisting}[language=C++]
#include <string>

template<class It>
int eval(It &it, int prec) {
  if (prec == 0) {
    int sign = 1, res = 0;
    for (; *it == '-'; it++) {
      sign *= -1;
    }
    if (*it == '(') {
      res = eval(++it, 2);
      it++;
    } else
      while (*it >= '0' && *it <= '9') {
        res = 10 * res + (*(it++) - '0');
      }
    return sign * res;
  }
  int num = eval(it, prec - 1);
  while (!((prec == 2 && *it != '+' && *it != '-') || (prec == 1 && *it != '*' && *it != '/'))) {
    switch (*(it++)) {
      case '+':
        num += eval(it, prec - 1);
        break;
      case '-':
        num -= eval(it, prec - 1);
        break;
      case '*':
        num *= eval(it, prec - 1);
        break;
      case '/':
        num /= eval(it, prec - 1);
        break;
    }
  }
  return num;
}

int eval(const std::string &s) {
  std::string expr = s + '\0';
  auto it = expr.begin();
  return eval(it, 2);
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(eval("1++1") == 2);
  assert(eval("1+2*3*4+3*(2+2)-100") == -63);
  return 0;
}
\end{lstlisting}
\subsection{Recursive Descent Parsing}
Evaluate an expression with custom operand types, prefix unary operators, binary operators, and precedences. Evaluation is performed by recursive descent: the parser recursively evaluates tighter precedence levels before looser ones. Parentheses are supported, but multiplication by juxtaposition is not.

An arbitrary operand type is supported by changing the \inlinecode{Operand} alias and the static \inlinecode{is\_operand()} and \inlinecode{eval\_operand()} helpers. For maximum reliability, the string representation of operands should not use characters shared by any operator. For instance, the best practice instead of accepting \inlinecode{-1} as a valid operand (since the \inlinecode{-} sign may conflict with the identical binary operator), is to specify non-negative numbers as operands alongside the unary operator \inlinecode{-}.

Operators may be non-empty strings of any length, but should not contain any parentheses or shared characters with the string representations of operands. Ideally, operators should not be prefixes or suffixes of one another, else the tokenization process may be ambiguous. For example, if \inlinecode{++} and \inlinecode{+} are both operators, then \inlinecode{++} may be split into either \inlinecode{["+", "+"]} or \inlinecode{["++"]} depending on the lexicographical ordering of conflicting operators.

\begin{compactitem}
  \item \inlinecode{RecursiveDescentParser(unary\_op, binary\_op)} initializes a parser with operators specified by hash tables \inlinecode{unary\_op} (of operator to unary function object) and \inlinecode{binary\_op} (of operator to pair of binary function object and operator precedence). Operator precedences should be numbered upwards starting at 0 (lowest precedence, evaluated last).
  \item \inlinecode{split(s)} returns a vector of tokens for the expression \inlinecode{s}, split on the given operators during construction. Each parenthesis, operator, and operand satisfying \inlinecode{is\_operand()} will be split into a separate token. The algorithm is naive, matching operators lazily in the case of overlapping operators as mentioned above. Under these circumstances, the parse may not always succeed.
  \item \inlinecode{eval(lo, hi)} returns the evaluation of a range \inlinecode{[lo, hi)} of already split-up expression tokens, where \inlinecode{lo} and \inlinecode{hi} must be random-access iterators.
  \item \inlinecode{eval(s)} returns the evaluation of expression \inlinecode{s}, after first calling \inlinecode{split(s)} to obtain the tokens.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(m)$ per call to the constructor, where $m$ is the total number of operators.
  \item $O(n \cdot m \cdot k)$ per call to \inlinecode{split(s)}, where $n$ is the length of \inlinecode{s}, $m$ is the total number of operators defined for the parser instance, and $k$ is the maximum length for any operator representation.
  \item $O(n)$ expected per call to \inlinecode{eval(lo, hi)}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(n \cdot m \cdot k + n)$ expected per call to \inlinecode{eval(s)}, where $n$ is the length of \inlinecode{s}, and $m$ and $k$ are as defined previously.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m \cdot k)$ for storage of the $m$ operators, of maximum length $k$.
  \item $O(n)$ auxiliary stack space for \inlinecode{split(s)}, \inlinecode{eval(lo, hi)}, and \inlinecode{eval(s)}, where $n$ is the length of the argument.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cctype>
#include <functional>
#include <set>
#include <sstream>
#include <stdexcept>
#include <string>
#include <unordered_map>
#include <utility>
#include <vector>
using std::string;

// Define the custom operand type and representation below.
using Operand = double;
using UnaryOp = std::function<Operand(Operand)>;
using BinaryOp = std::function<Operand(Operand, Operand)>;

struct BinaryRule {
  BinaryOp op;
  int precedence;
};

class RecursiveDescentParser {
  using unary_op_map = std::unordered_map<string, UnaryOp>;
  using binary_op_map = std::unordered_map<string, BinaryRule>;
  unary_op_map unary_ops;
  binary_op_map binary_ops;
  std::set<string> op_tokens;
  int max_precedence;

  static bool is_operand(const string &s) {
    int npoints = 0;
    for (char c : s) {
      if (c == '.') {
        if (++npoints > 1) {
          return false;
        }
      } else if (!isdigit(static_cast<unsigned char>(c))) {
        return false;
      }
    }
    return !s.empty();
  }

  static Operand eval_operand(const string &s) {
    Operand res;
    std::stringstream ss(s);
    ss >> res;
    return res;
  }

  template<class StrIt>
  Operand eval_unary(StrIt &lo, StrIt hi) {
    if (is_operand(*lo)) {
      return eval_operand(*(lo++));
    }
    if (auto it = unary_ops.find(*lo); it != unary_ops.end()) {
      return (it->second)(eval_unary(++lo, hi));
    }
    if (*lo != "(") {
      throw std::runtime_error("Expected \"(\" during eval.");
    }
    Operand res = eval_binary(++lo, hi, 0);
    if (*lo != ")") {
      throw std::runtime_error("Expected \")\" during eval.");
    }
    ++lo;
    return res;
  }

  template<class StrIt>
  Operand eval_binary(StrIt &lo, StrIt hi, int precedence) {
    if (precedence > max_precedence) {
      return eval_unary(lo, hi);
    }
    Operand v = eval_binary(lo, hi, precedence + 1);
    while (lo != hi) {
      auto it = binary_ops.find(*lo);
      if (it == binary_ops.end() || it->second.precedence != precedence) {
        return v;
      }
      v = (it->second.op)(v, eval_binary(++lo, hi, precedence + 1));
    }
    return v;
  }

  static string strip(string s) {
    auto not_space = [](unsigned char c) { return !std::isspace(c); };
    s.erase(s.begin(), std::find_if(s.begin(), s.end(), not_space));
    s.erase(std::find_if(s.rbegin(), s.rend(), not_space).base(), s.end());
    return s;
  }

 public:
  RecursiveDescentParser(const unary_op_map &unary_ops, const binary_op_map &binary_ops)
      : unary_ops(unary_ops), binary_ops(binary_ops) {
    for (const auto &[op, _] : unary_ops) {
      op_tokens.insert(op);
    }
    max_precedence = 0;
    for (const auto &[op, fn_prec] : binary_ops) {
      op_tokens.insert(op);
      max_precedence = std::max(max_precedence, fn_prec.precedence);
    }
  }

  std::vector<string> split(const string &s) {
    std::vector<string> res;
    for (int i = 0; i < static_cast<int>(s.size()); i++) {
      if (s[i] == ' ') {
        continue;
      }
      int next_paren = static_cast<int>(s.size());
      for (int j = i; j < static_cast<int>(s.size()); j++) {
        if (s[j] == '(' || s[j] == ')') {
          next_paren = j;
          break;
        }
      }
      while (i < next_paren) {
        int found = next_paren;
        string found_op;
        for (int j = i; j < next_paren && found == next_paren; j++) {
          for (const auto &op : op_tokens) {
            if (s.substr(j, op.size()) == op) {
              found = j;
              found_op = op;
              break;
            }
          }
        }
        string term = strip(s.substr(i, found - i));
        if (!term.empty()) {
          res.push_back(term);
          if (!is_operand(term)) {
            throw std::runtime_error("Failed to split term: \"" + term + "\".");
          }
        }
        if (found < next_paren) {
          res.push_back(found_op);
          i = found + static_cast<int>(found_op.size());
        } else {
          i = next_paren;
        }
      }
      if (next_paren < static_cast<int>(s.size())) {
        res.emplace_back(1, s[next_paren]);
      }
    }
    return res;
  }

  template<class StrIt>
  Operand eval(StrIt lo, StrIt hi) {
    Operand res = eval_binary(lo, hi, 0);
    if (lo != hi) {
      throw std::runtime_error("Eval failed at token " + *lo + ".");
    }
    return res;
  }

  Operand eval(const string &s) {
    std::vector<string> tokens = split(s);
    return eval(tokens.begin(), tokens.end());
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cmath>
using namespace std;

#define EQ(a, b) (fabs((a) - (b)) < 1e-7)

int main() {
  unordered_map<string, UnaryOp> unary_ops;
  unary_ops["+"] = [](double x) { return +x; };
  unary_ops["-"] = [](double x) { return -x; };

  unordered_map<string, BinaryRule> binary_ops;
  binary_ops["+"] = {[](double a, double b) { return a + b; }, 0};
  binary_ops["-"] = {[](double a, double b) { return a - b; }, 0};
  binary_ops["*"] = {[](double a, double b) { return a * b; }, 1};
  binary_ops["/"] = {[](double a, double b) { return a / b; }, 1};
  binary_ops["^"] = {[](double a, double b) { return std::pow(a, b); }, 2};

  RecursiveDescentParser p(unary_ops, binary_ops);
  assert(EQ(p.eval("-+-((--(-+1)))"), -1));
  assert(EQ(p.eval("5*(3+3)-2-2"), 26));
  assert(EQ(p.eval("1+2*3*4+3*(+2)-100"), -69));
  assert(EQ(p.eval("3*3*3*3*3*3-2*2*2*2*2*2*2*2"), 473));
  assert(EQ(p.eval("3.14 + 3 * (7.7/9.8^32.9  )"), 3.14));
  assert(EQ(p.eval("5*(3+2)/-1*-2+(-2-2-2+3)-3-(-2)+15/2/2/2+(-2)"), 45.875));
  assert(EQ(p.eval("123456789./3/3/3*2*2*2+456/6-23/3"), 36579857.6666666667));
  assert(EQ(p.eval("10/3+10/4+10/5+10/6+10/7+10/8+10/9+10/10+15*23456"), 351854.28968253968));
  assert(
      EQ(p.eval(
             "-(5-(5-(5-(5-(5-2)))))+(3-(3-(3-(3-(3+3)))))*"
             "(7-(7-(7-(7-(7-7+4*5)))))"
         ),
         117)
  );
  return 0;
}
\end{lstlisting}
\subsection{Shunting Yard Parsing}
Evaluate an expression with custom operand types, prefix unary operators, binary operators, and precedences. Evaluation is performed using the shunting yard algorithm, which maintains operator and value stacks while respecting precedence and parentheses. Multiplication by juxtaposition is not supported.

An arbitrary operand type is supported by changing the \inlinecode{Operand} alias and the static \inlinecode{is\_operand()} and \inlinecode{eval\_operand()} helpers. For maximum reliability, the string representation of operands should not use characters shared by any operator. For instance, the best practice instead of accepting \inlinecode{-1} as a valid operand (since the \inlinecode{-} sign may conflict with the identical binary operator), is to specify non-negative numbers as operands alongside the unary operator \inlinecode{-}.

Operators may be non-empty strings of any length, but should not contain any parentheses or shared characters with the string representations of operands. Ideally, operators should not be prefixes or suffixes of one another, else the tokenization process may be ambiguous. For example, if \inlinecode{++} and \inlinecode{+} are both operators, then \inlinecode{++} may be split into either \inlinecode{["+", "+"]} or \inlinecode{["++"]} depending on the lexicographical ordering of conflicting operators.

\begin{compactitem}
  \item \inlinecode{ShuntingYardParser(unary\_op, binary\_op)} initializes a parser with operators specified by hash tables \inlinecode{unary\_op} (of operator to unary function object) and \inlinecode{binary\_op} (of operator to pair of binary function object and operator precedence). Operator precedences should be numbered upwards starting at 0 (lowest precedence, evaluated last).
  \item \inlinecode{split(s)} returns a vector of tokens for the expression \inlinecode{s}, split on the given operators during construction. Each parenthesis, operator, and operand satisfying \inlinecode{is\_operand()} will be split into a separate token. The algorithm is naive, matching operators lazily in the case of overlapping operators as mentioned above. Under these circumstances, the parse may not always succeed.
  \item \inlinecode{eval(lo, hi)} returns the evaluation of a range \inlinecode{[lo, hi)} of already split-up expression tokens, where \inlinecode{lo} and \inlinecode{hi} must be random-access iterators.
  \item \inlinecode{eval(s)} returns the evaluation of expression \inlinecode{s}, after first calling \inlinecode{split(s)} to obtain the tokens.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(m)$ per call to the constructor, where $m$ is the total number of operators.
  \item $O(n \cdot m \cdot k)$ per call to \inlinecode{split(s)}, where $n$ is the length of \inlinecode{s}, $m$ is the total number of operators defined for the parser instance, and $k$ is the maximum length for any operator representation.
  \item $O(n)$ expected per call to \inlinecode{eval(lo, hi)}, where $n$ is the distance between \inlinecode{lo} and \inlinecode{hi}.
  \item $O(n \cdot m \cdot k + n)$ expected per call to \inlinecode{eval(s)}, where $n$ is the length of \inlinecode{s}, and $m$ and $k$ are as defined previously.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m \cdot k)$ for storage of the $m$ operators, of maximum length $k$.
  \item $O(n)$ auxiliary stack space for \inlinecode{split(s)}, \inlinecode{eval(lo, hi)}, and \inlinecode{eval(s)}, where $n$ is the length of the argument.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <cctype>
#include <functional>
#include <set>
#include <sstream>
#include <stack>
#include <stdexcept>
#include <string>
#include <unordered_map>
#include <utility>
#include <vector>
using std::string;

// Define the custom operand type and representation below.
using Operand = double;
using UnaryOp = std::function<Operand(Operand)>;
using BinaryOp = std::function<Operand(Operand, Operand)>;

struct BinaryRule {
  BinaryOp op;
  int precedence;
};

class ShuntingYardParser {
  using UnaryOpMap = std::unordered_map<string, UnaryOp>;
  using BinaryOpMap = std::unordered_map<string, BinaryRule>;
  UnaryOpMap unary_ops;
  BinaryOpMap binary_ops;
  std::set<string> op_tokens;

  static Operand eval_operand(const string &s) {
    Operand res;
    std::stringstream ss(s);
    ss >> res;
    return res;
  }

  static bool is_operand(const string &s) {
    int npoints = 0;
    for (char c : s) {
      if (c == '.') {
        if (++npoints > 1) {
          return false;
        }
      } else if (!isdigit(static_cast<unsigned char>(c))) {
        return false;
      }
    }
    return !s.empty();
  }

  static string strip(string s) {
    auto not_space = [](unsigned char c) { return !std::isspace(c); };
    s.erase(s.begin(), std::find_if(s.begin(), s.end(), not_space));
    s.erase(std::find_if(s.rbegin(), s.rend(), not_space).base(), s.end());
    return s;
  }

 public:
  ShuntingYardParser(const UnaryOpMap &unary_ops, const BinaryOpMap &binary_ops)
      : unary_ops(unary_ops), binary_ops(binary_ops) {
    for (const auto &[op, fn] : unary_ops) {
      op_tokens.insert(op);
    }
    for (const auto &[op, fn_prec] : binary_ops) {
      op_tokens.insert(op);
    }
  }

  std::vector<string> split(const string &s) {
    std::vector<string> res;
    for (int i = 0; i < static_cast<int>(s.size()); i++) {
      if (s[i] == ' ') {
        continue;
      }
      int next_paren = static_cast<int>(s.size());
      for (int j = i; j < static_cast<int>(s.size()); j++) {
        if (s[j] == '(' || s[j] == ')') {
          next_paren = j;
          break;
        }
      }
      while (i < next_paren) {
        int found = next_paren;
        string found_op;
        for (int j = i; j < next_paren && found == next_paren; j++) {
          for (const auto &op : op_tokens) {
            if (s.substr(j, op.size()) == op) {
              found = j;
              found_op = op;
              break;
            }
          }
        }
        string term = strip(s.substr(i, found - i));
        if (!term.empty()) {
          res.push_back(term);
          if (!is_operand(term)) {
            throw std::runtime_error("Failed to split term: \"" + term + "\".");
          }
        }
        if (found < next_paren) {
          res.push_back(found_op);
          i = found + static_cast<int>(found_op.size());
        } else {
          i = next_paren;
        }
      }
      if (next_paren < static_cast<int>(s.size())) {
        res.emplace_back(1, s[next_paren]);
      }
    }
    return res;
  }

  template<class StrIt>
  Operand eval(StrIt lo, StrIt hi) {
    std::stack<Operand> vals;
    std::stack<std::pair<string, bool>> op_stack;
    op_stack.emplace("(", false);
    StrIt prev = hi;
    do {
      string curr = (lo == hi) ? ")" : *lo;
      if (is_operand(curr)) {
        vals.push(eval_operand(curr));
      } else if (curr == "(") {
        op_stack.emplace(curr, false);
      } else if (
          unary_ops.find(curr) != unary_ops.end() &&
          (prev == hi || *prev == "(" || binary_ops.find(*prev) != binary_ops.end())
      ) {
        op_stack.emplace(curr, true);
      } else {
        while (true) {
          auto [op, is_unary] = op_stack.top();
          auto it1 = binary_ops.find(op);
          auto it2 = binary_ops.find(curr);
          if (!is_unary && (it1 == binary_ops.end() ? -1 : it1->second.precedence) <
                               (it2 == binary_ops.end() ? -1 : it2->second.precedence)) {
            break;
          }
          op_stack.pop();
          if (op == "(") {
            break;
          }
          Operand b = vals.top();
          vals.pop();
          if (is_unary) {
            if (auto it = unary_ops.find(op); it != unary_ops.end()) {
              vals.push((it->second)(b));
            } else {
              throw std::runtime_error("Failed to eval unary op: " + op);
            }
          } else {
            Operand a = vals.top();
            vals.pop();
            if (it1 == binary_ops.end()) {
              throw std::runtime_error("Failed to eval binary op: " + op);
            }
            vals.push((it1->second.op)(a, b));
          }
        }
        if (curr != ")") {
          op_stack.emplace(*lo, false);
        }
      }
      prev = lo;
    } while (lo++ != hi);
    return vals.top();
  }

  Operand eval(const string &s) {
    std::vector<string> tokens = split(s);
    return eval(tokens.begin(), tokens.end());
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cmath>
using namespace std;

#define EQ(a, b) (fabs((a) - (b)) < 1e-7)

int main() {
  unordered_map<string, UnaryOp> unary_ops;
  unary_ops["+"] = [](double x) { return +x; };
  unary_ops["-"] = [](double x) { return -x; };

  unordered_map<string, BinaryRule> binary_ops;
  binary_ops["+"] = {[](double a, double b) { return a + b; }, 0};
  binary_ops["-"] = {[](double a, double b) { return a - b; }, 0};
  binary_ops["*"] = {[](double a, double b) { return a * b; }, 1};
  binary_ops["/"] = {[](double a, double b) { return a / b; }, 1};
  binary_ops["^"] = {[](double a, double b) { return std::pow(a, b); }, 2};

  ShuntingYardParser p(unary_ops, binary_ops);
  assert(EQ(p.eval("-+-((--(-+1)))"), -1));
  assert(EQ(p.eval("5*(3+3)-2-2"), 26));
  assert(EQ(p.eval("1+2*3*4+3*(+2)-100"), -69));
  assert(EQ(p.eval("3*3*3*3*3*3-2*2*2*2*2*2*2*2"), 473));
  assert(EQ(p.eval("3.14 + 3 * (7.7/9.8^32.9  )"), 3.14));
  assert(EQ(p.eval("5*(3+2)/-1*-2+(-2-2-2+3)-3-(-2)+15/2/2/2+(-2)"), 45.875));
  assert(EQ(p.eval("123456789./3/3/3*2*2*2+456/6-23/3"), 36579857.6666666667));
  assert(EQ(p.eval("10/3+10/4+10/5+10/6+10/7+10/8+10/9+10/10+15*23456"), 351854.28968253968));
  assert(
      EQ(p.eval(
             "-(5-(5-(5-(5-(5-2)))))+(3-(3-(3-(3-(3+3)))))*"
             "(7-(7-(7-(7-(7-7+4*5)))))"
         ),
         117)
  );
  return 0;
}
\end{lstlisting}

\section{\texorpdfstring{Encoding and Compression}{3.8 Encoding and Compression}}
\setcounter{section}{8}
\setcounter{subsection}{0}
\subsection{Entropy and Frequency Tables}
Computes byte frequency tables and empirical information-theoretic quantities for strings. These primitives are useful when analyzing compression, building coding schemes such as Huffman coding, or comparing symbol distributions.

For a string with symbol probabilities $p_i$, empirical entropy is $H = -\sum_i p_i \log_2 p_i$ bits per symbol. This is a lower bound on the average number of bits per symbol achievable by any uniquely decodable code for the same independent symbol model.

\begin{compactitem}
  \item \inlinecode{byte\_frequencies(s)} returns a length-256 table of byte counts in string \inlinecode{s}.
  \item \inlinecode{entropy(freq)} returns empirical entropy in bits per symbol for a frequency table.
  \item \inlinecode{entropy(s)} returns empirical entropy in bits per symbol for string \inlinecode{s}.
  \item \inlinecode{expected\_code\_length(freq, length)} returns the average encoded bits per symbol for code lengths \inlinecode{length[c]}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n + m)$ per call to \inlinecode{entropy(s)}, where $n$ is the string length and $m = 256$.
  \item $O(m)$ per call to \inlinecode{entropy(freq)} and \inlinecode{expected\_code\_length(freq, length)}.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m)$ auxiliary heap space for \inlinecode{byte\_frequencies(s)}.
  \item $O(1)$ auxiliary for the other operations.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <cmath>
#include <string>
#include <vector>
using std::string;

std::vector<int> byte_frequencies(const string &s) {
  std::vector<int> freq(256, 0);
  for (int i = 0; i < static_cast<int>(s.size()); i++) {
    freq[static_cast<unsigned char>(s[i])]++;
  }
  return freq;
}

double entropy(const std::vector<int> &freq) {
  int total = 0;
  for (int i = 0; i < static_cast<int>(freq.size()); i++) {
    total += freq[i];
  }
  if (total == 0) {
    return 0;
  }
  double res = 0;
  for (int i = 0; i < static_cast<int>(freq.size()); i++) {
    if (freq[i] == 0) {
      continue;
    }
    double p = static_cast<double>(freq[i]) / total;
    res -= p * (log(p) / log(2.0));
  }
  return res;
}

double entropy(const string &s) {
  return entropy(byte_frequencies(s));
}

double expected_code_length(const std::vector<int> &freq, const std::vector<int> &length) {
  int total = 0;
  for (int i = 0; i < static_cast<int>(freq.size()); i++) {
    total += freq[i];
  }
  if (total == 0) {
    return 0;
  }
  double res = 0;
  for (int i = 0; i < static_cast<int>(freq.size()); i++) {
    res += static_cast<double>(freq[i]) * length[i] / total;
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
#include <cmath>
using namespace std;

bool close(double a, double b) {
  return fabs(a - b) < 1e-9;
}

int main() {
  vector<int> freq = byte_frequencies("aabb");
  assert(freq['a'] == 2);
  assert(freq['b'] == 2);
  assert(close(entropy(freq), 1.0));
  assert(close(entropy("aaaa"), 0.0));

  vector<int> length(256, 0);
  length['a'] = 1;
  length['b'] = 1;
  assert(close(expected_code_length(freq, length), 1.0));
  return 0;
}
\end{lstlisting}
\subsection{Run-Length Encoding}
Encodes consecutive equal characters as (character, count) runs. Run-length encoding is a simple lossless compression technique that works well when strings contain long repeated blocks and poorly when repetitions are rare.

The representation below stores runs in a vector of pairs rather than formatting them as a string. This avoids ambiguity when the original text contains digits or separator characters.

\begin{compactitem}
  \item \inlinecode{run\_length\_encode(s)} returns the sequence of runs in string \inlinecode{s}.
  \item \inlinecode{run\_length\_decode(runs)} reconstructs the original string from a sequence of runs.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ per call, where $n$ is the input or output length.

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(r)$ for encoded output, where $r$ is the number of runs.
  \item $O(n)$ for decoded output.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <string>
#include <utility>
#include <vector>
using std::string;

std::vector<std::pair<char, int>> run_length_encode(const string &s) {
  std::vector<std::pair<char, int>> res;
  for (int i = 0; i < static_cast<int>(s.size()); i++) {
    if (res.empty() || res.back().first != s[i]) {
      res.emplace_back(s[i], 1);
    } else {
      res.back().second++;
    }
  }
  return res;
}

string run_length_decode(const std::vector<std::pair<char, int>> &runs) {
  string res;
  for (int i = 0; i < static_cast<int>(runs.size()); i++) {
    res.append(runs[i].second, runs[i].first);
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  string s = "aaabccccdd";
  vector<pair<char, int>> runs = run_length_encode(s);
  assert(runs.size() == 4);
  assert(runs[0] == (pair<char, int>{'a', 3}));
  assert(runs[2] == (pair<char, int>{'c', 4}));
  assert(run_length_decode(runs) == s);
  return 0;
}
\end{lstlisting}
\subsection{Base64 Encoding}
Encodes binary data as printable ASCII using Base64, and decodes Base64 text back to bytes. Base64 is an encoding, not encryption or compression: it preserves all input data exactly, but usually increases size by about one third. Each group of three input bytes is split into four 6-bit values, each selecting one character of a 64-symbol alphabet, with \inlinecode{=} padding marking a final group of only one or two bytes.

This implementation uses the standard alphabet with \inlinecode{+}, \inlinecode{/}, and \inlinecode{=} padding. Whitespace is not skipped by the decoder; clean the input first if line breaks or spaces may appear.

\begin{compactitem}
  \item \inlinecode{base64\_encode(bytes)} returns the Base64 representation of string \inlinecode{bytes}.
  \item \inlinecode{base64\_decode(text)} returns the decoded byte string. It assumes \inlinecode{text} is valid padded Base64.
\end{compactitem}

\textbf{Time Complexity}: $O(n)$ per call, where $n$ is the input length.

\textbf{Space Complexity}: $O(n)$ for the output string.

\begin{lstlisting}[language=C++]
#include <string>
#include <vector>
using std::string;

const string BASE64_ALPHABET = "ABCDEFGHIJKLMNOPQRSTUVWXYZabcdefghijklmnopqrstuvwxyz0123456789+/";

string base64_encode(const string &bytes) {
  string res;
  int val = 0, bits = -6;
  for (int i = 0; i < static_cast<int>(bytes.size()); i++) {
    val = (val << 8) + static_cast<unsigned char>(bytes[i]);
    bits += 8;
    while (bits >= 0) {
      res.push_back(BASE64_ALPHABET[(val >> bits) & 63]);
      bits -= 6;
    }
  }
  if (bits > -6) {
    res.push_back(BASE64_ALPHABET[((val << 8) >> (bits + 8)) & 63]);
  }
  while (res.size() % 4) {
    res.push_back('=');
  }
  return res;
}

string base64_decode(const string &text) {
  std::vector<int> table(256, -1);
  for (int i = 0; i < 64; i++) {
    table[static_cast<unsigned char>(BASE64_ALPHABET[i])] = i;
  }
  string res;
  int val = 0, bits = -8;
  for (int i = 0; i < static_cast<int>(text.size()); i++) {
    if (text[i] == '=') {
      break;
    }
    int x = table[static_cast<unsigned char>(text[i])];
    if (x < 0) {
      break;
    }
    val = (val << 6) + x;
    bits += 6;
    if (bits >= 0) {
      res.push_back(static_cast<char>(((val >> bits) & 255)));
      bits -= 8;
    }
  }
  return res;
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  assert(base64_encode("hello") == "aGVsbG8=");
  assert(base64_decode("aGVsbG8=") == "hello");
  assert(base64_decode(base64_encode("abc123!?")) == "abc123!?");
  return 0;
}
\end{lstlisting}
\subsection{Huffman Coding}
Builds a Huffman code for a string and uses it to encode and decode that string. Huffman coding is a lossless prefix-code compression algorithm: more frequent characters receive shorter bit strings, and no code is a prefix of another code. The code tree is built greedily with a priority queue by repeatedly merging the two lowest-frequency subtrees until one tree remains; each character's bit string is then read off its root-to-leaf path.

The implementation below stores encoded bits as a string of \inlinecode{\textquotesingle{}0\textquotesingle{}} and \inlinecode{\textquotesingle{}1\textquotesingle{}} characters for clarity. For real compression, pack those bits into bytes. The tree is also needed to decode the bit string, so compressed data normally stores enough metadata to reconstruct the same tree.

\begin{compactitem}
  \item \inlinecode{HuffmanTree(text)} constructs a Huffman tree from the character frequencies in \inlinecode{text}.
  \item \inlinecode{codes()} returns a table mapping each byte value to its bit string.
  \item \inlinecode{encode(text)} returns the encoded bit string.
  \item \inlinecode{decode(bits)} returns the decoded text.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n + m \log m)$ to construct the tree, where $n$ is the input length and $m$ is the number of distinct characters.
  \item $O(n)$ to encode or decode, plus the number of produced or consumed bits.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Space Complexity}:
\begin{compactitem}
  \item $O(m)$ for the tree and code table, where $m$ is the number of distinct characters.
  \item $O(n)$ for the encoded or decoded output.
\end{compactitem}

\begin{lstlisting}[language=C++]
#include <queue>
#include <string>
#include <vector>
using std::string;

class HuffmanTree {
  struct Node {
    int freq;
    unsigned char ch;
    int left, right;

    Node(int freq = 0, unsigned char ch = 0, int left = -1, int right = -1)
        : freq(freq), ch(ch), left(left), right(right) {}
  };

  struct ByFrequency {
    const std::vector<Node> *nodes;
    explicit ByFrequency(const std::vector<Node> *nodes = nullptr) : nodes(nodes) {}
    bool operator()(int a, int b) const { return (*nodes)[a].freq > (*nodes)[b].freq; }
  };

  std::vector<Node> nodes;
  int root;
  std::vector<string> code;

  void build_codes(int u, const string &path) {
    if (nodes[u].left == -1 && nodes[u].right == -1) {
      code[nodes[u].ch] = path.empty() ? "0" : path;
      return;
    }
    build_codes(nodes[u].left, path + '0');
    build_codes(nodes[u].right, path + '1');
  }

 public:
  explicit HuffmanTree(const string &text) : root(-1), code(256) {
    std::vector<int> freq(256, 0);
    for (int i = 0; i < static_cast<int>(text.size()); i++) {
      freq[static_cast<unsigned char>(text[i])]++;
    }
    ByFrequency cmp(&nodes);
    std::priority_queue<int, std::vector<int>, ByFrequency> pq(cmp);
    for (int c = 0; c < 256; c++) {
      if (freq[c] > 0) {
        nodes.emplace_back(freq[c], static_cast<unsigned char>(c));
        pq.push(static_cast<int>(nodes.size()) - 1);
      }
    }
    if (pq.empty()) {
      return;
    }
    while (static_cast<int>(pq.size()) > 1) {
      int a = pq.top();
      pq.pop();
      int b = pq.top();
      pq.pop();
      nodes.emplace_back(nodes[a].freq + nodes[b].freq, 0, a, b);
      pq.push(static_cast<int>(nodes.size()) - 1);
    }
    root = pq.top();
    build_codes(root, "");
  }

  std::vector<string> codes() const { return code; }

  string encode(const string &text) const {
    string bits;
    for (int i = 0; i < static_cast<int>(text.size()); i++) {
      bits += code[static_cast<unsigned char>(text[i])];
    }
    return bits;
  }

  string decode(const string &bits) const {
    string text;
    if (root == -1) {
      return text;
    }
    // Degenerate tree: the input had a single distinct character, so the root is itself a leaf and
    // each symbol was encoded as one bit. Walking children would dereference node -1.
    if (nodes[root].left == -1 && nodes[root].right == -1) {
      return string(bits.size(), static_cast<char>(nodes[root].ch));
    }
    int u = root;
    for (int i = 0; i < static_cast<int>(bits.size()); i++) {
      u = bits[i] == '0' ? nodes[u].left : nodes[u].right;
      if (nodes[u].left == -1 && nodes[u].right == -1) {
        text.push_back(static_cast<char>(nodes[u].ch));
        u = root;
      }
    }
    return text;
  }
};
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>

int main() {
  string text = "banana bandana";
  HuffmanTree h(text);
  string bits = h.encode(text);
  assert(h.decode(bits) == text);
  assert(bits.size() < text.size() * 8);
  return 0;
}
\end{lstlisting}
\subsection{Burrows-Wheeler Transform}
The Burrows-Wheeler transform (BWT) permutes a string into one that tends to be far more compressible while remaining perfectly invertible. It writes down every cyclic rotation of the string, sorts them lexicographically, and reads off the last character of each sorted rotation. Characters that occur in similar contexts are brought together, producing long runs that run-length and move-to-front coders exploit; it is the heart of the bzip2 compressor and a building block of compressed-text indexes.

To make the permutation invertible, a unique sentinel character smaller than every real character is appended before transforming, so the rotations are all distinct and one of them is distinguished as the original. Sorting the rotations of the sentineled string is the same as sorting its suffixes, so the forward transform is computed with a prefix-doubling rotation sort. The inverse rebuilds the string with the last-to-first (LF) mapping: stably ordering the transformed characters yields, for each position of the sorted first column, the row that supplies the next character, and following that mapping from the sentinel row walks the original string out one character at a time.

The input must not contain the sentinel character \inlinecode{\textquotesingle{}\textbackslash{}0\textquotesingle{}}.

\begin{compactitem}
  \item \inlinecode{burrows\_wheeler\_transform(s)} returns the BWT of \inlinecode{s}. The result has length \inlinecode{s.size() + 1} and contains the sentinel \inlinecode{\textquotesingle{}\textbackslash{}0\textquotesingle{}} exactly once.
  \item \inlinecode{burrows\_wheeler\_inverse(code)} returns the original string, inverting \inlinecode{burrows\_wheeler\_transform}. The argument must be a string produced by that function.
\end{compactitem}

\Needspace*{3\baselineskip}
\textbf{Time Complexity}:
\begin{compactitem}
  \item $O(n \log^{2} n)$ per call to \inlinecode{burrows\_wheeler\_transform(s)}, where $n$ is the length of \inlinecode{s}.
  \item $O(n \log n)$ per call to \inlinecode{burrows\_wheeler\_inverse(code)}.
\end{compactitem}

\textbf{Space Complexity}: $O(n)$ auxiliary heap space for both functions.

\begin{lstlisting}[language=C++]
#include <algorithm>
#include <numeric>
#include <string>
#include <vector>

std::string burrows_wheeler_transform(std::string s) {
  s.push_back('\0');  // Unique sentinel, smaller than every real character.
  int n = static_cast<int>(s.size());
  // Sort the indices of the cyclic rotations by prefix doubling; with the sentinel this matches
  // sorting the suffixes.
  std::vector<int> order(n), rank(n), tmp(n);
  for (int i = 0; i < n; i++) {
    order[i] = i;
    rank[i] = static_cast<unsigned char>(s[i]);
  }
  for (int k = 1; k < n; k <<= 1) {
    auto cmp = [&](int a, int b) {
      if (rank[a] != rank[b]) {
        return rank[a] < rank[b];
      }
      return rank[(a + k) % n] < rank[(b + k) % n];
    };
    std::sort(order.begin(), order.end(), cmp);
    tmp[order[0]] = 0;
    for (int i = 1; i < n; i++) {
      tmp[order[i]] = tmp[order[i - 1]] + (cmp(order[i - 1], order[i]) ? 1 : 0);
    }
    rank = tmp;
    if (rank[order[n - 1]] == n - 1) {
      break;
    }
  }
  std::string code(n, '\0');
  for (int i = 0; i < n; i++) {
    code[i] = s[(order[i] + n - 1) % n];  // Last character of the i-th sorted rotation.
  }
  return code;
}

std::string burrows_wheeler_inverse(const std::string &code) {
  int n = static_cast<int>(code.size());
  // first[k] is the row supplying the first column's k-th character; following it from row 0 (the
  // sentinel row, sorted first) reconstructs the sentinel followed by the original string.
  std::vector<int> first(n);
  std::iota(first.begin(), first.end(), 0);
  std::stable_sort(first.begin(), first.end(), [&](int a, int b) {
    return static_cast<unsigned char>(code[a]) < static_cast<unsigned char>(code[b]);
  });
  std::string res;
  res.reserve(n);
  for (int i = 0, row = 0; i < n; i++) {
    row = first[row];
    res.push_back(code[row]);
  }
  return res.substr(1);  // Drop the leading sentinel.
}
\end{lstlisting}
\Needspace*{4\baselineskip}
\textbf{Example Usage}\par
\begin{lstlisting}[language=C++]
#include <cassert>
using namespace std;

int main() {
  string code = burrows_wheeler_transform("banana");
  assert(code.size() == 7);  // "banana" plus the sentinel.
  assert(burrows_wheeler_inverse(code) == "banana");

  // The transform clusters equal characters: "mississippi" yields long runs.
  string text = "mississippi";
  string t = burrows_wheeler_transform(text);
  assert(burrows_wheeler_inverse(t) == text);

  // Round-trips on assorted strings, including repeats and a single character.
  for (const string &s :
       {string(""), string("a"), string("aaaa"), string("abracadabra"),
        string("the quick brown fox"), string("\1\2abc", 5)}) {
    assert(burrows_wheeler_inverse(burrows_wheeler_transform(s)) == s);
  }
  return 0;
}
\end{lstlisting}
